---
title: "Bayesian Estimation of a Generalized Additive Model to Describe the Stage-Discharge Relation"
author: "Dave Holtschlag"
date: "`r format(Sys.Date(), '%A %b %d, %Y') `"
output: 
  html_document:
    toc: true
    toc_depth: 3
bibliography: bibliography.bib
---

## Abstract 
A Bayesian approach is developed to estimate parameters of a generalized additive model (GAM) [@Wood2006] to describe the dynamic relation between stage and discharge (flow) at a streamgage.  The GAM model provides a basis function and parameter estimates to describe the smooth, monotonically increasing relation between stage and flow using data on direct measurements of stage and flow commonly used to develop a so called rating curve. Bayesian estimation then uses the GAM basis function and re-estimates parameters to explicitly account for the uncertainty in the stage measurements. Data for USGS streamgage 04112500 Red Cedar River at East Lansing, MI are used to illustrate the approach and results of the generally applicable method.  Computations are implemented in the R programming environment [@cite_R].

### Setup computational environment

```{r setup, include=FALSE}
library( "rethinking" )
options(mc.cores = parallel::detectCores())
library(dataRetrieval)
library(lubridate)
library(tidyverse)
library(ggthemes)
library(splines)
library(mgcv)
library(fitdistrplus)
library(stringr)
library(kableExtra)
library(ggrepel)
library(ggcorrplot)
# Package expss contains the function vlookup
library(expss)
# Package to plot distribution
library(ggfortify)
# Multivariate Autoregressive State Space Model
library(MARSS)
knitr::opts_chunk$set(echo = TRUE)
```

## Specify Streamgage for Data Retrieval

```{r specify_gage}

# initialize counter for figures
fig_no <- 1

# Retrieve gage info for specified site Number
site_no    <- '04122500'

gage_info  <- readNWISsite(site_no)

# Standardize field names
gage_info  <- renameNWISColumns(gage_info)

# 
station_nm <- str_to_title( gage_info$station_nm )

print(head(gage_info))

```

### Retrieve Stage-Flow Rating

```{r read_rating_data, fig.width = 9}
# Rating curve at the selected site.  The 'exsa' parameter is an extended table
#   provide detailed info on stage discharge
rating19 <- read.csv('Data/04122500/04122500_rating_19.0.txt', header = FALSE, sep = '\t') 
colnames( rating19 ) <- c('rated_flow_19', 'stage')

rating20 <- read.csv('Data/04122500/04122500_rating_20.0.txt', header = FALSE, sep = '\t') 
colnames( rating20 ) <- c('rated_flow_20', 'stage')


rating21 <- read.csv('Data/04122500/04122500_rating_21.0.txt', header = FALSE, sep = '\t') 
colnames( rating21 ) <- c('rated_flow_21', 'stage')

ratings <- rating19 %>% 
  left_join( rating20 ) %>% 
  left_join( rating21)  %>% 
  dplyr::select(stage, everything() )
  
rating_long <- ratings %>% 
  gather( key = 'rating_no', value = 'rated_flow', rated_flow_19, rated_flow_20, rated_flow_21)

rating_long %>% 
  ggplot( aes( x = rated_flow, y = stage,  group = rating_no)) +
  geom_line( aes( color = rating_no, linetype = rating_no ), size = 1.5, alpha = 0.50) +
  theme_bw() +
  theme( legend.position = 'bottom' ) +
  xlab( 'Flow, in cubic feet per second' ) +
  ylab( 'Stage, in feet above gage datum' ) +
  labs( title =  'Figure 1. Selected rating curves at 04122500 Pere Marquette River at Scottville, MI' )
  
```




```{r gam_rating19, fig.width = 9}



# Estimate generalized additive model of stage - discharge rating 19 
gam1 <- gam( rated_flow_19 ~ s(stage), data = rating19 )

# Print summary of gam1
print(summary(gam1))


gam1_pred          <- predict(gam1, se.fit = TRUE)

rating19$gam_flow    <- gam1_pred$fit
rating19$se_fit      <- gam1_pred$se.fit
rating19$error_025   <- rating19$rated_flow_19 - (gam1_pred$fit + qnorm(0.025) * gam1_pred$se.fit)   
rating19$error_975   <- rating19$rated_flow_19 - (gam1_pred$fit + qnorm(0.975) * gam1_pred$se.fit) 
rating19$flow_error  <- gam1$residuals
rating19$pct_error   <- gam1$residuals / gam1$fitted.values * 100

rating19 %>% 
  # dplyr::arrange( stage ) %>% 
  ggplot(    aes( y = pct_error, x = stage  )) +
  geom_point( color = 'red', size = 0.5) +
  # geom_line( aes( y = error_025), color = 'salmon') +
  # geom_line( aes( y = error_975), color = 'salmon') +
  # geom_ribbon( aes( ymin = error_025, ymax = error_975), fill = 'blue', alpha = 0.25 ) +
  scale_y_continuous( breaks = seq( from = -0.5, to = 2, by = 0.5)) +
  ylab( 'Flow Error, in percent') +
  xlab( 'Stage, in feet') +
  theme_bw() +
  geom_hline( yintercept = 0, color = 'blue', linetype = 'dashed') +
  theme( legend.position = 'bottom') +
  labs( title = paste('Figure 2. Errors in Flow for GAM of rating curve 19 at', site_no, station_nm))

```




### Retrieve Stage-Flow Measurement Pairs

```{r read_meas_stage_flow}

meas_data   <- readNWISmeas(site_no, expanded = TRUE, convertType = TRUE)

# Standardize field names
meas_data  <- renameNWISColumns(meas_data)


meas_data <- meas_data %>% 
  rename('stage_round'   = 'gage_height_va',
         'flow'          = 'discharge_va',
         'qualifier'     = 'measured_rating_diff',
         'control'       = 'control_type_cd')      %>% 
  filter(qualifier      != 'Unspecified' | qualifier == !is.na(qualifier)) %>%
  mutate('qualifier'     =  ordered(qualifier, levels = c('Excellent', 'Good', 'Fair', 'Poor')),
         'log_flow'      = log10(flow),
         # rnd refers to the approximation of stage to 0.01 ft.
         'log_stage_round' = log10(stage_round))

# Populate field to combine similar Control attributes
meas_data$control_degree <- meas_data$control
meas_data$control_degree  = case_when(str_detect(meas_data$control, 'Heavy')       ~ 'Heavy',
                                      str_detect(meas_data$control, 'Moderate')    ~ 'Moderate',
                                      str_detect(meas_data$control, 'Light'   )    ~ 'Light',
                                      str_detect(meas_data$control, 'Clear'   )    ~ 'Clear',
                                      str_detect(meas_data$control, 'Ice'     )    ~ 'Ice',
                                      str_detect(meas_data$control, 'Unspecifed')  ~ 'Unspecifed')

meas_data$control_degree = ordered(meas_data$control_degree,
                                  levels = c('Unspecifed', 'Clear', 'Light', 'Moderate', 
                                             'Heavy', 'Ice'))

# Only cases where flow and stage are measured
meas_data <- meas_data[complete.cases(meas_data$stage, meas_data$flow),]
```

## 

```{r unit_daily_stage}

stage_unit <- read.csv('Data/04122500/stage/Gage_height.ft@04122500.EntireRecorda.csv', header = TRUE, sep = ',',
                       stringsAsFactors = FALSE, comment = '#')

# Convert from character field to datetime referenced to EST
stage_unit$datetime <- as_datetime(stage_unit$Timestamp..UTC.05.00, tz = 'America/New_York')

# Summarize unit stage to daily mean stage
stage_day <- stage_unit %>% 
  rename( stage = Value ) %>% 
  group_by(day = floor_date( datetime, 'day')) %>% 
  summarize( stage_mean_day  = mean( stage, na.rm = TRUE ),
             n         = n()) %>% 
  ungroup() %>% 
  arrange( day ) %>% 
  mutate( delta_day = as.numeric(day - lag( day ), units = 'days' ),
          day_date  = as.Date( day ))

# Measured flows and daily stages for the interval when rating 19 was active
#  Note: Discrete measurement timing is approximated to days
stage_flow_day_19 <- stage_day %>% 
  left_join( meas_data[, c('measurement_dt', 'flow', 'qualifier', 'meas_type')], 
             by = c( 'day_date'  = 'measurement_dt')) %>% 
  filter( day >= as.Date('1994-10-25', format = '%Y-%m-%d') & day <= as.Date('2011-03-01', format = '%Y-%m-%d' ))

```


## Inital state space model
yt is flow
Xt is the 10 x 1 state vector of GAM parameters
Q  is the fixed 10X10 covariance matrix of paramaters
R  is an estimated constant measurement variance
Zt is the matrix of fixed measurement components (smooth spline terms)
Yt is sequential daily values of log_stage_round 

```{r marss_m1}

# stage_flow_day_19

Yt <- stage_flow_day_19$flow

# Predict the intercept and slope components for the spline smooth 
Zt <- predict( gam1, newdata = data.frame( stage = stage_flow_day_19$stage_mean_day),
                                           type = 'lpmatrix')

Xt <- as.matrix( as.numeric(coefficients(gam1)), 10, 1)

# Vp is the estimated covariance matrix for the parameters. This is a Bayesian
# posterior covariance matrix that results from adopting a particular Bayesian
# model of the smoothing process. Paricularly useful for creating
# credible/confidence intervals

#   Under ML or REML smoothing parameter estimation it is possible to correct
#   the covariance matrix Vp for smoothing parameter uncertainty. This is the
#   corrected version

Q  <- as.list(as.matrix( as.numeric(gam1$Vp), 10, 10))
write.table(Q, file = 'Qmatrix.csv', row.names = F, col.names = F)

# write out matrix Q


# Allocate matrix with NAs to contain Xp components for measured log_stage values to days of measurement
Xp_aug <- matrix(NA, nrow = length(seq(ndx_min:ndx_max)), ncol = ncol(Xp))
# Replace NA indicators with measured log_stage values 
Xp_aug[ndx_log_stage,] <- Xp

# Identify measured flows during daily flow period in analysis

# Find indices of measured flows in daily analysis range
ndx_meas_flow <- which( meas_data$measurement_dt >= daily_data$measurement_dt[ndx_min] &
                        meas_data$measurement_dt <= daily_data$measurement_dt[ndx_max])

Yt <- rep(NA, length(seq(27:996)))
Yt[ndx_log_stage] <- meas_data$log_flow[ndx_meas_flow] 

R = matrix( list( "R_sigma2"), 1, 1)



# State-Space Model Component Z
Zt = t(Xp_aug)

dim(Zt) <- c(1, ncol(Xp), length(Yt))
x0 = matrix(GAM_coef_est, nrow = ncol(Xp), ncol = 1)

model.dat = Yt
model.gen = list(Z = Zt, x0 = x0, R = R )

m1 <- MARSS(model.dat, model.gen)


```


```{r }



```


```{r flow_meas_error}
# ---------------------------------
# Code  Description
# ---------------------------------
# E     Excellent    The data is within 2% (percent) of the actual flow
# G     Good         The data is within 5% (percent) of the actual flow
# F     Fair         The data is within 8% (percent) of the actual flow
# P     Poor         The data are >8% (percent) of the actual flow
# https://help.waterdata.usgs.gov/codes-and-parameters/discharge-measurement-quality-code

# qualifier <- .15; flow = 10000; stderr <- (log10(flow * (1 + qualifier)) - log10(flow))/2

meas_data <- meas_data %>% 
  mutate(qualifier_err = ifelse( qualifier == 'Excellent', 0.02,
                                 ifelse( qualifier == 'Good', 0.05, 
                                         ifelse( qualifier == 'Fair', 0.08,
                                                 ifelse( qualifier == 'Poor', 0.15, NA)))),
         meas_lflow_stderr = (log10( flow * (1 + qualifier_err)) - log_flow ) / 2.)


meas_data %>% 
  ggplot( aes( x = log_flow, y = meas_lflow_stderr, color = qualifier)) +
  geom_point()


```


Stage values are typically recorded to the nearest 100th of a foot. For streamgages that included many stage-flow measurements, this rounding may creates non-uniquesness that will result in averaging of flows at the same stage.  Given that different equal-stage measurments may have different timings and uncertainties, a small, zero-mean random value is added to each stage value to provide uniqueness.  


```{r make_stage_unique}
# Create version of log_stage with small uniform random number for uniqueness
#   Uniqueness ensures that each measurement is used even with stage values
#   rounded to the 100th of a foot.  The uniqueness helps in the application to
#   a matrix multiply in the Bayesian estimation.
#  Random seed is set to facilitate reproducibility


print(paste('There are',nrow(meas_data),'stage values, but only', 
            length(unique(meas_data$log_stage)), 'of these values are unique.'))

# Setting a random seed for reproducibility
set.seed(3472792)
meas_data <- meas_data %>% 
  mutate( log_stage_unique   = log_stage_round + runif(nrow(meas_data), -0.0000049, 0.0000049)) %>% 
  arrange(log_stage_unique)

```

### Summarize measurement data

Provides summary statistics for measurements grouped by log_stage_round


```{r summ_meas_data}

meas_summary <- meas_data %>% 
  group_by( log_stage_round ) %>% 
  dplyr::summarize( mean_log_flow = mean(log_flow),
                    sd_log_flow   = sd(log_flow),
                    ivar_log_flow = 1 / sd(log_flow)^2,
                    cnt_log_flow  = n())

meas_summary <- meas_summary %>% 
  mutate(stage_round = 10^log_stage_round,
  flow_mean = 10^mean_log_flow) %>%  
  dplyr::select(stage_round, everything())

table(meas_summary$cnt_log_flow)

kable(meas_summary) %>% 
  kable_styling() %>% 
  scroll_box(width = "700px", height = "500px")


```

### Plot empirical distributions of measured flow 

Plot indicates that the log10 of flows are approximately normally distributed.

```{r dist_flow}

meas_data %>% 
  ggplot( aes(log_flow) ) +
  geom_density(fill = 'blue', color = 'blue', alpha = 0.2, show.legend = TRUE) +
  stat_function(fun  = dnorm, 
                args = list( mean = mean(log10(meas_data$flow)),
                             sd   =   sd(log10(meas_data$flow))),
                  lwd = 1, color = 'purple', show.legend = TRUE) +
  theme_bw() + 
  xlab('Log_{10} of Measured Flow, in cubic feet per second') +
  ylab('Probability Density') +
  theme( legend.position = ('right')) +
  labs(title = paste0('Figure ', fig_no,'. Empirical Distribution of Measured Log_10 Flows with Normal Density Reference')) 
fig_no <- fig_no + 1

```

## Distribution of Stage


```{r dist_stage, eval = FALSE}

# Gamma distribution is questionable applicable because of arbitrary offset
# Fit gamma distribution
fit.gamma.logstage <- fitdist(meas_data$log_stage_unique - min(meas_data$log_stage_unique)+.0001,
                              distr = 'gamma', method = 'mle')

summary(fit.gamma.logstage)

meas_data %>% 
  ggplot( aes(log_stage_unique - min(log_stage_unique)),  color = 'blue' ) +
  geom_density(fill = 'blue', alpha = 0.2, show.legend = TRUE) +
  stat_function(fun   = dgamma, 
                args  = list(shape = fit.gamma.logstage$estimate[[1]],
                             rate  = fit.gamma.logstage$estimate[[2]])) +
                #              sd   =   sd(log10(measData$stage))),
                #  lwd = 1, color = 'purple', show.legend = TRUE) +
  theme_bw() + 
  xlab('Log10 of Measured Stage minus point of zero flow offset, in feet') +
  ylab('Probability Density') +
  labs(title = paste0('Figure ', fig_no, ' Empirical Distribution of log10 Measured Stages with Gamma Density Reference'),
       subtitle = 'All stages reduced to reflect point of zero flow.')

fig_no <- fig_no + 1

```



## Plot Rating and Measurements


```{r plot_rating_group_quality, fig.height = 7}

rating_data %>% 
  ggplot( aes(y = flow, x = stage)) +
  geom_line( color = 'blue') + 
  geom_point(data = meas_data, aes(x = stage_round, y = flow, 
                                  color = quality)) +
  theme_bw() +
  theme(legend.position = "bottom") +
 
  scale_y_continuous(name   = 'Flow, in cubic feet per second',
                     trans  = "log10",
                     breaks = c(1, 10, 50, 100, 200, 500, 1000, 2000, 4000, 6000)) +
  scale_x_continuous(name   = "Stage, in feet",
                     trans  = "log10",
                     breaks = c(seq(3, 12, by = 1))) +
  ggtitle(label = paste('Figure 5. Stage-flow relation', rating_no, 'at streamgage', site_no, station_nm))

```
## Plot Rating and Measurements with Control Characteristics

```{r plot_meas_control, fig.height = 7}

meas_data %>% 
  ggplot( aes(y = flow, x = stage_round)) +
  geom_point(aes(x = stage_round, y = flow, 
                 color = Control_Degree), size = 2, alpha = 0.5) +
    scale_colour_brewer() +
  
  geom_line(data = rating_data, aes(x = stage, y = flow), color = 'darkgreen', width = 2) + 
  
  # geom_vline(data = ratingKnots, aes(xintercept = stage_knots), 
  #            color = 'salmon', linetype = 'dashed' ) +
  
  theme_bw() +
  theme(legend.position = "bottom") +
 
  scale_y_continuous(name   = 'Flow, in cubic feet per second',
                     trans  = "log10",
                     breaks = c(1, 10, 50, 100, 200, 500, 1000, 2000, 4000, 6000)) +
  scale_x_continuous(name   = "Stage, in feet",
                     trans  = "log10",
                     breaks = c(seq(3, 12, by = 1))) +
  ggtitle(label = paste('Figure 1. Stage-flow relation at streamgage', site_no, station_nm,
                        'showing effects of Controls'))
```


## Frequency of measurement qualities

The quality of measurements may need to be recomputed based on the rating in place at the time of measurement.

```{r table_quality, eval = FALSE}

meas_data$flow_exp <- approx( rating_data$stage, rating_data$flow, xout = meas_data$stage, rule = 2 )$y

meas_data$flow_err <- (meas_data$flow - meas_data$flow_exp) / (meas_data$flow_exp) * 100.

# Frequency of measurement qualities
print(table(meas_data$Quality))


meas_data %>% 
  ggplot( aes( Quality, flow_err) ) +
  geom_boxplot() +
  xlab('Rated Quality of Measurement') +
  ylab('Percent error in flow measurement') +
  geom_hline( aes(yintercept = 0) , color = 'red', linetype = 'dashed' )


```


### GAM for Stage-Flow Rating


```{r gam_estimate_of_rating, fig.height = 7 }

# Put meas_data in date order

meas_data <- meas_data %>% 
  # arrange( meas_data$measurement_dt)
  arrange(meas_data$log_stage_round)

log_stage <- meas_data$log_stage_round 

# Compute GAM 
Gmodel  <- gam(meas_data$log_flow ~ s(log_stage))

plot(Gmodel)

print( summary(Gmodel) )

# Extract basis matrix with rows equal to the number of unique log_stage values and columns equal to the dim of the basis)
Xp      <- predict( Gmodel, type = 'lpmatrix')


GAM_coef_est <- coefficients(Gmodel)

# State-space model component
#  The initial value of the state vector is the coefficients of gam model
Xo <- as.numeric(GAM_coef_est)

GAM_coef_sd  <- sqrt(diag(Gmodel$Vp))
# Xp_df   <- data.frame(Xp)

# State-space model component
Q <- Gmodel$Vp


 as.data.frame(Xp) %>%
  gather( key = 'basis', value = 'smo_stage' ) %>% 
  mutate( basis_row = rep(1:776, 10),
          log_stage_unique = rep(meas_data$log_stage_unique, 10)) %>% 
   ggplot( aes( x = log_stage_unique, y = smo_stage, color = basis)) +
   geom_line() +
   theme_bw()  +
   theme( legend.position = 'bottom') +
   xlab('Log10 of Measured Stage, in feet') +
   ylab('Basis Coefficient') + 
   scale_x_continuous( limits = c(0.5, 1.1), breaks = seq(0.5, 1.1, by = 0.1)) + 
   labs( title = 'Figure 7. Relation between log(stage) and GAM Smoothing Spline' )
  

```

```{r daily_data}





# Create data.frame of daily values
daily_data <- data.frame( measurement_dt = seq( as.Date('1970-10-01'), as.Date('2019-09-30'), 'days'))


daily_data <- left_join( daily_data, meas_data[, c('measurement_dt', 'flow', 'log_stage_round', 
                                                   'log_flow', 'meas_lflow_stderr', 'qualifier',
                                                   'measurement_tm', 'tz_cd_reported', )] )


```
## Retrieve daily stage and flow values 

```{r get_daily_values}

daily_values_pub <- readNWISdv(site_no, c('00065', '00060'))

startDate <- "1970-01-01"


# daily_values_pub <- readNWISuv(site_no, c('00065', '00060'), startDate )

unit_values_pub  <- renameNWISColumns(daily_values_pub)


```




```{r }



```



## Inital state space model
Xt is the 10x 1 state vector of GAM parameters
Q  is the fixed 10X10 covariance matrix of paramaters
R  is an estimated constant measurement variance
Zt is the matrix of fixed measurement components (smooth spline terms)
Yt is sequential daily values of log_stage_round 

```{r marss_m1}

# Min and max indices of daily_values defines range of daily values for analysis
ndx_min  <-  27
ndx_max  <- 996

# Find indices of non-zero daily values of log_stage_round in analysis range
ndx_daily_stage <- which(!is.na(daily_data$log_stage_round[ndx_min:ndx_max]))

# Predict the intercept and slope components for the spline smooth 
Xp      <- predict(Gmodel, newdata = data.frame(log_stage = daily_data$log_stage_round[ndx_min:ndx_max][ndx_daily_stage]),
                   type = 'lpmatrix')

# Allocate matrix with NAs to contain Xp components for measured log_stage values to days of measurement
Xp_aug <- matrix(NA, nrow = length(seq(ndx_min:ndx_max)), ncol = ncol(Xp))
# Replace NA indicators with measured log_stage values 
Xp_aug[ndx_log_stage,] <- Xp

# Identify measured flows during daily flow period in analysis

# Find indices of measured flows in daily analysis range
ndx_meas_flow <- which( meas_data$measurement_dt >= daily_data$measurement_dt[ndx_min] &
                        meas_data$measurement_dt <= daily_data$measurement_dt[ndx_max])

Yt <- rep(NA, length(seq(27:996)))
Yt[ndx_log_stage] <- meas_data$log_flow[ndx_meas_flow] 

R = matrix( list( "R_sigma2"), 1, 1)



# State-Space Model Component Z
Zt = t(Xp_aug)

dim(Zt) <- c(1, ncol(Xp), length(Yt))
x0 = matrix(GAM_coef_est, nrow = ncol(Xp), ncol = 1)

model.dat = Yt
model.gen = list(Z = Zt, x0 = x0, R = R )

m1 <- MARSS(model.dat, model.gen)


```


### Random Component to log stage values modelled as gamma density



```{r Bayes_parameter_estimates_dgamma, fig.height = 6}

# Use quadratic approximation posterior distribution (quap) 
baye_gam_dgamma <- quap(
  alist(
    log_flow  ~ dnorm( mu , sigma ) ,
    mu       <- Xp %*% w ,
    w         ~ dnorm(0,1),
    sigma     ~ dgamma(1)
),
data=list( log_flow = meas_data$log_flow , Xp = Xp ) ,
start=list( w = rep( 0 , ncol(Xp) ) ) )


# Parameter estimates from Bayesian estimation
precis( baye_gam_dgamma, depth = 2, prob = 0.95, digits = 5)

# Extract parameter estimates
post_dgamma   <- extract.samples( baye_gam_dgamma, n = 1e4 )
# head(post_dnorm)

gammaPars <- fitdistr(post_dgamma$sigma,"gamma")
shapeML   <- gammaPars$estimate["shape"]
rateML    <- gammaPars$estimate["rate"]

xval <- seq(0,max(post_dgamma$sigma),len=length(post_dgamma$sigma))

data.frame(post_dgamma) %>% 
  ggplot( aes(sigma ) ) +
  geom_density( stat = 'density') +
  stat_function(aes(x = xval, y = ..y..), fun = dgamma, colour="brown", 
                n = length(post_dgamma$sigma), args = list(shape=shapeML, rate=rateML)) +
  scale_x_continuous( limits = c(0.055, 0.070)) + 
  labs( title = paste0('Figure 7. Distribution of Bayes dgamma parameter sigma with reference dgamma(shape = ',
                       format(shapeML, digits = 4),', rate = ', format(rateML, digits = 4), ')')) +
  theme_bw()


```


### Random Component to log stage values modelled as normal density


```{r Bayes_parameter_estimates_dnorm, fig.height = 6}

# Use quadratic approximation posterior distribution (quap) 
baye_gam_dnorm <- quap(
  alist(
    log_flow  ~ dnorm( mu , sigma ) ,
    mu       <- Xp %*% w ,
    w         ~ dnorm(0,1),
    sigma     ~ dnorm(1,1)
),
data=list( log_flow = meas_data$log_flow , Xp = Xp ) ,
start=list( w = rep( 0 , ncol(Xp) ) ) )


# Parameter estimates from Bayesian estimation
precis( baye_gam, depth = 2, prob = 0.95, digits = 5)

# Extract parameter estimates
post_dnorm   <- extract.samples( baye_gam, n = 1e4 )
head(post_dnorm)

# dens( post$sigma, norm.comp=TRUE , main="Distribution of sigma estimates" , show.HPDI=TRUE)


data.frame(post_dnorm) %>% 
  ggplot( aes(sigma )) +
  geom_density( stat = 'density', alpha = 0.25, color = 'blue', size = 1.5 ) +
  stat_function(fun = dnorm, 
                args = list( mean = mean(post_dnorm$sigma), 
                             sd   = sd(  post_dnorm$sigma)),   color = 'red',  size = 1.5, ) +
  # scale_color_manual(name = "Density Plot", values = c("Empirical" = 'blue' ) ) +
  scale_x_continuous( limits = c(0.0575, 0.070)) + 
  labs( title = paste0('Figure 6. Distribution of sigma for  Bayes dnorm parameter sigma with reference dnorm(mean = ',
                       format(mean(post_dnorm$sigma), digits = 4),', sd = ', 
                       format(  sd(post_dnorm$sigma), digits = 4), ')')) +
  theme_bw()
  


```

### Include qualifier info to account for varying uncertainty


```{r Bayes_parameter_estimates_uncert_x, fig.height = 6}
x0 = Xp[1:776,1]; x1 = Xp[1:776,2]; x2 = Xp[1:776,3]; x3 = Xp[1:776,4]; x4 = Xp[1:776, 5];
x5 = Xp[1:776,6]; x6 = Xp[1:776,7]; x7 = Xp[1:776,8]; x8 = Xp[1:776,9]; x9 = Xp[1:776,10];


# Use quadratic approximation posterior distribution (quap) 
baye_gam_cert_x_dnorm1 <- ulam(
  alist(
    lflow_meas   ~ dnorm( lflow_true , lflow_meas_sd ) ,
    vector[N]:lflow_true ~ dnorm(lflow_rate, lflow_true_sd),
    lflow_rate   <- b0 * x0 + b1 * x1 + b3 * x3 + b4 * x4 + b5 * x5 + b6 * x6 + b7 * x7 + b7 * x7 + b8 * x8 + b9 * x9,
    b0            ~ dnorm(  2.12203099, 0.002302916 ),
    b1            ~ dnorm( -0.31305639, 0.011874773 ),
    b2            ~ dnorm( -0.87836073, 0.085117576 ),
    b3            ~ dnorm( -0.23794731, 0.017755688 ),
    b4            ~ dnorm(  0.53667040, 0.069286763 ),
    b5            ~ dnorm(  0.08390154, 0.009977110 ),
    b6            ~ dnorm( -0.42387766, 0.065635509 ),
    b7            ~ dnorm( -0.25612051, 0.030899320 ),
    b8            ~ dnorm(  1.00733829, 0.200192569 ),
    b9            ~ dnorm(  0.97733403, 0.039447235 ),
    lflow_true_sd ~ dnorm(  0 , 1)
),
data=list( lflow_meas    = meas_data$log_flow, 
           x0 = x0, x1  = x1, x2 = x2, x3 = x3, x4 = x4, x5 = x5, x6 = x6, x7 = x7, x8 = x8, x9 = x9, 
           lflow_meas_sd = meas_data$meas_lflow_stderr, lflow_meas = meas_data$log_flow , N = nrow(meas_data)),
start=list(mu0 = 2.12203099,  mu1 = -0.31305639, mu2 = -0.87836073, mu3 = -0.23794731, mu4 =  0.53667040,
           mu5 = 0.08390154,  mu6 = -0.42387766, mu7 = -0.25612051, mu8 =  1.00733829, mu9 =  0.97733403,
           sd0 = 0.002302916, sd1 = 0.011874773, sd2 = 0.085117576, sd3 = 0.017755688, sd4 = 0.069286763,
           sd5 = 0.009977110, sd6 = 0.065635509, sd7 = 0.030899320, sd8 = 0.200192569, sd9 = 0.039447235),
iter = 5000, warmup = 1000, chains = 3, cores = 6)

# control= list(max_treedepth = 20, adapt_delta = 0.95))


# Parameter estimates from Bayesian estimation
precis( baye_gam_cert_x_dnorm, depth = 1, prob = 0.95, digits = 5)

# Extract parameter estimates
post_dnorm   <- extract.samples( baye_gam_cert_x_dnorm, n = 1e4 )
# head(post_dnorm)

# dens( post$sigma, norm.comp=TRUE , main="Distribution of sigma estimates" , show.HPDI=TRUE)


data.frame(post_dnorm) %>% 
  ggplot( aes(lflow_true_sd )) +
  geom_density( stat = 'density', alpha = 0.25, color = 'blue', size = 1.5 ) +
  stat_function(fun = dnorm, 
                args = list( mean = mean(post_dnorm$lflow_true_sd), 
                             sd   = sd(  post_dnorm$lflow_true_sd)),   color = 'red',  size = 1.5, ) +
  # scale_color_manual(name = "Density Plot", values = c("Empirical" = 'blue' ) ) +
  # scale_x_continuous( limits = c(0.0575, 0.070)) + 
  labs( title = paste0('Figure 6. Distribution of sigma for  Bayes dnorm parameter sigma with reference dnorm(mean = ',
                       format(mean(post_dnorm$lflow_true_sd), digits = 4),', sd = ', 
                       format(  sd(post_dnorm$lflow_true_sd), digits = 4), ')')) +
  theme_bw()
  
tmp <- precis( baye_gam_cert_x_dnorm, depth = 2, prob = 0.95, digits = 5)


resid <- tmp$mean[1:776] - meas_data$log_flow
plot( meas_data$log_flow, resid )
abline( a = 0, b = 0, col = 'red')

ggdistribution(dgamma, x = seq(0, 2, 0.02), shape = 2, scale = 0.2)
ggdistribution(dchisq, x = seq(0, 10, 0.02), df = 2.25)
```

## Re-estimate with better staring values


```{r }

dlist <- list(
  Q_obs = standardize( meas_data$log_flow ),
  Q_sd =  meas_data$ d$Divorce.SE / sd( d$Divorce ),
  S_obs = standardize( d$Marriage ),
  S_sd = d$Marriage.SE / sd( d$Marriage ),
  A = standardize( d$MedianAgeMarriage ),
  N = nrow(d)
)


baye_gam_uncert_x <- ulam(
  alist(
    lflow_mea  ~ dnorm( lflow_true , lflow_sd ) ,
    vector[N]:lflow_true ~ dnorm( lflow_rat, rat_sd),
    lflow_rat <- b0 * x0 + b1 * x1 + b3 * x3 + b4 * x4 + b5 * x5 + b6 * x6 + b7 * x7 + b7 * x7 + b8 * x8 + b9 * x9,
    lstage_mea ~
    vector[N]:lstage_true ~ 
      
    b0         ~ dnorm(mu0,sd0),
    b1         ~ dnorm(mu1,sd1),
    b2         ~ dnorm(mu2,sd2),
    b3         ~ dnorm(mu3,sd3),
    b4         ~ dnorm(mu4,sd4),
    b5         ~ dnorm(mu5,sd5),
    b6         ~ dnorm(mu6,sd6),
    b7         ~ dnorm(mu7,sd7),
    b8         ~ dnorm(mu8,sd8),
    b9         ~ dnorm(mu9,sd9),
    lflow_mea  ~ dnorm(lflow_est, lflow_sig),
    sigma      ~ dexp(1)
),
data=list( log_flow  = meas_data$log_flow, 
           x0 = x0, x1 = x1, x2 = x2, x3 = x3, x4 = x4, x5 = x5, x6 = x6, x7 = x7, x8 = x8, x9 = x9, 
           lflow_sig = meas_data$meas_lflow_stderr, N = length(log_flow)),
start=list(lflow_est = meas_data$log_flow ,
           mu0 = 2.12203099,  mu1 = -0.31305639, mu2 = -0.87836073, mu3 = -0.23794731, mu4 =  0.53667040,
           mu5 = 0.08390154,  mu6 = -0.42387766, mu7 = -0.25612051, mu8 =  1.00733829, mu9 =  0.97733403,
           sd0 = 0.002302916, sd1 = 0.011874773, sd2 = 0.085117576, sd3 = 0.017755688, sd4 = 0.069286763,
           sd5 = 0.009977110, sd6 = 0.065635509, sd7 = 0.030899320, sd8 = 0.200192569, sd9 = 0.039447235),
iter = 2000, warmup = 1000, chains = 3, cores = 6,
control=list(max_treedepth = 20, adapt_delta = 0.95))

data(gmcv)

dlist <- list(
  D_obs = standardize( d$Divorce ),
  D_sd = d$Divorce.SE / sd( d$Divorce ),
  M_obs = standardize( d$Marriage ),
  M_sd = d$Marriage.SE / sd( d$Marriage ),
  A = standardize( d$MedianAgeMarriage ),
  N = nrow(d)
)
m15.2 <- ulam(
  alist(
    D_obs ~ dnorm( D_est , D_sd ),
    vector[N]:D_est ~ dnorm( mu , sigma ),
    mu <- a + bA*A + bM*M_est[i],
    M_obs ~ dnorm( M_est , M_sd ),
    vector[N]:M_est ~ dnorm( 0 , 1 ),
    a ~ dnorm(0,0.2),
    bA ~ dnorm(0,0.5),
    bM ~ dnorm(0,0.5),
    sigma ~ dexp( 1 )
  ) , data=dlist , chains=4 , cores=4 )

```



```{r }

# Correlation among parameters
ggcorrplot( cov2cor( vcov( baye_gam_dnorm)), type = 'lower', 
                     insig = 'blank',  lab = TRUE, digits = 2)

```


# Compare Bayesian and GAM model parameter estimates


```{r compare_gam_bayes_param}
coef_gam <- data.frame( gam_coef = coef(Gmodel), baye_coef = coef(baye_gam)[1:10], 
                        id     = c('W1', 'W2', 'W3', 'W4', 'W5', 'W6', 'W7', 'W8', 'W9', 'W10'))

coef_gam %>% 
  ggplot( aes( x = gam_coef, y = baye_coef, label = id)) +
  geom_point( color = 'blue', size = 2) +
  geom_abline( slope = 1, intercept = 0, color = 'red') +
  geom_label_repel(aes(label = id), size = 2,
                     box.padding = 0.25, 
                   point.padding = 0.25,
                   label.padding = 0.25,
                  segment.color = 'grey50') +
  theme_classic() +
  labs( x = 'GAM Parameter Estimate',  y = 'Bayesian Parameter Estimate',
        title = 'Figure 4. Compare parameters estimated with and without random component in log stage values')
  
```
```{r extract_bayes}
# Extract parameter estimates
post   <- extract.samples( baye_gam, n = 1e4 )
head(post, n = 5)

dens( post$sigma, norm.comp=TRUE , main="Distribution of sigma estimates" , show.HPDI=TRUE)


data.frame(post) %>% 
  ggplot( aes(sigma ) ) +
  geom_density( stat = 'density') +
  stat_function(fun = dnorm, 
                args = list( mean =  mean(post$sigma), sd = sd(post$sigma)), color = 'red' )

```


```{r }
# Use quadratic approximation posterior distribution (quap) 
baye_gam_dexp <- quap(
  alist(
    log_flow  ~ dnorm( mu , sigma ) ,
    mu       <- Xp %*% w ,
    w         ~ dnorm(0,1),
    sigma     ~ dexp(1)
),
data=list( log_flow = meas_data$log_flow , Xp = Xp ) ,
start=list( w = rep( 0 , ncol(Xp) ) ) )


# Parameter estimates from Bayesian estimation
precis( baye_gam, depth = 2, prob = 0.95, digits = 5)



# Link uses the quap approximation, and samples from the posterior distribution, and computes 
#   µ for each case in the data and sample from the posterior distribution.
mu     <- link( baye_gam )
str(mu)

# Generage a sequence of log_stages for analysis

log_stage_seq <- seq( from = min(log_stage), to = max(log_stage), length.out = 50 )

data_seq      <- data.frame( log_stage = log_stage_seq )

Xp_seq        <- predict( Gmodel, newdata = data_seq, type = 'lpmatrix')

mu_seq        <- link( baye_gam, data = list(Xp = Xp_seq))[,1:50]

tmp025 <- as.data.frame(mu_seq) %>% 
  dplyr::summarize_all(list(q025 = quantile), probs = 0.025)
tmp975 <- as.data.frame(mu_seq) %>% 
  dplyr::summarize_all(list(q025 = quantile), probs = 0.975)
tmpave <- as.data.frame(mu_seq) %>% 
  dplyr::summarize_all(list(mean = mean))



baye_rating <- data.frame(t(rbind(log_stage_seq,as.matrix(tmp025), as.matrix(tmpave), as.matrix(tmp975))))

colnames(baye_rating) <- c('Stage','q025', 'mean', 'q975')
rownames(baye_rating) <- 1:50

baye_rating %>% 
  ggplot( aes( x = Stage, y = mean)) +
  geom_line(   )


mu_seq  <- link( baye_gam, data = data.frame( Xp = Xp_new ) )


# Compare bayesian estimated parameters with gam model estimates





```


## Evaluate spline model

```{r eval_bs, eval = FALSE}

post    <- extract.samples( n_bs_reg_02 )

posta   <- mean(post$a)
posts   <- mean(post$s)
postw   <- apply( post$w, 2, mean)

mu   <- link( n_bs_reg_02 )

mu_PI <- apply(mu, 2, PI, 0.95)

plot( measData$log_stage, measData$log_flow, col = col.alpha(rangi2, 0.5))

shade( mu_PI, measData$log_stage, col = col.alpha("black", 0.5))

stage.seq <- seq(from = min(measData$log_stage), to = max(measData$log_stage), length = 100)

stage.mu  <- link( n_bs_reg_02, data = data.frame(log_stage = stage.seq))

# Find minimum difference between entry in 
meas_data_ndx <- rep(NA, length(stage.seq))
for (i in 1:length(stage.seq)){
  adiff <- abs(stage.seq[i] - log_stage)
  meas_data_ndx[i] <- which.min(adiff)
  print( paste(i, ndx, adiff[ndx]))
}

# summarize the distribution of stage.mu
mu.mean <- apply( stage.mu[,meas_data_ndx] , 2 , mean )
mu.PI   <- apply( stage.mu[,meas_data_ndx] , 2 , PI , prob=0.95 )

smooth.spline(x = log_stage, y = log_flow)
# , w = NULL, df = measData, lambda = NULL, cv = FALSE,
#               all.knots = FALSE, nknots = .nknots.smspl,
#               keep.data = TRUE, df.offset = 0, penalty = 1,
#               control.spar = list(), tol = 1e-6 * IQR(log_stage), keep.stuff = FALSE)

require(graphics)
plot(dist ~ speed, data = cars, main = "data(cars)  &  smoothing splines")
cars.spl <- with(cars, smooth.spline(speed, dist))
cars.spl
## This example has duplicate points, so avoid cv = TRUE

lines(cars.spl, col = "blue", lwd = 2)
ss10 <- smooth.spline(cars[,"speed"], cars[,"dist"], df = 10)
lines(ss10, lty = 2, col = "red")
legend(5,120,c(paste("default [C.V.] => df =",round(cars.spl$df,1)),
               "s( * , df = 10)"), col = c("blue","red"), lty = 1:2,
       bg = 'bisque')



plot(log_flow ~ log_stage, data = measData, main = "data(cars)  &  smoothing splines",
     col = 'grey')
rate1.spl <- with(measData, smooth.spline(log_stage, log_flow))
rate1.spl
## This example has duplicate points, so avoid cv = TRUE

lines(rate1.spl, col = "blue", lwd = 2)
rate2.spl <- smooth.spline(measData[,"log_stage"], measData[,"log_flow"], df = 8)
lines(rate2.spl, lty = 2, col = "red")

legend(5,120,c(paste("default [C.V.] => df =",round(cars.spl$df,1)),
               "s( * , df = 10)"), col = c("blue","red"), lty = 1:2,
       bg = 'bisque')


# plot raw data
# fading out points to make line and interval more visible
plot( log_flow ~ log_stage , data=measData , col=col.alpha(rangi2,0.5) )
# plot the MAP line, aka the mean mu for each weight
lines( stage.seq , mu.mean )
# plot a shaded region for 89% PI
shade( mu.PI , stage.seq )


# pred      <- link( m_bs_reg_01, data = measData  )
# 
# 
# post <- extract.samples(m_bs_reg_01)
# w <- apply( post$w , 2 , mean )
# plot( NULL , xlim=range(tmp$log_stage) , ylim=c(-5,5) ,
# xlab="year" , ylab="basis * weight" )
# for ( i in 1:ncol(Bspline) ) lines( tmp$log_stage , w[i]*Bspline[,i] )
# 
# mu <- link( m_bs_reg_01 )
# mu_PI <- apply(mu,2,PI,0.95)
# plot( tmp$log_stage , tmp$log_flow , col=col.alpha(rangi2,0.3) , pch=16 )
# shade( mu_PI , tmp$log_stage , col=col.alpha("black",0.5) )


```


```{r gam_rate, eval = FALSE}

measData <- measData %>% 
  mutate( log_flow = log10(flow), log_stage = log10(stage))

log_stage <- measData$log_stage_
log_flow  <- measData$log_flow

gam_spline <- gam(log_flow ~ s(log_stage), data = measData, method = "ML")
measData$gam_hat <- gam_spline$fitted.values


measData %>% 
  ggplot( aes( x = log_stage, y = log_flow)) +
  geom_point() +
  geom_line( aes( x = log_stage, y = gam_hat), color = 'red')


plot(gam_spline)

m_bs_reg_01 <- quap(
  alist(
    log_flow ~  dnorm( mu, sigma ),
    mu       <- a + Bspline %*% w,
    a        ~  dnorm( 2, 1.5 ),
    w        ~  dnorm( 0, 1   ),
    sigma    ~  dexp(1)
  ),
  data = list( log_flow = log_flow, log_stage = log_stage, B = Bspline),
  start = list(w = rep( 0, ncol(Bspline)))
)

precis( m_bs_reg_01, depth = 2, prob = 0.95, digits = 5)




```


}