---
title: "Bayesian Estimation of a Generalized Additive Model to Describe the Stage-Discharge Relation"
author: "Dave Holtschlag"
date: "`r format(Sys.Date(), '%A %b %d, %Y') `"
output: 
  html_document:
    toc: true
    toc_depth: 3
bibliography: bibliography.bib
---

## Abstract 
A Bayesian approach is developed to estimate parameters of a generalized additive model (GAM) [@Wood2006] to describe the dynamic relation between stage and discharge (flow) at a streamgage.  The GAM model provides a basis function and parameter estimates to describe the smooth, monotonically increasing relation between stage and flow using data on direct measurements of stage and flow commonly used to develop a so called rating curve. Bayesian estimation then uses the GAM basis function and re-estimates parameters to explicitly account for the uncertainty in the stage measurements. Data for USGS streamgage 04112500 Red Cedar River at East Lansing, MI are used to illustrate the approach and results of the generally applicable method.  Computations are implemented in the R programming environment [@cite_R].

### Setup computational environment

```{r setup, include=FALSE}
library( "rethinking" )
options(mc.cores = parallel::detectCores())
library(dataRetrieval)
library(lubridate)
library(tidyverse)
library(ggthemes)
library(splines)
library(mgcv)
library(fitdistrplus)
library(stringr)
library(kableExtra)
library(ggrepel)
library(ggcorrplot)
# strcmp and other Matlab like functions
library(pracma)
# Package expss contains the function vlookup
library(expss)
# Package to plot distribution
library(ggfortify)
# Multivariate Autoregressive State Space Model
library(MARSS)
library(matrixcalc)
# The broom package takes the messy output of built-in functions in R, such as lm, nls, or t.test, and turns them into tidy data frames.
library(broom)
library(gganimate)
library(gifski)
knitr::opts_chunk$set(echo = TRUE)
```

## Retrieve Streamgaging Site Inforamtion 

```{r specify_gage}

# initialize counter for figures
fig_no <- 1

# Retrieve gage info for specified site Number
site_no    <- '04122500'

gage_info  <- readNWISsite(site_no)

# Standardize field names
gage_info  <- renameNWISColumns(gage_info)

# 
gage_info$station_nm <- str_to_title( gage_info$station_nm )

gage_info[1,c('site_no','station_nm','drain_area_va', 'dec_lat_va', 'dec_long_va', 'alt_va','alt_datum_cd')] %>% 
  kable( caption = 'Table 1. Summary streamgage information' ) %>% 
  kable_styling()

```

### Retrieve Stage-Flow Rating

Reads and plots rating curves 19, 20, and 21 for streamgage 04122500 Pere Marquette River at Scottville, MI and plots the curves

```{r read_rating_data, fig.width = 9}

# Rating curve at the selected site.  The 'exsa' parameter is an extended table
#   provide detailed info on stage discharge

rating18.0 <- read.csv('Data/04122500/04122500_rating_18.0.txt', header = FALSE, sep = '\t') 
colnames( rating18.0 ) <- c('rated_flow_18.0', 'stage')

rating19.0 <- read.csv('Data/04122500/04122500_rating_19.0.txt', header = FALSE, sep = '\t') 
colnames( rating19.0 ) <- c('rated_flow_19.0', 'stage')

rating20.0 <- read.csv('Data/04122500/04122500_rating_20.0.txt', header = FALSE, sep = '\t') 
colnames( rating20.0 ) <- c('rated_flow_20.0', 'stage')

rating20.1 <- read.csv('Data/04122500/04122500_rating_20.1.txt', header = FALSE, sep = '\t') 
colnames( rating20.1 ) <- c('rated_flow_20.1', 'stage')

rating21.0 <- read.csv('Data/04122500/04122500_rating_21.0.txt', header = FALSE, sep = '\t') 
colnames( rating21.0 ) <- c('rated_flow_21.0', 'stage')


rating_wide<- rating19.0   %>% 
  left_join(  rating20.0 ) %>% 
  left_join(  rating20.1)  %>% 
  left_join(  rating21.0 ) %>% 
  dplyr::select(stage, everything() )
  
rating_long <- rating_wide%>% 
  gather( key = 'rating_no', value = 'rated_flow', 
          rated_flow_19.0, rated_flow_20.0, rated_flow_20.1, rated_flow_21.0)

rating_long %>% 
  ggplot( aes( x = rated_flow, y = stage,  group = rating_no)) +
  geom_line( aes( color = rating_no, linetype = rating_no ), size = 1.5, alpha = 0.50) +
  theme_bw() +
  theme( legend.position = 'bottom' ) +
  xlab( 'Flow, in cubic feet per second' ) +
  ylab( 'Stage, in feet above gage datum' ) +
  labs( title =  'Figure 1. Selected rating curves at 04122500 Pere Marquette River at Scottville, MI' )
  
```


## Develop a generatized additive model (GAM) for rating 19

```{r gam_rating19, fig.width = 9}



# Estimate generalized additive model of stage - discharge rating 19 
gam_rate_19 <- gam( rated_flow_19 ~ s(stage), data = rating19 )

# Print summary of gam_rate_19
print(summary(gam_rate_19))

# glance, tidy, and augment from boom package provide alternative summaries of the model
# glance(gam_rate_19)
# tidy(gam_rate_19)
# augment(gam_rate_19) %>% kable() %>% kable_styling()


gam_rate_19_pred     <- predict(gam_rate_19, se.fit = TRUE)

rating19$gam_flow    <- gam_rate_19_pred$fit
rating19$se_fit      <- gam_rate_19_pred$se.fit
rating19$error_025   <- rating19$rated_flow_19 - (gam_rate_19_pred$fit + qnorm(0.025) * gam_rate_19_pred$se.fit)   
rating19$error_975   <- rating19$rated_flow_19 - (gam_rate_19_pred$fit + qnorm(0.975) * gam_rate_19_pred$se.fit) 
rating19$flow_error  <- gam_rate_19$residuals
rating19$pct_error   <- gam_rate_19$residuals / gam_rate_19$fitted.values * 100

rating19 %>% 
  # dplyr::arrange( stage ) %>% 
  ggplot(    aes( y = pct_error, x = stage  )) +
  geom_point( color = 'red', size = 0.5) +
  # geom_line( aes( y = error_025), color = 'salmon') +
  # geom_line( aes( y = error_975), color = 'salmon') +
  # geom_ribbon( aes( ymin = error_025, ymax = error_975), fill = 'blue', alpha = 0.25 ) +
  scale_y_continuous( breaks = seq( from = -0.5, to = 2, by = 0.5)) +
  ylab( 'Flow Error, in percent') +
  xlab( 'Stage, in feet') +
  theme_bw() +
  geom_hline( yintercept = 0, color = 'blue', linetype = 'dashed') +
  theme( legend.position = 'bottom') +
  labs( title = paste('Figure 2. Errors in Flow for GAM of rating curve 19 at', site_no, station_nm))

```




### Retrieve Stage-Flow Measurement Pairs

```{r read_meas_stage_flow}

meas_data   <- readNWISmeas(site_no, expanded = TRUE, convertType = TRUE)

# Standardize field names
meas_data  <- renameNWISColumns(meas_data)


control_type <- 
  recode_factor( meas_data$control_type_cd, 
                 'Clear'    = 'Clear',
          'VegetationLight' = 'Light Vegetation',
          'DebrisLight'     = 'Light Debris',   
          'IceShore'        = 'Shore Ice',
          'DebrisModerate'  = 'Moderate Debris',
          'IceCover'        = 'Ice Cover', .ordered = TRUE ) 

meas_data$control_type <- control_type


meas_data <- meas_data %>% 
  rename('stage'          = 'gage_height_va',
         'flow_meas'      = 'discharge_va',
         'qualifier'      = 'measured_rating_diff') %>% 
  filter(qualifier       != 'Unspecified' | qualifier == !is.na(qualifier)) %>%
  mutate('qualifier'      =  ordered(qualifier, levels = c('Excellent', 'Good', 'Fair', 'Poor')))


# ---------------------------------
# Code  Description
# ---------------------------------
# E     Excellent    The data is within 2% (percent) of the actual flow {95 percent of the time}
# G     Good         The data is within 5% (percent) of the actual flow
# F     Fair         The data is within 8% (percent) of the actual flow
# P     Poor         The data are >8% (percent) of the actual flow

meas_data <- meas_data %>% 
  mutate( se_pct = case_when( qualifier == 'Excellent' ~  1.0,
                              qualifier == 'Good'      ~  2.5,
                              qualifier == 'Fair'      ~  4.0,
                              qualifier == 'Poor'      ~ 10.0),
          se_flow = se_pct/100 * flow_meas ) 



# Populate field to combine similar Control attributes

meas_data$control_degree  = case_when(str_detect(meas_data$control, 'Heavy')       ~ 'Heavy',
                                      str_detect(meas_data$control, 'Moderate')    ~ 'Moderate',
                                      str_detect(meas_data$control, 'Light'   )    ~ 'Light',
                                      str_detect(meas_data$control, 'Clear'   )    ~ 'Clear',
                                      str_detect(meas_data$control, 'Ice'     )    ~ 'Ice',
                                      str_detect(meas_data$control, 'Unspecifed')  ~ 'Unspecifed')

meas_data$control_degree = ordered(meas_data$control_degree,
                                  levels = c('Clear', 'Light','Moderate', 'Heavy', 
                                             'Ice_Affected','Unspecifed'))



# Only cases where flow and stage are measured
meas_data <- meas_data[complete.cases(meas_data$stage, meas_data$flow),]


table( meas_data$control_type, meas_data$qualifier ) %>%
  kable( digits = 4, 
         caption = 'Table 1. Frequency of flow measurement qualifiers and controls at 04122500 Pere Marquette River at Scottville, MI.') %>%
  kable_styling() %>% 
  add_header_above(c("Degree of Control", "Qualifier", 'Qualifier', "Qualifier", "Qualifier"))


df <- data.frame( 'days_between_measurements' = diff(meas_data$measurement_dt))

df %>% 
  ggplot( aes( x = days_between_measurements )) + 
  geom_density( fill = 'blue', alpha = 0.25 ) +
  theme_bw() +
  geom_vline( xintercept = 42.75, color = 'black', linetype = 'dashed' ) +
  xlab( 'Time between flow measurements, in days' ) +
  ylab( 'Probability Density') +
  labs( title = 'Figure 2. Distribution of times between consecutive flow measurements at \n04122500 Pere River at Scottville, MI from 1985 to 2019' )
  
  


```

## Read unit stage values for 04122500 Pere Marquette River at Scottville, MI

```{r unit_daily_stage}

stage_unit <- read.csv('Data/04122500/stage/Gage_height.ft@04122500.EntireRecorda.csv', header = TRUE, sep = ',',
                       stringsAsFactors = FALSE, comment = '#')

# Convert from character field to datetime referenced to EST
stage_unit$datetime <- as_datetime(stage_unit$Timestamp..UTC.05.00, tz = 'America/New_York')

# Summarize unit stage to daily mean stage
stage_day <- stage_unit %>% 
  rename( stage = Value ) %>% 
  group_by(day = floor_date( datetime, 'day')) %>% 
  summarize( stage_mean_day  = mean( stage, na.rm = TRUE ),
             n         = n()) %>% 
  ungroup() %>% 
  arrange( day ) %>% 
  mutate( delta_day = as.numeric(day - lag( day ), units = 'days' ),
          day_date  = as.Date( day ))

# joint stage_day with full daily value


# Measured flows and daily stages for the interval when rating 19 was active
#  Note: Discrete measurement timing is approximated to days

df_stage_flow_day_19 <- data.frame(day = seq.Date(from = date('1994-10-25'), to = date('2019-11-01'), by = '1 day'))


df_stage_flow_day_19 <-df_stage_flow_day_19 %>% 
  left_join( meas_data[, c('measurement_dt', 'flow_meas', 'qualifier', 'meas_type', 'se_flow')], 
             by = c( 'day'  = 'measurement_dt')) %>% 
  left_join( stage_day[, c('day_date', 'stage_mean_day')], by = c( 'day' = 'day_date' )) %>% 
  filter( day >= as.Date('1994-10-25', format = '%Y-%m-%d') & day <= as.Date('2019-11-01', format = '%Y-%m-%d' ))

# find indices of missing stage_mean_day values
ndx_ina <- which( is.na(df_stage_flow_day_19$stage_mean_day ))
df_stage_flow_day_19$stage_mean_day[ndx_ina] <- 2.110


```


## Inital state space model
Yt is measured flow referenced to the day of measurement during the period when rating 19 was active. Most values are NA
Xt is the 10 x 1 state vector of GAM parameters used to approximate rating 19
Q  is the fixed 10X10 covariance matrix of paramaters
R  is an estimated constant measurement variance
Zt is the matrix of fixed measurement components 
Yt is sequential daily values of log_stage_round 


Note: A multiplicative form for the Q matrix can be specified as:
Q <- matrix(list('qb', 0, 0, 0, 0, 0, 0, '10*qb'), 3, 3)

```{r marss_m1}

# stage_flow_day_19 as 1 x T matrix
Yt <- matrix(df_stage_flow_day_19$flow_meas, nrow = 1)

# Predict the intercept and slope components for the spline smooth 
Zt <- t(predict( gam_rate_19, newdata = data.frame( stage = df_stage_flow_day_19$stage_mean_day),
                                           type = 'lpmatrix'))

rownames(Zt) <- NULL



# Vp is the estimated covariance matrix for the parameters. This is a Bayesian
# posterior covariance matrix that results from adopting a particular Bayesian
# model of the smoothing process. Paricularly useful for creating
# credible/confidence intervals

#   Under ML or REML smoothing parameter estimation it is possible to correct
#   the covariance matrix Vp for smoothing parameter uncertainty. This is the
#   corrected version

# Qn  <- as.matrix( as.numeric(gam_rate_19$Vp), 10, 10)
Qn  <- gam_rate_19$Vp

# Qc is a character matrix formed from a numeric matrix
#   Initialize Qc matrix
Qc <- matrix(NA, 10, 10)

Q = "diagonal and unequal"

zt <- array(Zt, dim = c(1, 10, 9142))

Z  <- zt


A  <- 'zero'
# R  <- matrix(list(0),1,1); diag(R) = "r"
R  <- array(0, dim = c(1, 1, 9142))
ndx_ian <- which(!is.na(Yt))
# R[1,ndx_ian] <- paste0('"r ', as.character(Yt[ndx_ian]),'"', sep='')
R[1,1, ndx_ian] <- (df_stage_flow_day_19$se_flow[ndx_ian])^2
U  <- 'zero'
# Q  <-  Qn * 0.02
# Q <- matrix(list('qicpt' ,0,0,0,0,0,0,0,0,0,0,
#                  'qknot' ,0,0,0,0,0,0,0,0,0,0,
#                  'qknot' ,0,0,0,0,0,0,0,0,0,0,
#                  'qknot' ,0,0,0,0,0,0,0,0,0,0,
#                  'qknot' ,0,0,0,0,0,0,0,0,0,0,
#                  'qknot' ,0,0,0,0,0,0,0,0,0,0,
#                  'qknot' ,0,0,0,0,0,0,0,0,0,0,
#                  'qknot' ,0,0,0,0,0,0,0,0,0,0,
#                  'qknot' ,0,0,0,0,0,0,0,0,0,0,
#                  'qknot'), 10, 10)

Q <- matrix(list('vvincpt',     0,      0,      0,       0,       0,       0,       0,        0,       0,
                      0,  'vvkk',     0,      0,       0,       0,       0,       0,        0,       0,
                      0,      0, 'vvkk',     0,       0,       0,       0,       0,        0,       0,
                      0,      0,      0,  'vvkk',      0,       0,       0,       0,        0,       0,
                      0,      0,      0,      0,   'vvkk',      0,       0,       0,        0,       0,
                      0,      0,      0,      0,       0,   'vvkk',      0,       0,        0,       0,
                      0,      0,      0,      0,       0,       0,   'vvkk',      0,        0,       0,
                      0,      0,      0,      0,       0,       0,       0,   'vvkk',       0,       0,
                      0,      0,      0,      0,       0,       0,       0,       0,    'vvkk' ,     0,
                      0,      0,      0,      0,       0,       0,       0,       0,        0,   'vvkk'),
            10, 10)

# Q <- matrix(list('qicpt','qincv',     0,      0,       0,       0,       0,       0,        0,       0,
#                  'qincv','qknot', 'qcvar',    0,       0,       0,       0,       0,        0,       0,
#                       0, 'qcvar', 'qknot', 'qcvar',    0,       0,       0,       0,        0,       0,
#                       0,      0,  'qcvar', 'qknot', 'qcvar',    0,       0,       0,        0,       0,
#                       0,      0,      0,   'qcvar', 'qknot', 'qcvar',    0,       0,        0,       0,
#                       0,      0,      0,      0,    'qcvar', 'qknot', 'qcvar',    0,        0,       0,
#                       0,      0,      0,      0,       0,    'qcvar', 'qknot', 'qcvar',     0,       0,
#                       0,      0,      0,      0,       0,       0,    'qcvar', 'qknot',  'qcvar',    0,
#                       0,      0,      0,      0,       0,       0,       0,    'qcvar',  'qknot' ,'qcvar',
#                       0,      0,      0,      0,       0,       0,       0,       0,     'qcvar', 'qknot'),
#             10, 10)



Q <- matrix(list('vvii', 'cvik', 'cvik', 'cvik', 'cvik', 'cvik', 'cvik', 'cvik',  'cvik', 'cvik',
                 'cvik', 'vvkk', 'cvkk', 'cvkk', 'cvkk', 'cvkk', 'cvkk', 'cvkk',  'cvkk', 'cvkk',
                 'cvik', 'cvkk', 'vvkk', 'cvkk', 'cvkk', 'cvkk', 'cvkk', 'cvkk',  'cvkk', 'cvkk',
                 'cvik', 'cvkk', 'cvkk', 'vvkk', 'cvkk', 'cvkk', 'cvkk', 'cvkk',  'cvkk', 'cvkk',
                 'cvik', 'cvkk', 'cvkk', 'cvkk', 'vvkk', 'cvkk', 'cvkk', 'cvkk',  'cvkk', 'cvkk',
                 'cvik', 'cvkk', 'cvkk', 'cvkk', 'cvkk', 'vvkk', 'cvkk', 'cvkk',  'cvkk', 'cvkk',
                 'cvik', 'cvkk', 'cvkk', 'cvkk', 'cvkk', 'cvkk', 'vvkk', 'cvkk',  'cvkk', 'cvkk',
                 'cvik', 'cvkk', 'cvkk', 'cvkk', 'cvkk', 'cvkk', 'cvkk', 'vvkk',  'cvkk', 'cvkk',
                 'cvik', 'cvkk', 'cvkk', 'cvkk', 'cvkk', 'cvkk', 'cvkk', 'cvkk',  'vvkk', 'cvkk',
                 'cvik', 'cvkk', 'cvkk', 'cvkk', 'cvkk', 'cvkk', 'cvkk', 'cvkk',  'cvkk', 'vvkk'),
            10, 10)

inits = list(Q=matrix(c( 1300 ,  0.01 ,  0.01 ,  0.01 ,  0.01 ,  0.01 ,  0.01 ,  0.01 ,   0.01 ,  0.01 ,
                         0.01 ,  0.10 ,  0.01 ,  0.01 ,  0.01 ,  0.01 ,  0.01 ,  0.01 ,   0.01 ,  0.01 ,
                         0.01 ,  0.01 ,  0.10 ,  0.01 ,  0.01 ,  0.01 ,  0.01 ,  0.01 ,   0.01 ,  0.01 ,
                         0.01 ,  0.01 ,  0.01 ,  0.10 ,  0.01 ,  0.01 ,  0.01 ,  0.01 ,   0.01 ,  0.01 ,
                         0.01 ,  0.01 ,  0.01 ,  0.01 ,  0.10 ,  0.01 ,  0.01 ,  0.01 ,   0.01 ,  0.01 ,
                         0.01 ,  0.01 ,  0.01 ,  0.01 ,  0.01 ,  0.10 ,  0.01 ,  0.01 ,   0.01 ,  0.01 ,
                         0.01 ,  0.01 ,  0.01 ,  0.01 ,  0.01 ,  0.01 ,  0.10 ,  0.01 ,   0.01 ,  0.01 ,
                         0.01 ,  0.01 ,  0.01 ,  0.01 ,  0.01 ,  0.01 ,  0.01 ,  0.10 ,   0.01 ,  0.01 ,
                         0.01 ,  0.01 ,  0.01 ,  0.01 ,  0.01 ,  0.01 ,  0.01 ,  0.01 ,   0.10 ,  0.01 ,
                         0.01 ,  0.01 ,  0.01 ,  0.01 ,  0.01 ,  0.01 ,  0.01 ,  0.01 ,   0.01 ,  0.10 ),
            10, 10))



X0 <- as.matrix( as.numeric(coefficients(gam_rate_19)), 10, 1)
V0 <- Qn

model.gen=list(Z=Z,A=A,R=R,U=U,Q=Q,x0=X0,V0=V0,tinitx=0)
```



```{r estimate_marss, eval = FALSE}

# Save command for the state space model
# saveRDS(kemfit_por, file = 'Models/kemfit_por.rds')
saveRDS(kemfit_por_Qdiag_unequal, file = 'Models/kemfit_por_Qdiag_unequal.rds')



# Check if kemfit model exists and read, else recompute
if( file.exists('Models/kemfit_por.rds')){
  kemfit_por <- readRDS('Models/kemfit_por.rds')
} else {
  kemfit_por_Qdiag_unequal <- MARSS(Yt, model = model.gen, control = list(maxit = 5000 ))
}

# Computes states (xtT, xtt, xtt1, variances (VtT, Vtt, vtt1), and similar stats on kemfit object)
kemfit_por_kfss <- MARSSkfss(kemfit_por)

# Error: Stopped in MARSSboot() due to problem(s) with function arguments.
#   paramCIs_innovation = MARSSparamCIs(kemfit_por, method = 'innovations')


# Innovations bootstrapping uses the innovations resampling and can only be done if there are no missing values in the data.
#   MARSSboot( kemfit_por, nboot = 1000, output = 'parameters', sim = 'innovations', param.gen = 'MLE' )



```
## Results from state-space modeling

Success! abstol and log-log tests passed at 571 iterations.
Alert: conv.test.slope.tol is 0.5.
Test with smaller values (<0.1) to ensure convergence.

MARSS fit is
Estimation method: kem 
Convergence test: conv.test.slope.tol = 0.5, abstol = 0.001
Estimation converged in 571 iterations. 
Log-likelihood: -583.9928 
AIC: 1171.986   AICc: 1172.117   
 
        Estimate
Q.qicpt  172.694
Q.qknot    0.105
Initial states (x0) defined at t=0

Standard errors have not been calculated. 
Use MARSSparamCIs to compute CIs and bias estimates.

## GAM for stage_flow_day_19 meaurements

```{r eval_state_space, fig.width = 9}

# Store results of MARSS modeling
df_stage_flow_day_19$ytT    <- as.vector(kemfit_por$ytT)
df_stage_flow_day_19$ytT.se <- as.vector(kemfit_por$ytT.se)

# Compare rating with kfgam estimates

stage_knots_19 <- predict( gam_rate_19, newdata = data.frame( stage = df_stage_flow_day_19$stage_mean_day),
                                           type = 'lpmatrix')

stage_knots_19 <- predict( gam_rate_19, type = 'lpmatrix')



stage_knots_19 <- cbind(rating19$stage, stage_knots_19) 
colnames(stage_knots_19)[1] <- 'stage'

tmp <- meas_data %>% 
  left_join( df_stage_flow_day_19, by = c('measurement_dt' = 'day') )

ndx_kfss <- which( df_stage_flow_day_19$day %in% meas_data$measurement_dt )
flow_xtT <- matrix(NA, nrow = 731, ncol = length(ndx_kfss) )

for (i in 1:length(ndx_kfss)){
  flow_xtT[ , i ] <- stage_knots_19_matrix %*% matrix(kemfit_por_kfss$xtT[ 1:10, ndx_kfss[ i ]], 10, 1)
}

sd_stage <- apply(flow_xtT, 1, FUN = identity, na.rm = TRUE)

plot(sd_stage)

stage_knots_19_matrix <- as.matrix(stage_knots_19[,2:11])


plot(sd_stage, rating19.0$stage)

state_coefs_19.0  <- matrix(kemfit_por_kfss$xtT[ 1:10,    1], 10, 1)
state_coefs_20.0  <- matrix(kemfit_por_kfss$xtT[ 1:10, 5972], 10, 1)
state_coefs_20.1  <- matrix(kemfit_por_kfss$xtT[ 1:10, 6463], 10, 1)
state_coefs_21.0  <- matrix(kemfit_por_kfss$xtT[ 1:10, 6918], 10, 1)
state_coefs_22.0  <- matrix(kemfit_por_kfss$xtT[ 1:10, 8746], 10, 1)

rating19$flow_kfgam19.0 <- stage_knots_19_matrix %*% state_coefs_19.0
rating19$flow_kfgam20.0 <- stage_knots_19_matrix %*% state_coefs_20.0
rating19$flow_kfgam20.1 <- stage_knots_19_matrix %*% state_coefs_20.1
rating19$flow_kfgam21.0 <- stage_knots_19_matrix %*% state_coefs_21.0
rating19$flow_kfgam22.0 <- stage_knots_19_matrix %*% state_coefs_22.0


# Retrieve daily flows for rating period 19

rating19[, c('stage', 'rated_flow_19', 'flow_kfgam19.0', 'flow_kfgam20.0', 'flow_kfgam20.1',
                    'flow_kfgam21.0', 'flow_kfgam22.0')] %>% 
  rename( flow_rating19.0 = rated_flow_19) %>% 
  gather( key = 'Estimator', value = 'flow_cfs', flow_rating19.0, flow_kfgam19.0, flow_kfgam20.0, 
          flow_kfgam20.1, flow_kfgam21.0, flow_kfgam22.0, -stage) %>% 
  mutate( Estimator = factor(Estimator, levels = c('flow_rating19.0', 'flow_kfgam19.0', 'flow_kfgam20.0',
                                                  'flow_kfgam20.1', 'flow_kfgam21.0', 'flow_kfgam22.0'))) %>% 
  ggplot( aes(x = flow_cfs, y = stage, color =  Estimator )) +
  geom_line() +
  theme_bw() +
  theme( legend.position = 'right') +
  xlab( 'Streamflow, in cubic feet per second' ) +
  ylab( 'Stage, in feet above gage datum' ) + 
  labs(title = 'Stage-Flow Ratings at 04122500 Pere Marquette River at Scottville, MI') 





daily_means <- readNWISdv( '04122500', parameterCd = '00060', startDate = '1994-10-25',
                               endDate = '2011-03-01')

daily_means <- renameNWISColumns( daily_means )


df_stage_flow_day_19$flow_daily <- daily_means$Flow


rating19[, c('stage', 'rated_flow_19', 'flow_kfgam19.0', 'flow_kfgam20.0', 'flow_kfgam20.1',
                    'flow_kfgam21.0', 'flow_kfgam22.0')] %>% 
  rename( flow_rating19.0 = rated_flow_19) %>% 
  gather( key = 'Estimator', value = 'flow_cfs', flow_rating19.0, flow_kfgam19.0, flow_kfgam20.0, 
          flow_kfgam20.1, flow_kfgam21.0, flow_kfgam22.0, -stage) %>% 
  mutate( Estimator = factor(Estimator, levels = c( 'flow_rating19.0', 'flow_kfgam19.0', 
          'flow_kfgam20.0', 'flow_kfgam20.1', 'flow_kfgam21.0', 'flow_kfgam22.0' ) )) %>% 
  ggplot( aes( x = day, y = flow_daily  )) +
  geom_line( color = 'blue', alpha = 0.5 ) +
  geom_line( aes( y = ytT ), color = 'red', alpha = 0.5 ) +
  theme_bw() + 
  ylab( 'Stage, in feet above streamgage datum') +
  xlab( 'Streamflow, in cubic feet per second' )
  





```




```{r plot_kf}

Zt_rate19 <- t(predict( gam_rate_19, type = 'lpmatrix'))
         q_gam_rate19 <- kemfit$states[,    1] %*% Zt_rate19
rating19$q_gam_rate19 <- as.vector(q_gam_rate19)






rating19$q_kf_rate_0001 <- as.vector( kemfit$states[, 0001] %*% Zt_rate19 )
rating19$q_kf_rate_5967 <- as.vector( kemfit$states[, 5967] %*% Zt_rate19 )

tmp <- rating19 %>% 
  left_join( rating20, by = 'stage')

tmp %>% 
  ggplot( aes( x = rated_flow_19, y = stage)) +
  geom_line( color = 'red' ) +
  geom_line( aes( x = rated_flow_20 ), color = 'blue') +
  geom_line( aes( x = q_kf_rate_0001), color = 'red',  linetype = 'dashed') +
  geom_line( aes( x = q_kf_rate_5967), color = 'blue', linetype = 'dashed')


```








```{r }

# Allocate matrix with NAs to contain Xp components for measured log_stage values to days of measurement
Xp_aug <- matrix(NA, nrow = length(seq(ndx_min:ndx_max)), ncol = ncol(Xp))
# Replace NA indicators with measured log_stage values 
Xp_aug[ndx_log_stage,] <- Xp

# Identify measured flows during daily flow period in analysis

# Find indices of measured flows in daily analysis range
ndx_meas_flow <- which( meas_data$measurement_dt >= daily_data$measurement_dt[ndx_min] &
                        meas_data$measurement_dt <= daily_data$measurement_dt[ndx_max])

Yt <- rep(NA, length(seq(27:996)))
Yt[ndx_log_stage] <- meas_data$log_flow[ndx_meas_flow] 

R = matrix( list( "R_sigma2"), 1, 1)



# State-Space Model Component Z
Zt = t(Xp_aug)

dim(Zt) <- c(1, ncol(Xp), length(Yt))
x0 = matrix(GAM_coef_est, nrow = ncol(Xp), ncol = 1)

model.dat = Yt
model.gen = list(Z = Zt, x0 = x0, R = R )

m1 <- MARSS(model.dat, model.gen)


```


```{r }



```


```{r flow_meas_error}
# ---------------------------------
# Code  Description
# ---------------------------------
# E     Excellent    The data is within 2% (percent) of the actual flow
# G     Good         The data is within 5% (percent) of the actual flow
# F     Fair         The data is within 8% (percent) of the actual flow
# P     Poor         The data are >8% (percent) of the actual flow
# https://help.waterdata.usgs.gov/codes-and-parameters/discharge-measurement-quality-code

# qualifier <- .15; flow = 10000; stderr <- (log10(flow * (1 + qualifier)) - log10(flow))/2

meas_data <- meas_data %>% 
  mutate(qualifier_err = ifelse( qualifier == 'Excellent', 0.02,
                                 ifelse( qualifier == 'Good', 0.05, 
                                         ifelse( qualifier == 'Fair', 0.08,
                                                 ifelse( qualifier == 'Poor', 0.15, NA)))),
         meas_lflow_stderr = (log10( flow * (1 + qualifier_err)) - log_flow ) / 2.)


meas_data %>% 
  ggplot( aes( x = log_flow, y = meas_lflow_stderr, color = qualifier)) +
  geom_point()


```


Stage values are typically recorded to the nearest 100th of a foot. For streamgages that included many stage-flow measurements, this rounding may creates non-uniquesness that will result in averaging of flows at the same stage.  Given that different equal-stage measurments may have different timings and uncertainties, a small, zero-mean random value is added to each stage value to provide uniqueness.  


```{r make_stage_unique}
# Create version of log_stage with small uniform random number for uniqueness
#   Uniqueness ensures that each measurement is used even with stage values
#   rounded to the 100th of a foot.  The uniqueness helps in the application to
#   a matrix multiply in the Bayesian estimation.
#  Random seed is set to facilitate reproducibility


print(paste('There are',nrow(meas_data),'stage values, but only', 
            length(unique(meas_data$log_stage)), 'of these values are unique.'))

# Setting a random seed for reproducibility
set.seed(3472792)
meas_data <- meas_data %>% 
  mutate( log_stage_unique   = log_stage_round + runif(nrow(meas_data), -0.0000049, 0.0000049)) %>% 
  arrange(log_stage_unique)

```

### Summarize measurement data

Provides summary statistics for measurements grouped by log_stage_round


```{r summ_meas_data}

meas_summary <- meas_data %>% 
  group_by( log_stage_round ) %>% 
  dplyr::summarize( mean_log_flow = mean(log_flow),
                    sd_log_flow   = sd(log_flow),
                    ivar_log_flow = 1 / sd(log_flow)^2,
                    cnt_log_flow  = n())

meas_summary <- meas_summary %>% 
  mutate(stage_round = 10^log_stage_round,
  flow_mean = 10^mean_log_flow) %>%  
  dplyr::select(stage_round, everything())

table(meas_summary$cnt_log_flow)

kable(meas_summary) %>% 
  kable_styling() %>% 
  scroll_box(width = "700px", height = "500px")


```

### Plot empirical distributions of measured flow 

Plot indicates that the log10 of flows are approximately normally distributed.

```{r dist_flow}

meas_data %>% 
  ggplot( aes(log_flow) ) +
  geom_density(fill = 'blue', color = 'blue', alpha = 0.2, show.legend = TRUE) +
  stat_function(fun  = dnorm, 
                args = list( mean = mean(log10(meas_data$flow)),
                             sd   =   sd(log10(meas_data$flow))),
                  lwd = 1, color = 'purple', show.legend = TRUE) +
  theme_bw() + 
  xlab('Log_{10} of Measured Flow, in cubic feet per second') +
  ylab('Probability Density') +
  theme( legend.position = ('right')) +
  labs(title = paste0('Figure ', fig_no,'. Empirical Distribution of Measured Log_10 Flows with Normal Density Reference')) 
fig_no <- fig_no + 1

```

## Distribution of Stage


```{r dist_stage, eval = FALSE}

# Gamma distribution is questionable applicable because of arbitrary offset
# Fit gamma distribution
fit.gamma.logstage <- fitdist(meas_data$log_stage_unique - min(meas_data$log_stage_unique)+.0001,
                              distr = 'gamma', method = 'mle')

summary(fit.gamma.logstage)

meas_data %>% 
  ggplot( aes(log_stage_unique - min(log_stage_unique)),  color = 'blue' ) +
  geom_density(fill = 'blue', alpha = 0.2, show.legend = TRUE) +
  stat_function(fun   = dgamma, 
                args  = list(shape = fit.gamma.logstage$estimate[[1]],
                             rate  = fit.gamma.logstage$estimate[[2]])) +
                #              sd   =   sd(log10(measData$stage))),
                #  lwd = 1, color = 'purple', show.legend = TRUE) +
  theme_bw() + 
  xlab('Log10 of Measured Stage minus point of zero flow offset, in feet') +
  ylab('Probability Density') +
  labs(title = paste0('Figure ', fig_no, ' Empirical Distribution of log10 Measured Stages with Gamma Density Reference'),
       subtitle = 'All stages reduced to reflect point of zero flow.')

fig_no <- fig_no + 1

```



## Plot Rating and Measurements


```{r plot_rating_group_quality, fig.height = 7}

rating_data %>% 
  ggplot( aes(y = flow, x = stage)) +
  geom_line( color = 'blue') + 
  geom_point(data = meas_data, aes(x = stage_round, y = flow, 
                                  color = quality)) +
  theme_bw() +
  theme(legend.position = "bottom") +
 
  scale_y_continuous(name   = 'Flow, in cubic feet per second',
                     trans  = "log10",
                     breaks = c(1, 10, 50, 100, 200, 500, 1000, 2000, 4000, 6000)) +
  scale_x_continuous(name   = "Stage, in feet",
                     trans  = "log10",
                     breaks = c(seq(3, 12, by = 1))) +
  ggtitle(label = paste('Figure 5. Stage-flow relation', rating_no, 'at streamgage', site_no, station_nm))

```
## Plot Rating and Measurements with Control Characteristics

```{r plot_meas_control, fig.height = 7}

meas_data %>% 
  ggplot( aes(y = flow, x = stage_round)) +
  geom_point(aes(x = stage_round, y = flow, 
                 color = Control_Degree), size = 2, alpha = 0.5) +
    scale_colour_brewer() +
  
  geom_line(data = rating_data, aes(x = stage, y = flow), color = 'darkgreen', width = 2) + 
  
  # geom_vline(data = ratingKnots, aes(xintercept = stage_knots), 
  #            color = 'salmon', linetype = 'dashed' ) +
  
  theme_bw() +
  theme(legend.position = "bottom") +
 
  scale_y_continuous(name   = 'Flow, in cubic feet per second',
                     trans  = "log10",
                     breaks = c(1, 10, 50, 100, 200, 500, 1000, 2000, 4000, 6000)) +
  scale_x_continuous(name   = "Stage, in feet",
                     trans  = "log10",
                     breaks = c(seq(3, 12, by = 1))) +
  ggtitle(label = paste('Figure 1. Stage-flow relation at streamgage', site_no, station_nm,
                        'showing effects of Controls'))
```


## Frequency of measurement qualities

The quality of measurements may need to be recomputed based on the rating in place at the time of measurement.

```{r table_quality, eval = FALSE}

meas_data$flow_exp <- approx( rating_data$stage, rating_data$flow, xout = meas_data$stage, rule = 2 )$y

meas_data$flow_err <- (meas_data$flow - meas_data$flow_exp) / (meas_data$flow_exp) * 100.

# Frequency of measurement qualities
print(table(meas_data$Quality))


meas_data %>% 
  ggplot( aes( Quality, flow_err) ) +
  geom_boxplot() +
  xlab('Rated Quality of Measurement') +
  ylab('Percent error in flow measurement') +
  geom_hline( aes(yintercept = 0) , color = 'red', linetype = 'dashed' )


```


### GAM for Stage-Flow Rating


```{r gam_estimate_of_rating, fig.height = 7 }

# Put meas_data in date order

meas_data <- meas_data %>% 
  # arrange( meas_data$measurement_dt)
  arrange(meas_data$log_stage_round)

log_stage <- meas_data$log_stage_round 

# Compute GAM 
Gmodel  <- gam(meas_data$log_flow ~ s(log_stage))

plot(Gmodel)

print( summary(Gmodel) )

# Extract basis matrix with rows equal to the number of unique log_stage values and columns equal to the dim of the basis)
Xp      <- predict( Gmodel, type = 'lpmatrix')


GAM_coef_est <- coefficients(Gmodel)

# State-space model component
#  The initial value of the state vector is the coefficients of gam model
Xo <- as.numeric(GAM_coef_est)

GAM_coef_sd  <- sqrt(diag(Gmodel$Vp))
# Xp_df   <- data.frame(Xp)

# State-space model component
Q <- Gmodel$Vp


 as.data.frame(Xp) %>%
  gather( key = 'basis', value = 'smo_stage' ) %>% 
  mutate( basis_row = rep(1:776, 10),
          log_stage_unique = rep(meas_data$log_stage_unique, 10)) %>% 
   ggplot( aes( x = log_stage_unique, y = smo_stage, color = basis)) +
   geom_line() +
   theme_bw()  +
   theme( legend.position = 'bottom') +
   xlab('Log10 of Measured Stage, in feet') +
   ylab('Basis Coefficient') + 
   scale_x_continuous( limits = c(0.5, 1.1), breaks = seq(0.5, 1.1, by = 0.1)) + 
   labs( title = 'Figure 7. Relation between log(stage) and GAM Smoothing Spline' )
  

```

```{r daily_data}





# Create data.frame of daily values
daily_data <- data.frame( measurement_dt = seq( as.Date('1970-10-01'), as.Date('2019-09-30'), 'days'))


daily_data <- left_join( daily_data, meas_data[, c('measurement_dt', 'flow', 'log_stage_round', 
                                                   'log_flow', 'meas_lflow_stderr', 'qualifier',
                                                   'measurement_tm', 'tz_cd_reported', )] )


```
## Retrieve daily stage and flow values 

```{r get_daily_values}

daily_values_pub <- readNWISdv(site_no, c('00065', '00060'))

startDate <- "1970-01-01"


# daily_values_pub <- readNWISuv(site_no, c('00065', '00060'), startDate )

unit_values_pub  <- renameNWISColumns(daily_values_pub)


```




```{r }



```



## Inital state space model
Xt is the 10x 1 state vector of GAM parameters
Q  is the fixed 10X10 covariance matrix of paramaters
R  is an estimated constant measurement variance
Zt is the matrix of fixed measurement components (smooth spline terms)
Yt is sequential daily values of log_stage_round 

```{r marss_m1}

# Min and max indices of daily_values defines range of daily values for analysis
ndx_min  <-  27
ndx_max  <- 996

# Find indices of non-zero daily values of log_stage_round in analysis range
ndx_daily_stage <- which(!is.na(daily_data$log_stage_round[ndx_min:ndx_max]))

# Predict the intercept and slope components for the spline smooth 
Xp      <- predict(Gmodel, newdata = data.frame(log_stage = daily_data$log_stage_round[ndx_min:ndx_max][ndx_daily_stage]),
                   type = 'lpmatrix')

# Allocate matrix with NAs to contain Xp components for measured log_stage values to days of measurement
Xp_aug <- matrix(NA, nrow = length(seq(ndx_min:ndx_max)), ncol = ncol(Xp))
# Replace NA indicators with measured log_stage values 
Xp_aug[ndx_log_stage,] <- Xp

# Identify measured flows during daily flow period in analysis

# Find indices of measured flows in daily analysis range
ndx_meas_flow <- which( meas_data$measurement_dt >= daily_data$measurement_dt[ndx_min] &
                        meas_data$measurement_dt <= daily_data$measurement_dt[ndx_max])

Yt <- rep(NA, length(seq(27:996)))
Yt[ndx_log_stage] <- meas_data$log_flow[ndx_meas_flow] 

R = matrix( list( "R_sigma2"), 1, 1)



# State-Space Model Component Z
Zt = t(Xp_aug)

dim(Zt) <- c(1, ncol(Xp), length(Yt))
x0 = matrix(GAM_coef_est, nrow = ncol(Xp), ncol = 1)

model.dat = Yt
model.gen = list(Z = Zt, x0 = x0, R = R )

m1 <- MARSS(model.dat, model.gen)


```


### Random Component to log stage values modelled as gamma density



```{r Bayes_parameter_estimates_dgamma, fig.height = 6}

# Use quadratic approximation posterior distribution (quap) 
baye_gam_dgamma <- quap(
  alist(
    log_flow  ~ dnorm( mu , sigma ) ,
    mu       <- Xp %*% w ,
    w         ~ dnorm(0,1),
    sigma     ~ dgamma(1)
),
data=list( log_flow = meas_data$log_flow , Xp = Xp ) ,
start=list( w = rep( 0 , ncol(Xp) ) ) )


# Parameter estimates from Bayesian estimation
precis( baye_gam_dgamma, depth = 2, prob = 0.95, digits = 5)

# Extract parameter estimates
post_dgamma   <- extract.samples( baye_gam_dgamma, n = 1e4 )
# head(post_dnorm)

gammaPars <- fitdistr(post_dgamma$sigma,"gamma")
shapeML   <- gammaPars$estimate["shape"]
rateML    <- gammaPars$estimate["rate"]

xval <- seq(0,max(post_dgamma$sigma),len=length(post_dgamma$sigma))

data.frame(post_dgamma) %>% 
  ggplot( aes(sigma ) ) +
  geom_density( stat = 'density') +
  stat_function(aes(x = xval, y = ..y..), fun = dgamma, colour="brown", 
                n = length(post_dgamma$sigma), args = list(shape=shapeML, rate=rateML)) +
  scale_x_continuous( limits = c(0.055, 0.070)) + 
  labs( title = paste0('Figure 7. Distribution of Bayes dgamma parameter sigma with reference dgamma(shape = ',
                       format(shapeML, digits = 4),', rate = ', format(rateML, digits = 4), ')')) +
  theme_bw()


```


### Random Component to log stage values modelled as normal density


```{r Bayes_parameter_estimates_dnorm, fig.height = 6}

# Use quadratic approximation posterior distribution (quap) 
baye_gam_dnorm <- quap(
  alist(
    log_flow  ~ dnorm( mu , sigma ) ,
    mu       <- Xp %*% w ,
    w         ~ dnorm(0,1),
    sigma     ~ dnorm(1,1)
),
data=list( log_flow = meas_data$log_flow , Xp = Xp ) ,
start=list( w = rep( 0 , ncol(Xp) ) ) )


# Parameter estimates from Bayesian estimation
precis( baye_gam, depth = 2, prob = 0.95, digits = 5)

# Extract parameter estimates
post_dnorm   <- extract.samples( baye_gam, n = 1e4 )
head(post_dnorm)

# dens( post$sigma, norm.comp=TRUE , main="Distribution of sigma estimates" , show.HPDI=TRUE)


data.frame(post_dnorm) %>% 
  ggplot( aes(sigma )) +
  geom_density( stat = 'density', alpha = 0.25, color = 'blue', size = 1.5 ) +
  stat_function(fun = dnorm, 
                args = list( mean = mean(post_dnorm$sigma), 
                             sd   = sd(  post_dnorm$sigma)),   color = 'red',  size = 1.5, ) +
  # scale_color_manual(name = "Density Plot", values = c("Empirical" = 'blue' ) ) +
  scale_x_continuous( limits = c(0.0575, 0.070)) + 
  labs( title = paste0('Figure 6. Distribution of sigma for  Bayes dnorm parameter sigma with reference dnorm(mean = ',
                       format(mean(post_dnorm$sigma), digits = 4),', sd = ', 
                       format(  sd(post_dnorm$sigma), digits = 4), ')')) +
  theme_bw()
  


```

### Include qualifier info to account for varying uncertainty


```{r Bayes_parameter_estimates_uncert_x, fig.height = 6}
x0 = Xp[1:776,1]; x1 = Xp[1:776,2]; x2 = Xp[1:776,3]; x3 = Xp[1:776,4]; x4 = Xp[1:776, 5];
x5 = Xp[1:776,6]; x6 = Xp[1:776,7]; x7 = Xp[1:776,8]; x8 = Xp[1:776,9]; x9 = Xp[1:776,10];


# Use quadratic approximation posterior distribution (quap) 
baye_gam_cert_x_dnorm1 <- ulam(
  alist(
    lflow_meas   ~ dnorm( lflow_true , lflow_meas_sd ) ,
    vector[N]:lflow_true ~ dnorm(lflow_rate, lflow_true_sd),
    lflow_rate   <- b0 * x0 + b1 * x1 + b3 * x3 + b4 * x4 + b5 * x5 + b6 * x6 + b7 * x7 + b7 * x7 + b8 * x8 + b9 * x9,
    b0            ~ dnorm(  2.12203099, 0.002302916 ),
    b1            ~ dnorm( -0.31305639, 0.011874773 ),
    b2            ~ dnorm( -0.87836073, 0.085117576 ),
    b3            ~ dnorm( -0.23794731, 0.017755688 ),
    b4            ~ dnorm(  0.53667040, 0.069286763 ),
    b5            ~ dnorm(  0.08390154, 0.009977110 ),
    b6            ~ dnorm( -0.42387766, 0.065635509 ),
    b7            ~ dnorm( -0.25612051, 0.030899320 ),
    b8            ~ dnorm(  1.00733829, 0.200192569 ),
    b9            ~ dnorm(  0.97733403, 0.039447235 ),
    lflow_true_sd ~ dnorm(  0 , 1)
),
data=list( lflow_meas    = meas_data$log_flow, 
           x0 = x0, x1  = x1, x2 = x2, x3 = x3, x4 = x4, x5 = x5, x6 = x6, x7 = x7, x8 = x8, x9 = x9, 
           lflow_meas_sd = meas_data$meas_lflow_stderr, lflow_meas = meas_data$log_flow , N = nrow(meas_data)),
start=list(mu0 = 2.12203099,  mu1 = -0.31305639, mu2 = -0.87836073, mu3 = -0.23794731, mu4 =  0.53667040,
           mu5 = 0.08390154,  mu6 = -0.42387766, mu7 = -0.25612051, mu8 =  1.00733829, mu9 =  0.97733403,
           sd0 = 0.002302916, sd1 = 0.011874773, sd2 = 0.085117576, sd3 = 0.017755688, sd4 = 0.069286763,
           sd5 = 0.009977110, sd6 = 0.065635509, sd7 = 0.030899320, sd8 = 0.200192569, sd9 = 0.039447235),
iter = 5000, warmup = 1000, chains = 3, cores = 6)

# control= list(max_treedepth = 20, adapt_delta = 0.95))


# Parameter estimates from Bayesian estimation
precis( baye_gam_cert_x_dnorm, depth = 1, prob = 0.95, digits = 5)

# Extract parameter estimates
post_dnorm   <- extract.samples( baye_gam_cert_x_dnorm, n = 1e4 )
# head(post_dnorm)

# dens( post$sigma, norm.comp=TRUE , main="Distribution of sigma estimates" , show.HPDI=TRUE)


data.frame(post_dnorm) %>% 
  ggplot( aes(lflow_true_sd )) +
  geom_density( stat = 'density', alpha = 0.25, color = 'blue', size = 1.5 ) +
  stat_function(fun = dnorm, 
                args = list( mean = mean(post_dnorm$lflow_true_sd), 
                             sd   = sd(  post_dnorm$lflow_true_sd)),   color = 'red',  size = 1.5, ) +
  # scale_color_manual(name = "Density Plot", values = c("Empirical" = 'blue' ) ) +
  # scale_x_continuous( limits = c(0.0575, 0.070)) + 
  labs( title = paste0('Figure 6. Distribution of sigma for  Bayes dnorm parameter sigma with reference dnorm(mean = ',
                       format(mean(post_dnorm$lflow_true_sd), digits = 4),', sd = ', 
                       format(  sd(post_dnorm$lflow_true_sd), digits = 4), ')')) +
  theme_bw()
  
tmp <- precis( baye_gam_cert_x_dnorm, depth = 2, prob = 0.95, digits = 5)


resid <- tmp$mean[1:776] - meas_data$log_flow
plot( meas_data$log_flow, resid )
abline( a = 0, b = 0, col = 'red')

ggdistribution(dgamma, x = seq(0, 2, 0.02), shape = 2, scale = 0.2)
ggdistribution(dchisq, x = seq(0, 10, 0.02), df = 2.25)
```

## Re-estimate with better staring values


```{r }

dlist <- list(
  Q_obs = standardize( meas_data$log_flow ),
  Q_sd =  meas_data$ d$Divorce.SE / sd( d$Divorce ),
  S_obs = standardize( d$Marriage ),
  S_sd = d$Marriage.SE / sd( d$Marriage ),
  A = standardize( d$MedianAgeMarriage ),
  N = nrow(d)
)


baye_gam_uncert_x <- ulam(
  alist(
    lflow_mea  ~ dnorm( lflow_true , lflow_sd ) ,
    vector[N]:lflow_true ~ dnorm( lflow_rat, rat_sd),
    lflow_rat <- b0 * x0 + b1 * x1 + b3 * x3 + b4 * x4 + b5 * x5 + b6 * x6 + b7 * x7 + b7 * x7 + b8 * x8 + b9 * x9,
    lstage_mea ~
    vector[N]:lstage_true ~ 
      
    b0         ~ dnorm(mu0,sd0),
    b1         ~ dnorm(mu1,sd1),
    b2         ~ dnorm(mu2,sd2),
    b3         ~ dnorm(mu3,sd3),
    b4         ~ dnorm(mu4,sd4),
    b5         ~ dnorm(mu5,sd5),
    b6         ~ dnorm(mu6,sd6),
    b7         ~ dnorm(mu7,sd7),
    b8         ~ dnorm(mu8,sd8),
    b9         ~ dnorm(mu9,sd9),
    lflow_mea  ~ dnorm(lflow_est, lflow_sig),
    sigma      ~ dexp(1)
),
data=list( log_flow  = meas_data$log_flow, 
           x0 = x0, x1 = x1, x2 = x2, x3 = x3, x4 = x4, x5 = x5, x6 = x6, x7 = x7, x8 = x8, x9 = x9, 
           lflow_sig = meas_data$meas_lflow_stderr, N = length(log_flow)),
start=list(lflow_est = meas_data$log_flow ,
           mu0 = 2.12203099,  mu1 = -0.31305639, mu2 = -0.87836073, mu3 = -0.23794731, mu4 =  0.53667040,
           mu5 = 0.08390154,  mu6 = -0.42387766, mu7 = -0.25612051, mu8 =  1.00733829, mu9 =  0.97733403,
           sd0 = 0.002302916, sd1 = 0.011874773, sd2 = 0.085117576, sd3 = 0.017755688, sd4 = 0.069286763,
           sd5 = 0.009977110, sd6 = 0.065635509, sd7 = 0.030899320, sd8 = 0.200192569, sd9 = 0.039447235),
iter = 2000, warmup = 1000, chains = 3, cores = 6,
control=list(max_treedepth = 20, adapt_delta = 0.95))

data(gmcv)

dlist <- list(
  D_obs = standardize( d$Divorce ),
  D_sd = d$Divorce.SE / sd( d$Divorce ),
  M_obs = standardize( d$Marriage ),
  M_sd = d$Marriage.SE / sd( d$Marriage ),
  A = standardize( d$MedianAgeMarriage ),
  N = nrow(d)
)
m15.2 <- ulam(
  alist(
    D_obs ~ dnorm( D_est , D_sd ),
    vector[N]:D_est ~ dnorm( mu , sigma ),
    mu <- a + bA*A + bM*M_est[i],
    M_obs ~ dnorm( M_est , M_sd ),
    vector[N]:M_est ~ dnorm( 0 , 1 ),
    a ~ dnorm(0,0.2),
    bA ~ dnorm(0,0.5),
    bM ~ dnorm(0,0.5),
    sigma ~ dexp( 1 )
  ) , data=dlist , chains=4 , cores=4 )

```



```{r }

# Correlation among parameters
ggcorrplot( cov2cor( vcov( baye_gam_dnorm)), type = 'lower', 
                     insig = 'blank',  lab = TRUE, digits = 2)

```


# Compare Bayesian and GAM model parameter estimates


```{r compare_gam_bayes_param}
coef_gam <- data.frame( gam_coef = coef(Gmodel), baye_coef = coef(baye_gam)[1:10], 
                        id     = c('W1', 'W2', 'W3', 'W4', 'W5', 'W6', 'W7', 'W8', 'W9', 'W10'))

coef_gam %>% 
  ggplot( aes( x = gam_coef, y = baye_coef, label = id)) +
  geom_point( color = 'blue', size = 2) +
  geom_abline( slope = 1, intercept = 0, color = 'red') +
  geom_label_repel(aes(label = id), size = 2,
                     box.padding = 0.25, 
                   point.padding = 0.25,
                   label.padding = 0.25,
                  segment.color = 'grey50') +
  theme_classic() +
  labs( x = 'GAM Parameter Estimate',  y = 'Bayesian Parameter Estimate',
        title = 'Figure 4. Compare parameters estimated with and without random component in log stage values')
  
```
```{r extract_bayes}
# Extract parameter estimates
post   <- extract.samples( baye_gam, n = 1e4 )
head(post, n = 5)

dens( post$sigma, norm.comp=TRUE , main="Distribution of sigma estimates" , show.HPDI=TRUE)


data.frame(post) %>% 
  ggplot( aes(sigma ) ) +
  geom_density( stat = 'density') +
  stat_function(fun = dnorm, 
                args = list( mean =  mean(post$sigma), sd = sd(post$sigma)), color = 'red' )

```


```{r }
# Use quadratic approximation posterior distribution (quap) 
baye_gam_dexp <- quap(
  alist(
    log_flow  ~ dnorm( mu , sigma ) ,
    mu       <- Xp %*% w ,
    w         ~ dnorm(0,1),
    sigma     ~ dexp(1)
),
data=list( log_flow = meas_data$log_flow , Xp = Xp ) ,
start=list( w = rep( 0 , ncol(Xp) ) ) )


# Parameter estimates from Bayesian estimation
precis( baye_gam, depth = 2, prob = 0.95, digits = 5)



# Link uses the quap approximation, and samples from the posterior distribution, and computes 
#   µ for each case in the data and sample from the posterior distribution.
mu     <- link( baye_gam )
str(mu)

# Generage a sequence of log_stages for analysis

log_stage_seq <- seq( from = min(log_stage), to = max(log_stage), length.out = 50 )

data_seq      <- data.frame( log_stage = log_stage_seq )

Xp_seq        <- predict( Gmodel, newdata = data_seq, type = 'lpmatrix')

mu_seq        <- link( baye_gam, data = list(Xp = Xp_seq))[,1:50]

tmp025 <- as.data.frame(mu_seq) %>% 
  dplyr::summarize_all(list(q025 = quantile), probs = 0.025)
tmp975 <- as.data.frame(mu_seq) %>% 
  dplyr::summarize_all(list(q025 = quantile), probs = 0.975)
tmpave <- as.data.frame(mu_seq) %>% 
  dplyr::summarize_all(list(mean = mean))



baye_rating <- data.frame(t(rbind(log_stage_seq,as.matrix(tmp025), as.matrix(tmpave), as.matrix(tmp975))))

colnames(baye_rating) <- c('Stage','q025', 'mean', 'q975')
rownames(baye_rating) <- 1:50

baye_rating %>% 
  ggplot( aes( x = Stage, y = mean)) +
  geom_line(   )


mu_seq  <- link( baye_gam, data = data.frame( Xp = Xp_new ) )


# Compare bayesian estimated parameters with gam model estimates





```


## Evaluate spline model

```{r eval_bs, eval = FALSE}

post    <- extract.samples( n_bs_reg_02 )

posta   <- mean(post$a)
posts   <- mean(post$s)
postw   <- apply( post$w, 2, mean)

mu   <- link( n_bs_reg_02 )

mu_PI <- apply(mu, 2, PI, 0.95)

plot( measData$log_stage, measData$log_flow, col = col.alpha(rangi2, 0.5))

shade( mu_PI, measData$log_stage, col = col.alpha("black", 0.5))

stage.seq <- seq(from = min(measData$log_stage), to = max(measData$log_stage), length = 100)

stage.mu  <- link( n_bs_reg_02, data = data.frame(log_stage = stage.seq))

# Find minimum difference between entry in 
meas_data_ndx <- rep(NA, length(stage.seq))
for (i in 1:length(stage.seq)){
  adiff <- abs(stage.seq[i] - log_stage)
  meas_data_ndx[i] <- which.min(adiff)
  print( paste(i, ndx, adiff[ndx]))
}

# summarize the distribution of stage.mu
mu.mean <- apply( stage.mu[,meas_data_ndx] , 2 , mean )
mu.PI   <- apply( stage.mu[,meas_data_ndx] , 2 , PI , prob=0.95 )

smooth.spline(x = log_stage, y = log_flow)
# , w = NULL, df = measData, lambda = NULL, cv = FALSE,
#               all.knots = FALSE, nknots = .nknots.smspl,
#               keep.data = TRUE, df.offset = 0, penalty = 1,
#               control.spar = list(), tol = 1e-6 * IQR(log_stage), keep.stuff = FALSE)

require(graphics)
plot(dist ~ speed, data = cars, main = "data(cars)  &  smoothing splines")
cars.spl <- with(cars, smooth.spline(speed, dist))
cars.spl
## This example has duplicate points, so avoid cv = TRUE

lines(cars.spl, col = "blue", lwd = 2)
ss10 <- smooth.spline(cars[,"speed"], cars[,"dist"], df = 10)
lines(ss10, lty = 2, col = "red")
legend(5,120,c(paste("default [C.V.] => df =",round(cars.spl$df,1)),
               "s( * , df = 10)"), col = c("blue","red"), lty = 1:2,
       bg = 'bisque')



plot(log_flow ~ log_stage, data = measData, main = "data(cars)  &  smoothing splines",
     col = 'grey')
rate1.spl <- with(measData, smooth.spline(log_stage, log_flow))
rate1.spl
## This example has duplicate points, so avoid cv = TRUE

lines(rate1.spl, col = "blue", lwd = 2)
rate2.spl <- smooth.spline(measData[,"log_stage"], measData[,"log_flow"], df = 8)
lines(rate2.spl, lty = 2, col = "red")

legend(5,120,c(paste("default [C.V.] => df =",round(cars.spl$df,1)),
               "s( * , df = 10)"), col = c("blue","red"), lty = 1:2,
       bg = 'bisque')


# plot raw data
# fading out points to make line and interval more visible
plot( log_flow ~ log_stage , data=measData , col=col.alpha(rangi2,0.5) )
# plot the MAP line, aka the mean mu for each weight
lines( stage.seq , mu.mean )
# plot a shaded region for 89% PI
shade( mu.PI , stage.seq )


# pred      <- link( m_bs_reg_01, data = measData  )
# 
# 
# post <- extract.samples(m_bs_reg_01)
# w <- apply( post$w , 2 , mean )
# plot( NULL , xlim=range(tmp$log_stage) , ylim=c(-5,5) ,
# xlab="year" , ylab="basis * weight" )
# for ( i in 1:ncol(Bspline) ) lines( tmp$log_stage , w[i]*Bspline[,i] )
# 
# mu <- link( m_bs_reg_01 )
# mu_PI <- apply(mu,2,PI,0.95)
# plot( tmp$log_stage , tmp$log_flow , col=col.alpha(rangi2,0.3) , pch=16 )
# shade( mu_PI , tmp$log_stage , col=col.alpha("black",0.5) )


```


```{r gam_rate, eval = FALSE}

measData <- measData %>% 
  mutate( log_flow = log10(flow), log_stage = log10(stage))

log_stage <- measData$log_stage_
log_flow  <- measData$log_flow

gam_spline <- gam(log_flow ~ s(log_stage), data = measData, method = "ML")
measData$gam_hat <- gam_spline$fitted.values


measData %>% 
  ggplot( aes( x = log_stage, y = log_flow)) +
  geom_point() +
  geom_line( aes( x = log_stage, y = gam_hat), color = 'red')


plot(gam_spline)

m_bs_reg_01 <- quap(
  alist(
    log_flow ~  dnorm( mu, sigma ),
    mu       <- a + Bspline %*% w,
    a        ~  dnorm( 2, 1.5 ),
    w        ~  dnorm( 0, 1   ),
    sigma    ~  dexp(1)
  ),
  data = list( log_flow = log_flow, log_stage = log_stage, B = Bspline),
  start = list(w = rep( 0, ncol(Bspline)))
)

precis( m_bs_reg_01, depth = 2, prob = 0.95, digits = 5)




```



```{r annimate_rating}

library(gganimate)
#> Loading required package: ggplot2

# We'll start with a static plot
p <- ggplot(iris, aes(x = Petal.Width, y = Petal.Length)) + 
  geom_line() +
  transition_states(Species,
                    transition_length = 2,
                    state_length = 1)

p


animate(p, renderer = file_renderer('~/animation/'))[1:6]
animate(p, 'output.gif')

p <- rating_long %>% 
  ggplot( aes(x = rated_flow, y = stage, color =  rating_no )) +
  geom_point() +
  transition_states(rating_no,
                    transition_length = 1,
                    state_length = 1) +
  theme_bw() +
  theme( legend.position = 'right') +
  xlab( 'Streamflow, in cubic feet per second' ) +
  ylab( 'Stage, in feet above gage datum' ) + 
  labs(title = 'Stage-Flow Ratings at 04122500 Pere Marquette River at Scottville, MI') +
  geom_line( data = rating18.0, aes( x = rated_flow_18.0, y = stage), color = 'black', linetype = 'dots' )

animate(p, nframes = 6, fps = 2)



tmp <- as.data.frame(  as.table( flow_xtT ))
colnames(tmp) <- c('Date', 'stage', 'flow')
tmp$stage <- rating19.0$stage
tmp$Date  <- rep( df_stage_flow_day_19$day[ndx_kfss] , each = 731)

rating_days_measured <- tmp %>% 
  ggplot( aes(x = flow, y = stage )) +
  geom_point( color = 'blue', size = 0.5 ) +
    transition_time(Date) +
  labs(title = 'Date of Measurement: {frame_time}' ) + 
  theme_bw() +
  theme( legend.position = 'right') +
  xlab( 'Streamflow, in cubic feet per second' ) +
  ylab( 'Stage, in feet above gage datum' ) + 
  # labs(title = 'Stage-Flow Ratings at 04122500 Pere Marquette River at Scottville, MI') +
  geom_line( data = rating18.0, aes( x = rated_flow_18.0, y = stage), color = 'black', linetype = 'dots' )

animate(rating_days_measured, nframes = 155, fps = 10)
```