---
title: "Dynamic Stage-Discharge Rating estimation using spline Model  Kalman Filter"
author: "David J Holtschlag"
date: "`r format(Sys.Date(), '%A %b %d, %Y') `"
output: 
  pdf_document:
bibliography: bibliography.bib
---

## Abstract
A Bayesian approach is developed to estimate parameters of a generalized additive model (GAM) [@Wood2006] to describe the dynamic relation between stage and discharge (flow) at a streamgage.  The GAM model provides a basis function and parameter estimates to describe the smooth, monotonically increasing relation between stage and flow using data on direct measurements of stage and flow commonly used to develop a so called rating curve. Bayesian estimation then uses the GAM basis function and re-estimates parameters to explicitly account for the uncertainty in the stage measurements. Data for USGS streamgage 04112500 Red Cedar River at East Lansing, MI are used to illustrate the approach and results of the generally applicable method.  Computations are implemented in the R programming environment [@cite_R].

### Setup computational environment

```{r setup, include=FALSE}
library( "rethinking" )
options(mc.cores = parallel::detectCores())
library(dataRetrieval)
library(lubridate)
library(tidyverse)
library(ggthemes)
library(splines)
library(mgcv)
library(fitdistrplus)
library(stringr)
library(kableExtra)
library(ggrepel)
library(ggcorrplot)
# strcmp and other Matlab like functions
library(pracma)
# Package expss contains the function vlookup
library(expss)
# Package to plot distribution
library(ggfortify)
# Multivariate Autoregressive State Space Model
library(MARSS)
library(matrixcalc)
# The broom package takes the messy output of built-in functions in R, such as lm, nls, or t.test, and turns them into tidy data frames.
library(broom)
library(gganimate)
library(gifski)
knitr::opts_chunk$set(echo = TRUE)
```

## Retrieve Streamgaging Site Inforamtion 

```{r specify_gage, echo = FALSE}

# initialize counter for figures
fig_no <- 1
tab_no <- 1

# Retrieve gage info for specified site Number
site_no    <- '04122500'

gage_info  <- readNWISsite(site_no)

# Standardize field names
gage_info  <- renameNWISColumns(gage_info)

# 
gage_info$station_nm <- str_to_title( gage_info$station_nm )

gage_info %>% 
  rename( drain_area_mi2 = drain_area_va,
          latitude       = dec_lat_va,
          longitude      = dec_long_va,
          altitude_ft    = alt_va ,
          altitude_datum = alt_datum_cd) %>% 
  dplyr::select( site_no, station_nm, drain_area_mi2, latitude, longitude, altitude_ft, altitude_datum) %>% 
  kable( caption = paste0('Table ',tab_no,'. Summary streamgage information' )) %>% 
  kable_styling()


tab_no <- tab_no + 1

# The starting and ending dates of the model
#   Based on the beginning of unit stage date and endin on the first flow measurement of water year 2020
date_beg_model  <- as.Date( '1994-10-25', format = '%Y-%m-%d' ) 
date_end_model  <- as.Date( '2019-10-08', format = '%Y-%m-%d' )

```
### Get rating period info and display

```{r rating_history, echo = FALSE}
rating_history <- read.csv( 'Data/04122500/rating_period_start_end.txt', sep = '\t', header = TRUE)

renderPlot({
  rating_history %>% 
  kable( caption = paste0('Table ', tab_no, '. Streamflow Rating Curve History at 04122500 Pere Marquette River at Scottvill, MI')) %>% kable_styling( full_width = FALSE )
tab_no = tab_no + 1
})
 
```


### Retrieve Stage-Flow Ratings

Reads and plots rating curves 18, 19, 20, and 21 for streamgage 04122500 Pere Marquette River at Scottville, MI and plots the curves

```{r read_rating_data, fig.width = 9, fig.height = 8, echo = FALSE}

# Rating curve at the selected site.  The 'exsa' parameter is an extended table
#   provide detailed info on stage discharge

rating18 <- read.csv('Data/04122500/04122500_rating_18.0.txt', header = FALSE, sep = '\t') 
colnames( rating18 ) <- c('rated_flow_18', 'stage')

rating19 <- read.csv('Data/04122500/04122500_rating_19.0.txt', header = FALSE, sep = '\t') 
colnames( rating19 ) <- c('rated_flow_19', 'stage')

rating20 <- read.csv('Data/04122500/04122500_rating_20.0.txt', header = FALSE, sep = '\t') 
colnames( rating20 ) <- c('rated_flow_20', 'stage')

rating21 <- read.csv('Data/04122500/04122500_rating_21.0.txt', header = FALSE, sep = '\t') 
colnames( rating21 ) <- c('rated_flow_21', 'stage')


rating_wide <- rating18   %>%  
  left_join(   rating19 ) %>% 
  left_join(   rating20 ) %>% 
  left_join(   rating21 ) %>% 
  dplyr::select(stage, everything() )
  
rating_long <- rating_wide %>% 
  gather( key = 'rating_no', value = 'rated_flow', rated_flow_18, 
          rated_flow_19, rated_flow_20, rated_flow_20, rated_flow_21)

renderPlot({
rating_long %>% 
  ggplot( aes( x = rated_flow, y = stage,  group = rating_no)) +
  geom_line( aes( color = rating_no ), size = 1.0, alpha = 0.50) +
  theme_bw() +
  theme( legend.position = 'bottom' ) +
  xlab( 'Flow, in cubic feet per second' ) +
  ylab( 'Stage, in feet above gage datum' ) +
  annotate('text', x = c(1300, 1950, 3200, 5200), y = c( 4.0, 4.7, 6.0, 7.0), 
           label = c('18', '19', '20', '21')) +
  labs( title =  paste0('Figure ', fig_no, '. Selected rating curves at 04122500 Pere Marquette River at Scottville, MI' ))
})
fig_no <- fig_no + 1
        
  
```


## Develop a generatized additive model (GAM) for rating 19

```{r gam_rating19, fig.width = 9, echo = FALSE }


# Estimate generalized additive model of stage - discharge rating 19 
gam_rate_19 <- gam( rated_flow_19 ~ s(stage), data = rating19 )

# Print summary of gam_rate_19
print(summary(gam_rate_19))

# Generalized Additive Model prediction for rating curve 18
gam_rate_19_pred     <- predict(gam_rate_19, se.fit = TRUE)

rating19$gam_flow    <- gam_rate_19_pred$fit
rating19$se_fit      <- gam_rate_19_pred$se.fit
rating19$error_025   <- rating19$rated_flow_19 - (gam_rate_19_pred$fit + qnorm(0.025) * gam_rate_19_pred$se.fit)   
rating19$error_975   <- rating19$rated_flow_19 - (gam_rate_19_pred$fit + qnorm(0.975) * gam_rate_19_pred$se.fit) 
rating19$flow_error  <- gam_rate_19$residuals
rating19$pct_error   <- gam_rate_19$residuals / gam_rate_19$fitted.values * 100

rating19 %>% 
  # dplyr::arrange( stage ) %>% 
  ggplot(    aes( y = pct_error, x = stage  )) +
  geom_point( color = 'red', size = 0.5) +
  # geom_line( aes( y = error_025), color = 'salmon') +
  # geom_line( aes( y = error_975), color = 'salmon') +
  # geom_ribbon( aes( ymin = error_025, ymax = error_975), fill = 'blue', alpha = 0.25 ) +
  scale_y_continuous( breaks = seq( from = -0.5, to = 2, by = 0.5)) +
  ylab( 'Flow Error, in percent') +
  xlab( 'Stage, in feet') +
  theme_bw() +
  geom_hline( yintercept = 0, color = 'blue', linetype = 'dashed') +
  theme( legend.position = 'bottom') +
  labs( title = paste0('Figure ', fig_no, '. Errors in Flow for GAM of rating curve 19 at ', gage_info$site_no, 
                       ' ', gage_info$station_nm))
fig_no <- fig_no + 1

```




### Retrieve Stage-Flow Measurement Pairs

```{r read_meas_stage_flow}

discrete_meas   <- readNWISmeas(site_no, expanded = TRUE, convertType = TRUE)

# Standardize field names
discrete_meas  <- renameNWISColumns(discrete_meas)


control_type <- 
  recode_factor( discrete_meas$control_type_cd, 
                 'Clear'    = 'Clear Channel',
          'VegetationLight' = 'Light Vegetation',
          'DebrisLight'     = 'Light Debris',   
          'IceShore'        = 'Shore Ice',
          'DebrisModerate'  = 'Moderate Debris',
          'IceCover'        = 'Ice Cover', .ordered = TRUE ) 

discrete_meas$control_type <- control_type


discrete_meas <- discrete_meas %>% 
  rename('stage_discrete_mea'  = 'gage_height_va',
         'flow_discrete_mea'   = 'discharge_va',
         'qualifier'           = 'measured_rating_diff') %>% 
  filter(qualifier            != 'Unspecified' | qualifier == !is.na(qualifier)) %>%
  mutate('qualifier'           =  ordered(qualifier, levels = c('Excellent', 'Good', 'Fair', 'Poor')))

# Source 
# https://help.waterdata.usgs.gov/codes-and-parameters/discharge-measurement-quality-code
# ---------------------------------
# Code  Description
# ---------------------------------
# E     Excellent    The data is within 2% (percent) of the actual flow {95 percent of the time}
# G     Good         The data is within 5% (percent) of the actual flow
# F     Fair         The data is within 8% (percent) of the actual flow
# P     Poor         The data are >8% (percent) of the actual flow

discrete_meas <- discrete_meas %>% 
  mutate( se_pct = case_when( qualifier == 'Excellent' ~  1.0,
                              qualifier == 'Good'      ~  2.5,
                              qualifier == 'Fair'      ~  4.0,
                              qualifier == 'Poor'      ~ 10.0),
          se_flow = se_pct/100 * flow_discrete_mea ) %>% 
  filter( measurement_dt >= date_beg_model & measurement_dt <= date_end_model )

# Only cases where flow and stage are measured
discrete_meas <- discrete_meas[complete.cases(discrete_meas$stage_discrete_mea, discrete_meas$flow_discrete_mea),]
```



```{r measurement_type_freq}
table( discrete_meas$control_type, discrete_meas$qualifier ) %>%
  kable( digits = 4, 
         caption = 'Table 1. Frequency of flow measurements by control type and measurement qualifiers at 04122500 Pere Marquette River at Scottville, MI, from 1985 to 2019.') %>%
  kable_styling() %>% 
  add_header_above(c("Type of Control", "Qualifier", 'Qualifier', "Qualifier", "Qualifier"))

# Remove ice cover measurements
discrete_meas <- discrete_meas %>% 
  dplyr::filter( control_type != 'Ice Cover' ) %>% 
  dplyr::select( 'measurement_nu', 'measurement_dt', 'measurement_tm', 'tz_cd_reported',
                 'stage_discrete_mea', 'flow_discrete_mea', 'qualifier',  'control_type',
                 'se_pct', 'se_flow' )


# Compute time between consequtive flow measurements
df <- data.frame( 'days_between_measurements' = diff(discrete_meas$measurement_dt))


# Plot the probability density of days between measurements
df %>% 
  ggplot( aes( x = days_between_measurements )) + 
  geom_density( fill = 'blue', alpha = 0.25 ) +
  theme_bw() +
  geom_vline( xintercept = 46, color = 'black', linetype = 'dashed' ) +
  xlab( 'Time between flow measurements, in days' ) +
  ylab( 'Probability Density') +
  labs( title = paste('Figure 2. Distribution of days between consecutive flow measurements at \n04122500 Pere River at Scottville, MI from', date_beg_model, 'to', date_end_model) ) +
  annotate('text', x = 40, y = 0.0265, label = '46')
  
  


```

## Read unit stage values for 04122500 Pere Marquette River at Scottville, MI

```{r unit_daily_stage}

stage_unit <- read.csv('Data/04122500/stage/Gage_height.ft@04122500.EntireRecorda.csv', header = TRUE, sep = ',',
                       stringsAsFactors = FALSE, comment = '#')

# Convert from character field to datetime referenced to EST
stage_unit$datetime <- as_datetime(stage_unit$Timestamp..UTC.05.00, tz = 'America/New_York')

# Summarize unit stage to daily mean stage
stage_day <- stage_unit %>% 
  rename( stage = Value ) %>% 
  group_by(day = floor_date( datetime, 'day')) %>% 
  summarize( stage_mean_day  = mean( stage, na.rm = TRUE ),
             n         = n()) %>% 
  ungroup() %>% 
  arrange( day ) %>% 
  mutate( delta_day = as.numeric(day - lag( day ), units = 'days' ),
          day_date  = as.Date( day )) 


# Measured flows and daily stages for the interval when rating 19 was active
#  Note: Discrete measurement timing is approximated to days

df_stage_flow_allDays <- data.frame(day = seq.Date(from = date_beg_model, to = date_end_model, by = '1 day'))

# calculate_mode function from https://exploratory.io/note/kanaugust/1701090969905358
#   Used in subsequent summarize
calculate_mode <- function(x) {
  uniqx <- unique(na.omit(x))
  uniqx[which.max(tabulate(match(x, uniqx)))]
}

# Take the mean flow and stage for days of multiple flow measurements
df_discrete_meas <- discrete_meas %>% 
  group_by( measurement_dt ) %>% 
  summarize( flow_discrete_mea       = mean( flow_discrete_mea ) ,
             stage_discrete_mea      = mean( stage_discrete_mea     ) ,
             n                       = n(),
             qualifier               = calculate_mode(qualifier),
             control_type            = calculate_mode(control_type),
             se_pct                  = mean(se_pct),
             se_flow                 = mean(se_flow))

# Integrate daily flow and discrete measurements 
df_stage_flow_allDays <-df_stage_flow_allDays %>% 
  left_join( df_discrete_meas[, c('measurement_dt', 'flow_discrete_mea', 'qualifier', 'control_type', 'se_flow')], 
             by = c( 'day'  = 'measurement_dt')) %>% 
  left_join( stage_day[, c('day_date', 'stage_mean_day')], by = c( 'day' = 'day_date' )) %>% 
  filter( day >= date_beg_model & day <= date_end_model )

```

### Estimate missing stage based on daily mean flow and rating curve 19

```{r missing_stage}


if( file.exists('flow_daily_pub.RData')){
  print('Loading published daily_means from RData file.')
  load('flow_daily_pub.RData')
} else {
  print('Retrieving daily flow from NWIS')
  flow_daily_pub <- readNWISdv( site_no, parameterCd = '00060', startDate = date_beg_model,
                             endDate = date_end_model)
  # Standardize names
  flow_daily_pub <- renameNWISColumns( flow_daily_pub )
  save(flow_daily_pub, file = 'flow_daily_pub.RData')
}




df_stage_flow_allDays <- df_stage_flow_allDays %>% 
  left_join( flow_daily_pub[, c('Date', 'Flow')], by = c( 'day' = 'Date' )) %>% 
  rename( 'Date' = 'day',
          'flow_daily_pub'    = 'Flow')


# find indices of missing stage_mean_day values
ndx_miss <- which( is.na(df_stage_flow_allDays$stage_mean_day ))

# Find dates for missing data
dates_missing <- df_stage_flow_allDays$Date[ndx_miss]

# Estimate daily mean stage from published daily mean flow
stage_missing <- approx( x = rating19$rated_flow_19, y = rating19$stage, 
                         xout =df_stage_flow_allDays$flow_daily_pub[ndx_miss])


df_stage_flow_allDays$stage_mean_day[ ndx_miss ] <- stage_missing$y

```


## Inital state space model
Yt is measured flow referenced to the day of measurement during the period when rating 19 was active. Most values are NA
Xt is the 10 x 1 state vector of GAM parameters used to approximate rating 19
Q  is the fixed 10X10 covariance matrix of paramaters
R  is an estimated constant measurement variance
Zt is the matrix of fixed measurement components 
Yt is sequential daily values of log_stage_round 


Note: A multiplicative form for the Q matrix can be specified as:
This is provided just as a reminder of the format for defining Q with a multiplicative constant.
Q <- matrix(list('qb', 0, 0, 0, 0, 0, 0, '10*qb'), 3, 3)

```{r marss_m1}

# Yt is the measurement vector that contains directly measured flows (not daily means)
#   stage_flow_day_19 as 1 x T matrix
Yt <- matrix(df_stage_flow_allDays$flow_discrete_mea, nrow = 1)

# Indexes for measured flows in vector
ndx_ian_Yt <- which( !is.na(Yt) )

print(paste('The measurement vector is', 
            format( length(which( is.na(Yt))) / length(Yt) * 100, digit = 3 ),'percent NA values.'))


# Zt is the time-varying design matrix in the measurement equation
#   Zt contains the intercept and knot components for the spline from the GA model 
Zt <- t(predict( gam_rate_19, newdata = data.frame( stage = df_stage_flow_allDays$stage_mean_day),
                                           type = 'lpmatrix'))
# Remove row names
rownames(Zt) <- NULL

# Populate the Zt array in the format that MARSS expects
Zt <- array(Zt, dim = c(1, 10, nrow(df_stage_flow_allDays)))

# A is an unused time-varying vector in the measurement equation
A  <- 'zero'


R  <- array(1e+06, dim = c(1, 1, nrow(df_stage_flow_allDays)))
# Measurement variance specification
#   The standard error of flow as a percentage that varies with perceived accuracy of the measured flow
R[1,1, ndx_ian_Yt] <- (df_stage_flow_allDays$se_flow[ndx_ian_Yt])^2



# Set up data frame to characterize measurement variances
df <- data.frame( measurement_variance <- R[1, 1, ndx_ian_Yt])
df %>% 
  ggplot( aes( x = measurement_variance)) +
  geom_density( fill = 'green', alpha = 0.25 ) + 
  scale_x_log10() +
  geom_vline( xintercept = 300) +
  theme_bw() +
  xlab( 'Measurement variance, in feet squared per second squared') +
  ylab( 'Probability density') +
  labs( title = paste0('Figure ', fig_no, '. Frequency density of streamflow measurement variance'))
fig_no <- fig_no + 1
  

# Drift vector
U  <- 'zero'
# Initial state vector from GA model
X0 <- as.matrix( as.numeric(coefficients(gam_rate_19)), 10, 1)
# Initial covariance of the state model
V0 <- gam_rate_19$Vp
# 
```

## Q matrix representing the error variance of the state vector specification options

```{r qmatrix, eval = FALSE}

# Alternate specification for the state covariance matrix Q
#   Diagonal matrix with unique specifications for the variance of two diagonal elements 
Q <- matrix(list('vvii',     0,      0,      0,       0,       0,       0,       0,        0,       0,
                     0,  'vvkk',     0,      0,       0,       0,       0,       0,        0,       0,
                     0,      0,  'vvkk',     0,       0,       0,       0,       0,        0,       0,
                     0,      0,      0,  'vvkk',      0,       0,       0,       0,        0,       0,
                     0,      0,      0,      0,   'vvkk',      0,       0,       0,        0,       0,
                     0,      0,      0,      0,       0,   'vvkk',      0,       0,        0,       0,
                     0,      0,      0,      0,       0,       0,   'vvkk',      0,        0,       0,
                     0,      0,      0,      0,       0,       0,       0,   'vvkk',       0,       0,
                     0,      0,      0,      0,       0,       0,       0,       0,    'vvkk' ,     0,
                     0,      0,      0,      0,       0,       0,       0,       0,        0,   'vvkk'),
            10, 10)

# Alternative specification for the state covariance matrix Q
Q = 'diagonal and equal'
Q = 'unconstrained'
Q = 'equalvarcov'


Q <- matrix(list('qb', 0, 0, 0, 0, 0, 0, '10*qb'), 3, 3)

Qgam_diag <- matrix(list('0.01193951*q',   0,   0,   0,   0,   0,   0,   0,   0,   0,
                    0 , '1.2730521*q',   0,   0,   0,   0,   0,   0,   0,   0,
                    0 ,   0, '10.550829*q',   0,   0,   0,   0,   0,   0,   0,
                    0 ,   0,   0, '0.7581285*q',   0,   0,   0,   0,   0,   0,
                    0 ,   0,   0,   0, '5.7352363*q',   0,   0,   0,   0,   0,
                    0 ,   0,   0,   0,   0, '1.0322490*q',   0,   0,   0,   0,
                    0 ,   0,   0,   0,   0,   0, '6.3561642*q',   0,   0,   0,
                    0 ,   0,   0,   0,   0,   0,   0, '1.4345857*q',   0,   0,
                    0 ,   0,   0,   0,   0,   0,   0,   0, '27.288992*q',   0,
                    0 ,   0,   0,   0,   0,   0,   0,   0,   0, '5.6616311*q'), 10, 10)

Qvo <- matrix(list(     '0.01193951*q',             0,    0,   0,   0,   0,   0,   0,   0,   0,
                    0 ,  '1.2730521*q', '-1.4770388*q',   0,   0,   0,   0,   0,   0,   0,
                    0 , '-1.4770388*q', '10.5508289*q',   0,   0,   0,   0,   0,   0,   0,
                    0 ,  '0.6363844*q', '-1.8314508*q',   0, '0.7581285*q',   0,   0,   0,   0,   0,   0,
                    0 ,  '1.1127029*q', '-7.3200227*q',   0, '5.7352363*q',   0,   0,   0,   0,   0,
                    0 ,  '0.7349047*q',  '1.8743107*q',   0,   0, '1.0322490*q',   0,   0,   0,   0,
                    0 , '-1.1603747*q',  '7.6315411*q',   0,   0,   0, '6.3561642*q',   0,   0,   0,
                    0 ,  '0.9505341*q', '-1.9276072*q',   0,   0,   0,   0, '1.4345857*q',   0,   0,
                    0 , '-2.3386594*q', '16.7687224*q',   0,   0,   0,   0,   0, '27.288992*q',   0,
                    0 , '-2.3704107*q',  '0.4574160*q',   0,   0,   0,   0,   0,   0, '5.6616311*q'), 10, 10)

Qgam_full <- matrix(list('0.0119395*q', 0,       0,              0,              0,              0,              0,               0,              0,              0,
            0,  '1.2730521*q', '-1.4770388*q',  '0.6363844*q',  '1.1127029*q', '-0.7349047*q', '-1.1603747*q',  '0.9505341*q',  '-2.3386594*q', '-2.3704107*q',
            0, '-1.4770388*q', '10.5508289*q', '-1.8314508*q', '-7.3200227*q',  '1.8743107*q',  '7.6315411*q', '-1.9276072*q',  '16.7687224*q',  '0.4574160*q',
            0,  '0.6363844*q', '-1.8314508*q',  '0.7581285*q',  '1.3384873*q', '-0.4494808*q', '-1.4025692*q',  '0.4963399*q',  '-2.9523673*q', '-1.0426028*q',
            0,  '1.1127029*q', '-7.3200227*q',  '1.3384873*q',  '5.7352363*q', '-1.3752243*q', '-5.3067451*q',  '1.4259650*q', '-12.0438808*q', '-0.4523644*q',
            0, '-0.7349047*q',  '1.8743107*q', '-0.4494808*q', '-1.3752243*q',  '1.0322490*q',  '1.4401821*q', '-0.4509775*q',   '3.0145552*q',  '1.3127885*q',
            0, '-1.1603747*q',  '7.6315411*q', '-1.4025692*q', '-5.3067451*q',  '1.4401821*q',  '6.3561642*q', '-1.4914250*q',  '12.6427408*q',  '0.4547216*q',
            0,  '0.9505341*q', '-1.9276072*q',  '0.4963399*q',  '1.4259650*q', '-0.4509775*q', '-1.4914250*q',  '1.4345857*q',  '-3.0854359*q', '-1.8801029*q',
            0, '-2.3386594*q', '16.7687224*q', '-2.9523673*q','-12.0438808*q',  '3.0145552*q', '12.6427408*q', '-3.0854359*q',  '27.2889916*q',  '0.5867990*q',
            0, '-2.3704107*q',  '0.4574160*q', '-1.0426028*q', '-0.4523644*q',  '1.3127885*q',  '0.4547216*q', '-1.8801029*q',   '0.5867990*q',  '5.6616311*q'), 10, 10)

Qvo <- matrix(list(0.0119395, 0,       0,              0,              0,              0,              0,               0,              0,              0,
            0,  1.2730521, -1.4770388,  0.6363844,  1.1127029, -0.7349047, -1.1603747,  0.9505341,  -2.3386594, -2.3704107,
            0, -1.4770388, 10.5508289, -1.8314508, -7.3200227,  1.8743107,  7.6315411, -1.9276072,  16.7687224,  0.4574160,
            0,  0.6363844, -1.8314508,  0.7581285,  1.3384873, -0.4494808, -1.4025692,  0.4963399,  -2.9523673, -1.0426028,
            0,  1.1127029, -7.3200227,  1.3384873,  5.7352363, -1.3752243, -5.3067451,  1.4259650, -12.0438808, -0.4523644,
            0, -0.7349047,  1.8743107, -0.4494808, -1.3752243,  1.0322490,  1.4401821, -0.4509775,   3.0145552,  1.3127885,
            0, -1.1603747,  7.6315411, -1.4025692, -5.3067451,  1.4401821,  6.3561642, -1.4914250,  12.6427408,  0.4547216,
            0,  0.9505341, -1.9276072,  0.4963399,  1.4259650, -0.4509775, -1.4914250,  1.4345857,  -3.0854359, -1.8801029,
            0, -2.3386594, 16.7687224, -2.9523673, 12.0438808,  3.0145552, 12.6427408, -3.0854359,  27.2889916,  0.5867990,
            0, -2.3704107,  0.4574160, -1.0426028, -0.4523644,  1.3127885,  0.4547216, -1.8801029,   0.5867990,  5.6616311), 10, 10)


for ( i in 1:10){
  for (j in 1:10){
    if( Qvo[i,j] != Qvo[j,i] ) {
      print(paste( i,j, Qvo[i,j], Qvo[j,i]))
    }
  }
}


Qtridiag_v_c_terms <-
       matrix(list('v',    'c',      0,      0,       0,       0,       0,       0,        0,       0,
                   'c',    'v',    'c',      0,       0,       0,       0,       0,        0,       0,
                     0,    'c',    'v',    'c',       0,       0,       0,       0,        0,       0,
                     0,      0,    'c',    'v',     'c',       0,       0,       0,        0,       0,
                     0,      0,      0,    'c',     'v',     'c',       0,       0,        0,       0,
                     0,      0,      0,      0,     'c',     'v',     'c',       0,        0,       0,
                     0,      0,      0,      0,       0,     'c',     'v',     'c',        0,       0,
                     0,      0,      0,      0,       0,       0,     'c',     'v',     'c',        0,
                     0,      0,      0,      0,       0,       0,       0,     'c',     'v' ,     'c',
                     0,      0,      0,      0,       0,       0,       0,       0,     'c',     'v'),
            10, 10)


Qv0 <- matrix(NA, ncol = 10, nrow = 10 )
for (i in 1:10){
  for (j in 1:10){
    Qv0[i,j] <- paste("qmult * ", format(V0[i,j], digits = 8))
  }
}

Qv0 <- matrix( list(0), 10, 10)
for (i in 1:10){
  Qv0[i,i] <- paste0('qmult*', V0[i,i])
}
Qvo <- unlist(Qv0)

q <- matrix(list(paste(     'qb', rep(0, 9), 
                 rep(0, 1), 'qb', rep(0, 8),
                 rep(0, 2), 'qb', rep(0, 7),
                 rep(0, 3), 'qb', rep(0, 6),
                 rep(0, 4), 'qb', rep(0, 5),
                 rep(0, 5), 'qb', rep(0, 4),
                 rep(0, 6), 'qb', rep(0, 3),
                 rep(0, 7), 'qb', rep(0, 2),
                 rep(0, 8), 'qb', rep(0, 1),
                 rep(0, 9), 'qb')), 10, 10)




```


## State-Space Estimation



```{r estimate_marss, eval = TRUE}
# Standardize kem file names

# model.gen=list(Z=Zt,A=A,R=R,U=U,Q=Q,x0=X0,V0=V0,tinitx=0)

# kemfit_19_Qdiag_equalfx <- MARSS(Yt, model = model.gen, control = list(maxit = 1000))
# saveRDS(kemfit_Q_full, file = 'kemfit_Q_full.rds')
# saveRDS(kemfit_Qgam_diag, file = 'kemfit_por_Qgam_full_unequal.rds')
kem_fit      <- readRDS( file = 'kemfit_19_Qdiag_equal_fx10.rds' )
kem_kfss     <- MARSSkfss( kem_fit )

# MARSSparamCI fails. Message: Error in dpari[time.varying] <- dparmat(MLEobj, time.varying, t = t) : 
# kem_paramCIs <- MARSSparamCIs( kem_fit  )

# MARSSboot looks like it will work, but the execution time for larger nboot values (1000) will be extensive
#  
# kem_boot     <- MARSSboot( kem_fit, param.gen = 'MLE', nboot = 2 )

# Error in dpari[time.varying] <- dparmat(MLEobj, time.varying, t = t) :
# kem_FisherI  <- MARSSFisherI( kem_fit )

# kemfit_Qtridiag_v_c_terms <- MARSS(Yt, model = model.gen, control = list(maxit = 2000))


# Save command for the state space model
# saveRDS(kemfit_por, file = 'Models/kemfit_por.rds')



# Innovations bootstrapping uses the innovations resampling and can only be done if there are no missing values in the data.
#   MARSSboot( kemfit_por, nboot = 1000, output = 'parameters', sim = 'innovations', param.gen = 'MLE' )

# kemfit_19_Qdiag_equal_fx  <- MARSS(Yt, model = model.gen, control = list(maxit = 4000))
# kemfit_19_Qdiag_equal_fx1 <- MARSS(Yt, model = model.gen, control = list(maxit = 1000), inits = coef(kemfit_19_Qdiag_equal_fx))


```

## Conmpute standard errors of flow estimates


```{r }
ytT.se <- matrix(NA, nrow( df_stage_flow_allDays))

for (i in 1:nrow( df_stage_flow_allDays)){
  ytT.se[i] <- matrix(Zt[1, , i], 1, 10) %*% matrix(kem_kfss$VtT[, , i], 10, 10) %*% matrix(Zt[1,,i], 10, 1)
}
ytT.se <- sqrt(ytT.se)


plot(ytT.se)
```

Success! abstol and log-log tests passed at 571 iterations.
Alert: conv.test.slope.tol is 0.5.
Test with smaller values (<0.1) to ensure convergence.

MARSS fit is
Estimation method: kem 
Convergence test: conv.test.slope.tol = 0.5, abstol = 0.001
Estimation converged in 571 iterations. 
Log-likelihood: -583.9928 
AIC: 1171.986   AICc: 1172.117   
 
        Estimate
Q.qicpt  172.694
Q.qknot    0.105
Initial states (x0) defined at t=0

Standard errors have not been calculated. 
Use MARSSparamCIs to compute CIs and bias estimates.

## GAM for stage_flow_day_19 meaurements

```{r eval_state_space, fig.width = 9, fig.height = 16}


# Store results of MARSS modeling
df_stage_flow_allDays$ytT     <- as.vector(kem_fit$ytT)
df_stage_flow_allDays$ytT.se  <- as.vector(ytT.se)
df_stage_flow_allDays$cl_025 <- df_stage_flow_allDays$ytT + qnorm(0.025) * df_stage_flow_allDays$ytT.se
df_stage_flow_allDays$cl_975 <- df_stage_flow_allDays$ytT + qnorm(0.975) * df_stage_flow_allDays$ytT.se


df_stage_flow_allDays %>% 
  rename( 'flow_daily_kem'  = 'ytT',
          'measured_flow'  = 'flow_discrete_mea', 
          'lower_95_flow'  = 'cl_025',
          'upper_95_flow'  = 'cl_975') %>% 
           addWaterYear  %>% 
  dplyr::filter( waterYear %in% c( 1995, 2000, 2005, 2010, 2015) ) %>% 
  ggplot( aes( x = Date, y = flow_daily_kem )) +
  geom_line( color = 'blue' ) +
  geom_point( aes( y = measured_flow, color = qualifier), size = 1.5) +
  geom_line(  aes( y = upper_95_flow ), color = 'salmon', linetype = 'solid') +
  geom_line(  aes( y = lower_95_flow ), color = 'salmon', linetype = 'solid') +
  # scale_y_log10() +
  theme_bw() +
  xlab( 'Date') + 
  ylab( 'Daily mean streamflow, in cubic feet per second') +
  theme( legend.position = 'bottom') +
  facet_wrap( ncol = 1, waterYear ~ ., scales = 'free' )


```


```{r compare_rating_kem, eval = FALSE}
# Compare rating with kfgam estimates

stage_knots_19 <- predict( gam_rate_19, newdata = data.frame( stage = df_stage_flow_allDays$stage_mean_day),
                                           type = 'lpmatrix')

stage_knots_19 <- predict( gam_rate_19, type = 'lpmatrix')



stage_knots_19 <- cbind(rating19$stage, stage_knots_19) 
colnames(stage_knots_19)[1] <- 'stage'

tmp <- discrete_meas %>% 
  left_join( df_stage_flow_allDays, by = c('measurement_dt' = 'day') )

ndx_kfss <- which( df_stage_flow_allDays$Date %in% discrete_meas$measurement_dt )
flow_xtT <- matrix(NA, nrow = 731, ncol = length( ndx_kfss ) )

stage_knots_19_matrix <- as.matrix(stage_knots_19[,2:11])

# for (i in 1:length(ndx_kfss)){
#   flow_xtT[ , i ] <- stage_knots_19_matrix %*% matrix(kemfit_por_Qdiag_unequal_kfss$xtT[ 1:10, ndx_kfss[ i ]], 10, 1)
# }


for (i in 1:length(ndx_kfss)){
  flow_xtT[ , i ] <- stage_knots_19_matrix %*% matrix(kem_kfss$xtT[ 1:10, ndx_kfss[ i ]], 10, 1)
}

state_coefs_19  <- matrix(kem_kfss$xtT[ 1:10,    1], 10, 1)
state_coefs_20  <- matrix(kem_kfss$xtT[ 1:10, 5972], 10, 1)
state_coefs_21  <- matrix(kem_kfss$xtT[ 1:10, 6918], 10, 1)
state_coefs_22  <- matrix(kem_kfss$xtT[ 1:10, 8746], 10, 1)

rating19$flow_kfgam19 <- stage_knots_19_matrix %*% state_coefs_19
rating19$flow_kfgam20 <- stage_knots_19_matrix %*% state_coefs_20
rating19$flow_kfgam21 <- stage_knots_19_matrix %*% state_coefs_21
rating19$flow_kfgam22 <- stage_knots_19_matrix %*% state_coefs_22


# Retrieve daily flows for rating period 19

rating19[, c('stage', 'rated_flow_19', 'flow_kfgam19', 'flow_kfgam20',
                    'flow_kfgam21', 'flow_kfgam22')] %>% 
  rename( flow_rating19 = rated_flow_19) %>% 
  gather( key = 'Estimator', value = 'flow_cfs', flow_rating19, flow_kfgam19, flow_kfgam20, 
          flow_kfgam21, flow_kfgam22, -stage) %>% 
  mutate( Estimator = factor(Estimator, levels = c('flow_rating19', 'flow_kfgam19', 'flow_kfgam20', 
                                                   'flow_kfgam21', 'flow_kfgam22'))) %>% 
  ggplot( aes(x = flow_cfs, y = stage, color =  Estimator )) +
  geom_line() +
  theme_bw() +
  theme( legend.position = 'right') +
  xlab( 'Streamflow, in cubic feet per second' ) +
  ylab( 'Stage, in feet above gage datum' ) + 
  labs(title = 'Stage-Flow Ratings at 04122500 Pere Marquette River at Scottville, MI') 


```



```{r summarize_kfgam_rating, eval = FALSE }

# Convert matrix containing flow ratings during por to dataframe
df_flow_xtT <- as.data.frame( flow_xtT)

# Flows for fixed stages
flow_xtT_min_stage <- apply(flow_xtT[,-76], 1, FUN = min)
flow_xtT_max_stage <- apply(flow_xtT, 1, FUN = max)
flow_xtT_ave_stage <- apply(flow_xtT, 1, FUN = mean)
flow_xtT_med_stage <- apply(flow_xtT, 1, FUN = median)


flow_xtT_p05_stage <- apply(flow_xtT, 1, FUN = quantile, probs = 0.05)
flow_xtT_p95_stage <- apply(flow_xtT, 1, FUN = quantile, probs = 0.95)

flow_xtT_p01_stage <- apply(flow_xtT, 1, FUN = quantile, probs = 0.01)
flow_xtT_p99_stage <- apply(flow_xtT, 1, FUN = quantile, probs = 0.99)



df_flow_xtT_stats  <- data.frame( stage = rating19$stage, flow_xtT_min_stage = flow_xtT_min_stage,
                                  flow_xtT_max_stage = flow_xtT_max_stage, 
                                  flow_xtT_ave_stage = flow_xtT_ave_stage,
                                  flow_xtT_med_stage = flow_xtT_med_stage,
                                  flow_xtT_p01_stage = flow_xtT_p01_stage,
                                  flow_xtT_p99_stage = flow_xtT_p99_stage)

df_flow_xtT_stats %>% 
  gather( key = 'stat', value = 'flow_cfs', flow_xtT_min_stage, flow_xtT_ave_stage, 
          flow_xtT_max_stage, flow_xtT_med_stage, flow_xtT_p01_stage, 
          flow_xtT_p99_stage) %>% 
  ggplot( aes( x = flow_cfs, y = stage, color = stat)) +
  geom_line() +
  theme_bw() +
  theme( legend.position = 'bottom') +
  scale_x_sqrt( breaks = c(1, 10, 50, 100, 500, 1000, 2000, 3000, 4000, 5000, 6000, 7000)) +
  scale_y_continuous( breaks = seq(1, 8, 1))


```

```{r eval = FALSE}


rating19[, c('stage', 'rated_flow_19', 'flow_kfgam19', 'flow_kfgam20',
                    'flow_kfgam21', 'flow_kfgam22')] %>% 
  rename( flow_rating19.0 = rated_flow_19) %>% 
  gather( key = 'Estimator', value = 'flow_cfs', flow_rating19, flow_kfgam19, flow_kfgam20, 
          flow_kfgam21, flow_kfgam22, -stage) %>% 
  mutate( Estimator = factor(Estimator, levels = c( 'flow_rating19', 'flow_kfgam19', 
          'flow_kfgam20', 'flow_kfgam21', 'flow_kfgam22' ) )) %>% 
  ggplot( aes( x = day, y = flow_daily  )) +
  geom_line( color = 'blue', alpha = 0.5 ) +
  geom_line( aes( y = ytT ), color = 'red', alpha = 0.5 ) +
  theme_bw() + 
  ylab( 'Stage, in feet above streamgage datum') +
  xlab( 'Streamflow, in cubic feet per second' )
  





```


```{r annimate_rating, eval = FALSE}

#> Loading required package: ggplot2

# We'll start with a static plot
#  p <- ggplot(iris, aes(x = Petal.Width, y = Petal.Length)) + 
#  geom_line() +
#   transition_states(Species,
#                     transition_length = 2,
#                     state_length = 1)
 
# p


# animate(p, renderer = file_renderer('~/animation/'))[1:6]
# animate(p, 'output.gif')

p <- rating_long %>% 
  ggplot( aes(x = rated_flow, y = stage, color =  rating_no )) +
  geom_point() +
  transition_states(rating_no,
                    transition_length = 1,
                    state_length = 1) +
  theme_bw() +
  theme( legend.position = 'right') +
  xlab( 'Streamflow, in cubic feet per second' ) +
  ylab( 'Stage, in feet above gage datum' ) + 
  labs(title = 'Stage-Flow Ratings at 04122500 Pere Marquette River at Scottville, MI') +
  geom_line( data = rating19, aes( x = rated_flow_19, y = stage), color = 'black', linetype = 'dots' )

animate(p, nframes = 6, fps = 2)
```

```{r }

rating_long %>% 
  ggplot( aes(x = rated_flow, y = stage, color =  rating_no )) +
  geom_point() +
  theme_bw() +
  theme( legend.position = 'right') +
  xlab( 'Streamflow, in cubic feet per second' ) +
  ylab( 'Stage, in feet above gage datum' ) + 
  labs(title = 'Stage-Flow Ratings at 04122500 Pere Marquette River at Scottville, MI') +
  geom_line( data = rating19, aes( x = rated_flow_19, y = stage ), color = 'black', linetype = 'dots' )

```


```{r, fig.height = 24 }

# Initialize vector to contaoin kem output
tmp_vec <- matrix( flow_xtT, ncol = 1, byrow = TRUE)

tmp_df <- data.frame( Date =  rep(df_stage_flow_allDays$Date[ndx_kfss], each = nrow(rating19)),
                      stage = rep(rating19$stage, times = length(ndx_kfss)),
                      kem_flow =  tmp_vec )
tmp_df <- tmp_df %>% addWaterYear()
rm(tmp_vec)


tmp_df %>% 
  dplyr::filter( waterYear %in% c( 1995, 2000, 2005, 2010, 2015, 2019) ) %>%
  ggplot( aes(x = kem_flow, y = stage, group = waterYear ), alpha = 0.25) +
  geom_line( size = 0.5, color = 'blue', alpha = 0.5) +
  labs(title = 'Date of Measurement:' ) + 
  theme_bw() +
  facet_wrap( ~ waterYear, ncol = 2) +
  theme( legend.position = 'right') +
  xlab( 'Streamflow, in cubic feet per second' ) +
  ylab( 'Stage, in feet above gage datum' ) + 
  labs(title = 'Stage-Flow Ratings at 04122500 Pere Marquette River at Scottville, MI') +
  geom_line( data = rating19, aes( x = rated_flow_19, y = stage, group = NULL), color = 'red', size = 1 ) +
  geom_line( data = rating21, aes( x = rated_flow_21, y = stage, group = NULL), color = 'forestgreen', size = 1 )

```



rating_days_measured <- tmp %>% 
  ggplot( aes(x = flow_daily, y = stage )) +
  geom_point( color = 'blue', size = 0.5 ) +
    transition_time(Date) +
  labs(title = 'Date of Measurement: {frame_time}' ) + 
  theme_bw() +
  theme( legend.position = 'right') +
  xlab( 'Streamflow, in cubic feet per second' ) +
  ylab( 'Stage, in feet above gage datum' ) + 
  # labs(title = 'Stage-Flow Ratings at 04122500 Pere Marquette River at Scottville, MI') +
  geom_line( data = rating19, aes( x = rated_flow_19, y = stage), color = 'black', linetype = 'dots' )

animate(rating_days_measured, nframes = 155, fps = 10)
```