---
title: "A State-Space Model for Adaptive Stage-Discharge Rating Estimation using Discrete Flow Measurements"
author: "David J Holtschlag"
date: "`r format(Sys.Date(), '%A %b %d, %Y') `"
output: 
  html_document:
    toc: true
    toc_depth: 4
  always_allow_html: yes
bibliography: bibliography.bib
---

## Abstract

An adaptive stage-discharge estimation approach is developed to track persistent changes in the relation between water-surface elevations (stages) and streamflow discharges (flows). The approach updates parameters of a cubic spline regression function describing the stage-flow relation using data from periodic discrete flow measurements. The spline parameters form the state vector of a state-space model, while the measurement vector contains the discrete flow measurements. The state-space model is solved iteratively using a Kalman filter. The approach is illustrated using daily data from U.S. Geological Survey streamgage 04122500 Pere Marquette River at Scottville, Mich. Results show the adaptation of a stage-flow relation consistent with rating 18 (applied at the streamgage from 1986-1994), to a relation consistent with rating 21 (applied at the streamgage from 2013-2019), based on 188 discrete flow measurements during the intervening period. In addition, the Kalman filter is used to compute the magnitudes and uncertainties of daily flows at the streamgage for selected years. Once the state-space model estimated from daily values, however, the model can be applied at regular unit (less than daily time intervals) time steps by a simple adjustment of the process variance in the state equation. Persistent or long-term changes in the stage-flow relation can be tracked with this univariate model.  Extension of this model to track intermittent changes, associated with variable ice-backwater for example, would require a multivariate approach involving a streamflow network that is beyond the scope of this model. 

## Background

The U.S. Geological Survey (USGS) computes records of streamflow at about 8,500 active streamgages throughout the United States. Streamgages measure flow from the upstream basin, which is generally topographically defined.  Computations are commonly based on continual measurements of stage at 10- to 60-minute intervals. The stage data is used to compute corresponding values of flow based on a empirical relation between stage and flow commonly referred to as a rating curve or table. The rating describes a monotonically increasing relation between stage and flow. This rating is developed and maintained on the basis of discrete, contemporaneous measurements of stage and flow that span the typical range of flow conditions at the streamgage. 

The stage-flow relation can change gradually over time due to persistent changes in the channel or overbank hydraulic characteristics, or change abruptly owing to intermittent effects, such as increased backwater associated with aquatic plant growth or ice formation. Once the cause of the intermittent changes are no longer a factor, the stage-flow relation may revert to nominal conditions, consistent with possible persistent changes at the stream site.  

The distinction between persistent and intermittent changes is somewhat arbitrary, but practically determined by the measurement frequency. Abrupt intermittent changes, such as ice backwater effects, can occur on a time scale of hours to days during periods when no discrete flow measurements occur to identify the onset and cessation of these events. The frequency of discrete flow measurements is constrained by the cost of field visits. 

### Scope

This paper describes the preliminary development and testing of an automated approach for estimating a stage-flow relation for a streamgage that sequentially adapts to persistent changes in rating conditions based on discrete flow measurement data. A persistent change is considered to be a slow change in rating conditions relative to the frequency of discrete measurements available to characterize the change. Adaptive tracking can be applied in real-time, considering only past discrete measurements, or offline, considering discrete flow information before and after the time of estimation. Only offline tracking is presented in the paper.  Implementation of the technique requires continual stage data and discrete flow data for implementation. In contrast to persistent changes, intermittent changes represent sudden changes in rating conditions that may be associated with transient effects like those due to ice-affected streamflow. Tracking intermittent changes requires more precise information on the timing of onset and cessation of the intermittent effects than is possible from discrete flow measurements.  Intermittent changes may benefit from analysis of multiple streamgages in a network, but is beyond the scope of this study. 


```{r setup, include=FALSE}
# Libraries needed in the analysis are loaded here
library( "rethinking" )
options(mc.cores = parallel::detectCores())
suppressPackageStartupMessages(library(dataRetrieval))
suppressPackageStartupMessages(library(lubridate))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggthemes))
suppressPackageStartupMessages(library(splines))
suppressPackageStartupMessages(library(mgcv))
suppressPackageStartupMessages(library(fitdistrplus))
suppressPackageStartupMessages(library(stringr))
suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(ggrepel))
suppressPackageStartupMessages(library(ggcorrplot))
# strcmp and other Matlab like functions
suppressPackageStartupMessages(library(pracma))
# Package expss contains the function vlookup
suppressPackageStartupMessages(library(expss))
# Package to plot distribution
suppressPackageStartupMessages(library(ggfortify))
# Multivariate Autoregressive State Space Model
suppressPackageStartupMessages(library(MARSS))
suppressPackageStartupMessages(library(matrixcalc))
# The broom package takes the messy output of built-in functions in R, such as lm, nls, or t.test, and turns them into tidy data frames.
suppressPackageStartupMessages(library(broom))
suppressPackageStartupMessages(library(gganimate))
suppressPackageStartupMessages(library(gifski))
# scales is used in density plot
suppressPackageStartupMessages(library(scales))
suppressPackageStartupMessages(library(ggmap))
suppressPackageStartupMessages(library(ggsn))
suppressPackageStartupMessages(library(sf))
suppressPackageStartupMessages(library(usmap))
suppressPackageStartupMessages(library(FAdist))
knitr::opts_chunk$set(echo = FALSE, fig.width = 8)
```



```{r specify_gage, echo = FALSE, results = 'hide' }

# The starting and ending dates of the model
#   Based on the beginning of unit stage date and endin on the first flow measurement of water year 2020
date_beg_model  <- as.Date( '1989-09-30', format = '%Y-%m-%d' ) 
date_end_model  <- as.Date( '2019-10-08', format = '%Y-%m-%d' )

# initialize counter for figures
fig_no <- 0
tab_no <- 0

# Retrieve gage info for specified site Number
site_no    <- '04122500'

# Used in accessing site_info
site_file  <- paste0('site_', site_no, '.RData')

if( file.exists(site_file)){
  # print(paste('Loading site_info for', site_no, 'from RData file.'))
  load(site_file)
} else {
  # print(paste('Retrieving site_info for', site_no, 'from NWIS' ))
  site_info  <- readNWISsite(site_no)

  # Standardize field names
  site_info  <- renameNWISColumns(site_info)
  # Standarize capitalization of site name
  # site_info$station_nm <- str_to_title( site_info$station_nm )
  site_info$station_nm <- "Pere Marquette River At Scottville, Mich."
  save(site_info, file = site_file)
}

```



```{r show_site_info, fig.height = 2, warning = 'hide', results = 'hide', eval = FALSE }

site_info %>% 
  rename( drain_area_mi2 = drain_area_va,
          latitude       = dec_lat_va,
          longitude      = dec_long_va,
          altitude_ft    = alt_va ,
          altitude_datum = alt_datum_cd) %>% 
  dplyr::select( site_no, station_nm, drain_area_mi2, latitude, longitude, altitude_ft, altitude_datum) %>% 
  kable( caption = paste0('Table ',tab_no,'. Summary streamgage information' )) %>% 
  kable_styling()


```
### Study Area

USGS streamgage 04122500 Pere Marquette River at Scottville, Mich. was selected for this study. (fig. 1). The Pere Marquette River drains a basin in the west central lower peninsula of Michigan.  The basin has a drainage area of 681 mi^2^ at the streamgage. The basin is heavily forested and includes parts of the Huron-Manistee National Forest. Streamflow has been computed continuously at the site starting in August 1939 and was active at the time of this paper (October 2020). The streamflow rating is commonly ice-affected during the winter months.



```{r register_key, echo = FALSE,  results = 'hide', message = 'hide', warning = 'hide'}

# this sets your google map for this session
# Read google key from file
# google_key <- read_file('maps/google_key.txt')

# Register the key and delete it following registry
# register_google(key = google_key)
# rm('google_key')

# check if key is saved
# has_google_key()

# Set up bounding box for map

# site_map <- get_map( location = c(lon = site_info$dec_long_va + 0.10, lat = site_info$dec_lat_va),
#                      zoom = 9, source = 'google', maptype = "terrain")
# 
# write_rds(site_map, path = "C:/Home/projects/spline_rating_curves/stage_discharge_rating/maps/site_map")

suppressMessages(map <- read_rds(path = "C:/Home/projects/spline_rating_curves/stage_discharge_rating/maps/site_map"))

invisible(watershed    <- st_read("C:/Home/ws_012/fhwa_flood_regions/gis/shapefiles/04122500.shp"))
watershed_df <- data.frame( lon = st_coordinates(watershed)[,1],
                           lat = st_coordinates(watershed)[,2]) 
```


```{r plot_site_map, fig.height = 7, fig.width = 7, message = FALSE, warning = 'hide' }
 fig_no <- fig_no + 1

# Site map with streamgage and drainage basin shown
vicinity <- map %>% 
  ggmap() +
  geom_point( data = data.frame( lon = site_info$dec_long_va, lat = site_info$dec_lat_va),
              aes( x = lon, y = lat), color = 'red', size = 3) +
  geom_polygon( data = watershed_df, aes( x = lon, y = lat), color = 'lightgrey', alpha = 0.15 ) +
  xlab( 'Longitude' ) + 
  ylab( 'Latitude') +
  scalebar(x.min = -86.95622, x.max = -85.29841, y.min =  43.40782, y.max =  44.5, 
           transform = TRUE, dist_unit = 'mi', dist = 10, location = 'bottomleft',
           model = "GRS80", height = 0.02, st.size = 3, st.dist = .015,
           border.size = 1, box.fill = c('black', 'white')) +
  annotate('text', x = -85.88, y = 43.85, label = 'Pere Marquette Basin' ) +
  annotate('text', x = -86.28, y = 43.99, label = '04122500') +
  ggtitle( paste0('Figure ', fig_no, '. U.S. Geological Survey streamgage 04122500 Pere Marquette \nRiver at Scottville, Mich. and drainage basin'))

mi_map <- map_data('state')

mi_inset <-mi_map %>% 
  dplyr::filter( region == 'michigan') %>% 
  ggplot( aes( x = long, y = lat, group = group)) +
  geom_polygon( fill = 'black') +
  geom_point( data = data.frame( long = site_info$dec_long_va, lat = site_info$dec_lat_va),
              aes( x = long, y = lat, group = NA), color = 'red', size = 2) + 
  xlab('') +
  ylab('')


vicinity + inset(grob = ggplotGrob(mi_inset), xmin = -86.85, xmax = -86.4, 
                                                      ymin =  44.25, ymax =  44.55)


```


## Data Components

Continual streamflow data are commonly computed on the basis of continual stage data, and an empirical relation between stage and flow. The continual intervals are generally less than an hour in length, and are referred to as unit data.  The relation between stage and flow is based on periodic discrete stage and flow measurements obtained during a field visit to the streamgage. A set of discrete measurements is used to manually define a curvilinear relation between stage and flow over an extended stage range. The stage-flow relation is commonly referred to as a rating curve or table. Discrete measurements continue throughout the period of streamgage operation to confirm, update, extend, and revise the stage-flow relation as needed accurately compute streamflow. 

Several data components are critical in computing unit streamflow, and derivative products, such as daily (mean) streamflow. Daily streamflow is the primary streamflow statistic used in many water-resource applications. Daily streamflow is commonly used, for example, to compute flow-duration and high- and low-frequency characteristics widely used in design and decision making. Therefore, daily streamflow was the first streamflow statistic to be accessible electronically. 

Historically, unit stage and flow  values have been considered ancillary data of secondary importance to daily streamflow. High cost of electronic store of these data were prohibitive. More recently, however, both unit stage and flow data has been stored in accessible formats along with information on discrete flow measurements. Ready access to these data have been a limiting factor in developing the analysis presented in this paper. 


### Daily Streamflow

Daily mean streamflow data for the Pere Marquette River were obtained for the period of record using the USGS National Water Information System (NWIS) by use of the R package 'dataRetrieval' [@DeCicco2019]. Some of the earlier ancillary data used in computing daily mean flows, however, is only available from archived paper records. In particular, unit stage data and discrete flow measurement data are generally available for Pere Marquette River in electronic format since `r format(date_beg_model, '%b %d, %Y')`. The most recent data used in the analysis was `r format(date_beg_model, '%b %d, %Y')`. In this paper, the period of electronic data availability is referred to as the period of analysis (POA). Data summaries presented in the following sections are restricted to this period.



```{r pub_daily_flow}
# Load daily flows to estimate missing daily mean stage 
if( file.exists('flow_daily_pub.RData')){
  # print('Loading published daily_means from RData file.')
  load('flow_daily_pub.RData')
} else {
  # print('Retrieving daily flow from NWIS')
  flow_daily_pub <- readNWISdv( site_no, parameterCd = '00060', startDate = date_beg_model,
                             endDate = date_end_model)
  # Standardize names
  flow_daily_pub <- renameNWISColumns( flow_daily_pub )
  save(flow_daily_pub, file = 'flow_daily_pub.RData')
}

flow_hpdi <- HPDI( flow_daily_pub$Flow, prob = 0.95 )

```

***

> Probability Distribution of Daily Flows


The empirical probability distribution describes the relative likelihood of daily flow magnitudes.  Daily flows at Pere Marquette River during the period of analysis varied from `r flow_hpdi[[1]]` and `r flow_hpdi[[2]]` ft^3^/s, 95-percent of the time.  The minimum and maximum daily flows during this period were `r min(flow_daily_pub$Flow)` and `r max(flow_daily_pub$Flow)`, respectively. The mean flow was 787 ft^3^/s. The probability distribution of the flows are asymmetric (right skewed). The natural logarithm of flows, however, is consistent with a three parameter gamma distribution (fig. 2). This distribution is an important consideration in specifying a model to track changes in stage-flow relations.   

```{r prob_den_flow, warning = FALSE }
# Estimate parameters for lgamma3
gam3_parm <- fitdist(log(flow_daily_pub$Flow), 'gamma3', start = list(shape = 1, scale = 1, thres = 1),
        upper = c(Inf, Inf, min(log(flow_daily_pub$Flow))))

gam3_seq <- seq(gam3_parm$estimate[3], max(log(flow_daily_pub$Flow)), by = 0.02)
gam3_den <- dgamma3(gam3_seq, shape = gam3_parm$estimate[1],
                    scale = gam3_parm$estimate[2], thres = gam3_parm$estimate[3])
gam3_df  <- data.frame(gam3_seq, gam3_den)


fig_no <- fig_no + 1

# Plot the empirical density of log(Flow) with theoretical GAMMA3 Density 

flow_daily_pub %>% 
  ggplot( aes( Flow)) +
  geom_density( fill = 'lightgrey', alpha = 0.5 ) +
  theme_bw() +
  scale_x_continuous( trans = 'log', breaks = c( 300, 500, 1000, 2000, 3000, 4000, 5000), limits = c(300, 5000) ) + 
  geom_vline( xintercept =  mean(flow_daily_pub$Flow), linetype = 'dashed') +
  annotate('text', x = 980, y = 0.2, label = paste0('Mean = ', format(mean(flow_daily_pub$Flow), digits = 0 ))) +
  geom_line( data = gam3_df, aes( x = exp(gam3_seq), y = gam3_den), color = 'red' ) +
  xlab('Flow, in cubic feet per second') +
  ylab('Probability density') +
  annotate('text', x = 350, y = 1.12, label = 'Theoretical \nGamma Density', color = 'red', size = 3.8) +
  # annotate('text', x = exp(7.65), y = 1.09, label = paste0('dgamma3( shape = ', format( gam3_parm$estimate[1], digits = 4),
  #                                   ', scale = ', format( gam3_parm$estimate[2], digits = 4),
  #                                   ', thres = ', format( gam3_parm$estimate[3], digits = 4),' )'), color = 'red', size = 3.8) +
  annotate('segment', x = 450, y = 1.10, xend = 570, yend = 1.10, color = 'red' ) +
  annotate('text', x = exp(6.26), y = 0.15, label = 'Empirical Density' ) +
  ggtitle( paste0( 'Figure ', fig_no, '. Distribution of daily flows at 04122500 Pere Marquette River at \nScottville, Mich. based on data from ', format(date_beg_model, '%B %d, %Y'), ' to ', format(date_end_model, '%B %d, %Y')) ) 

```

***

### Discrete Flow Measurements

```{r read_discrete_meas_stage_flow, echo = FALSE,  warning = FALSE }

if( file.exists('discrete_meas.RData')){
  # print('Loading discrete flow measurement data from RData file.')
  load('discrete_meas.RData')
} else {
  # print('Retrieving daily flow from NWIS')
  discrete_meas   <- readNWISmeas(site_no, expanded = TRUE, convertType = TRUE)
  # Standardize field names
  discrete_meas  <- renameNWISColumns(discrete_meas)
  save(discrete_meas, file = 'discrete_meas.RData')
}

# Standardize attributes names for hydraulic control
control_type <- 
  recode_factor( discrete_meas$control_type_cd, 
                 'Clear'    = 'Clear Channel',
          'VegetationLight' = 'Light Vegetation',
          'DebrisLight'     = 'Light Debris',   
          'IceShore'        = 'Shore Ice',
          'DebrisModerate'  = 'Moderate Debris',
          'IceCover'        = 'Ice Cover', .ordered = TRUE ) 

# Add control type to data.frame
discrete_meas$control_type <- control_type


# Source 
# https://help.waterdata.usgs.gov/codes-and-parameters/discharge-measurement-quality-code
# ---------------------------------
# Code  Description
# ---------------------------------
# E     Excellent    The data is within 2% (percent) of the actual flow {95 percent of the time}
# G     Good         The data is within 5% (percent) of the actual flow
# F     Fair         The data is within 8% (percent) of the actual flow
# P     Poor         The data are >8% (percent) of the actual flow

discrete_meas <- discrete_meas %>% 
  dplyr::rename( 'qualifier' = 'measured_rating_diff') %>% 
  mutate( se_pct = case_when( qualifier == 'Excellent' ~  2.0,
                              qualifier == 'Good'      ~  5.0,
                              qualifier == 'Fair'      ~  8.0,
                              qualifier == 'Poor'      ~ 12.0),
          se_flow = se_pct/100 * discharge_va ) 


# Read rating history
rating_history <- read.csv( 'Data/04122500/rating_period_start_end.txt', sep = '\t', header = TRUE,
                            colClasses = c('numeric', 'Date', 'Date'), comment.char = '#')

# Rename discrete measurements accuracy and period of analysis
discrete_meas <- discrete_meas %>% 
  rename('stage_discrete_mea'  = 'gage_height_va',
         'flow_discrete_mea'   = 'discharge_va') %>% 
  filter(qualifier            != 'Unspecified' | qualifier == !is.na(qualifier)) %>%
  mutate('qualifier'           =  ordered(qualifier, levels = c('Excellent', 'Good', 'Fair', 'Poor')),
         'Date'                =  measurement_dt,
         'rating_no'    =  case_when( Date >= rating_history$start_date[1] & Date <  rating_history$end_date[1] ~ 18,
                                             Date >= rating_history$start_date[2] & Date <  rating_history$end_date[2] ~ 19,
                                             Date >= rating_history$start_date[3] & Date <  rating_history$end_date[3] ~ 20,
                                             Date >= rating_history$start_date[4] & Date <  rating_history$end_date[4] ~ 21),
         'period_of_analysis'  =  ifelse( Date >= date_beg_model & Date <= date_end_model, TRUE, FALSE)) %>% 
  dplyr::select( 'Date', 'measurement_nu', 'stage_discrete_mea', 'flow_discrete_mea', 'qualifier', 
                 'control_type', 'rating_no', 'period_of_analysis', 'se_pct', 'se_flow') %>% 
  addWaterYear()
```


Figure `r fig_no + 1` shows a time series plot of discrete flow and stage measurements during the period for which these data were electronically accessible. There are no apparent trends in the magnitudes or variances of the flow. There may be a tendency, however, for minimum annual stages to trend downward during the period. In particular, annual minimum stage value in the 10-year period centered on 1990 seem to average about 1.5 ft. In contrast, annual minimum annual stage values in the 10-year period centered in 2015 seem to average about 0.5 ft lower. Given the apparent lack of trend in discrete flow values, the lowering of the minimum water levels may be the result of channel degradation. The decreasing trend in stage values not apparent in the loess regression line through all the range of data. 


```{r plot_discrete_data, message = FALSE }
# Plot discrete Flow and Stage
fig_no <- fig_no + 1
discrete_meas %>% 
  dplyr::select( Date, flow_discrete_mea, stage_discrete_mea ) %>% 
  dplyr::rename( Flow = flow_discrete_mea,
                 Stage = stage_discrete_mea ) %>% 
  gather( 'Parameter', 'Magnitude',  -Date ) %>% 
  ggplot( aes( x = Date, y = Magnitude, group = Parameter ) ) +
  geom_point() +
  facet_grid( Parameter ~ ., scales = 'free_y' )  +
  theme_bw() +
  geom_smooth() +
  scale_x_date( breaks =        seq(as.Date("1985/1/1"), as.Date("2020/1/1"), by = "5 years"),
                labels = format(seq(as.Date("1985/1/1"), as.Date("2020/1/1"), by = "5 years"), '%Y')) +
  # geom_vline( xintercept = c(date_beg_model, date_end_model), linetype = 'dashed') +
  # geom_vline( xintercept = as.Date( c('1986-10-01', '1994-10-25', '2011-03-01', '2013-10-01')) , color = 'blue') + 
  # annotate('segment', x = date_beg_model, xend = date_end_model, y = 3500, yend = 3500 ) +
  ylab( paste0('Stage, in feet above ', site_info$alt_va, ' ', site_info$alt_datum_cd,
  '                           Flow, in cubic feet per second') )  +
  ggtitle( paste0('Figure ', fig_no, '. Discrete flow measurements at 04122500 Pere Marquette River at Scottville, \nMich. during the period with accessible discrete measurement data.') )


```


***

> Hydraulic Controls and Accuracy Qualifiers

Discrete flow measurement data for Pere Marquette River occurring during the period of analysis were retrieved from the USGS National Water Information System (NWIS) using the R package dataRetrieval [@DeCicco2019]. Table `r tab_no + 1` summarizes the type of hydraulic control and the accuracy of discrete flow measurements provided by the qualifier. The most common hydraulic control conditions were a 'Clear Channel', and the most common measurement accuracy qualifier was 'Good.'  


The qualifiers assigned to discrete flow measurements by hydrographers in the field were used to estimate uncertainties of the measured flows.  In this paper, a qualifier of 'Excellent' was associated with a standard error of 2 percent; 'Good,' with a standard error of 5 percent; 'Fair,' with a standard error of 8 percent; and 'Poor,' with a standard error of 12-percent. The associations are loosely based on the 'Discharge Measurement Quality Code' descriptions by USGS NWIS at https://help.waterdata.usgs.gov/codes-and-parameters/discharge-measurement-quality-code, accessed Jan. 18, 2020.   
  
```{r measurement_type_freq, echo = FALSE, message = FALSE}

# Increment table counter
tab_no <- tab_no + 1
table( discrete_meas$control_type, discrete_meas$qualifier ) %>%
  rbind(c(sum(discrete_meas$qualifier == 'Excellent'),
          sum(discrete_meas$qualifier == 'Good'),
          sum(discrete_meas$qualifier == 'Fair'), 
          sum(discrete_meas$qualifier == 'Poor'))) %>% 
  cbind(c(sum(discrete_meas$control_type == 'Clear Channel'),
          sum(discrete_meas$control_type == 'Light Vegetation'),
          sum(discrete_meas$control_type == 'Light Debris'),
          sum(discrete_meas$control_type == 'Shore Ice'),
          sum(discrete_meas$control_type == 'Moderate Debris'),
          sum(discrete_meas$control_type == 'Ice Cover'),
          nrow(discrete_meas))) %>% 
  kable( digits = 4, 
         caption = paste0('Table ', tab_no, '. Frequency of flow measurements by stream channel control type and flow measurement qualifiers at ',
         site_info$site_no, ' ', site_info$station_nm, 'from ', date_beg_model, ' to ', date_end_model),
         col.names = c( 'Excellent', 'Good', 'Fair', 'Poor', 'Total') ) %>%
  kable_styling() %>% 
  add_header_above(c("Type of Control" = 1, "Discrete Flow Measurement Qualifier" = 4, ' '), align = 'center') %>% 
  footnote('Discrete measurements with the type of control "Ice Cover" were subsequently deleted from the analysis because they did not reflect persistent changes in the stage-flow relation. Also, check measurements, which are repeated flow measurements on the same day to confirm the accuracy of the initially measured flow, were made on two days during the period of analysis. An average flow was computed from the two measurements to represent the discrete flow on that day. Thus, the total number of discrete flow measurements was reduced to 188.')

# Remove ice cover measurements
discrete_meas <- discrete_meas %>% 
  dplyr::filter( control_type != 'Ice Cover' ) 
  # dplyr::select( 'measurement_nu', 'measurement_dt',
  #                'stage_discrete_mea', 'flow_discrete_mea', 'qualifier',  'control_type',
  #                'se_pct', 'se_flow' )


# Assign first 471 - 478 discrete measurements to rating 17
ndx_rat <- which(discrete_meas$measurement_nu < 479)
discrete_meas$rating_no[ ndx_rat ] <- 17
# Add measurements 479 and 480 to the set used for 18 because of the highest
ndx_rat <- which(discrete_meas$measurement_nu > 478 & discrete_meas$measurement_nu < 481)
discrete_meas$rating_no[ ndx_rat ] <- 18

# Compute time between consequtive flow measurements
df <- data.frame( 'days_between_measurements' = diff(discrete_meas$Date, 1))



```

***

> Time Between Discrete Flow Measurements  

The frequency of discrete flow measurements is an important factor in detecting persistent changes in the stage-flow rating and in determining the costs of field operations. During the period of analysis at Pere Marquette River, the mean time between discrete measurements was `r format(mean(df$days_between_measurements), digits = 1)`.  There are no extended gaps in discrete flow measurements.  The four measurements at zero time between measurements are from check measurements made to ensure the initally measured flow. 


```{r days_between_meas, messages = FALSE, warning = FALSE } 

# Increment figure counter
fig_no <- fig_no + 1

# Plot the probability density of days between measurements
df %>% 
  ggplot( aes( x = as.numeric(days_between_measurements) )) + 
  geom_histogram( fill = 'black', alpha = 0.50, binwidth = 5 ) +
  theme_bw() +
  geom_vline( xintercept = mean(df$days_between_measurements), color = 'blue', linetype = 'dashed' ) +
  xlab( 'Time between flow measurements, in days' ) +
  ylab( 'Probability Density') +
  ggtitle(paste('Figure', fig_no, '. Distribution of days between consecutive flow measurements at \nstreamgage 04122500 Pere River at Scottville, MI from', date_beg_model, 'to', date_end_model),
          subtitle = 'Bin widths include 5 days.') +
  annotate('text', x = 52, y = 25.5, label = 'Mean = 58 days', hjust = -0.25, vjust = 0, color = 'blue')

```

***

> Unit stage values per day

Until recently, unit and daily stage data were considered ancillary variables to the computation of unit and daily flows. Thus, the availability of historical unit and daily stage data is more limited than daily flow data. In particular, limitations in the availability of historical stage data constrained the start of the period of analysis utilized in this paper.  In addition, the number of unit values recorded per day varied during the period of analysis (fig. `r fig_no + 1`). From about October 1, 1989 to May 4, 1994, 24 values per day was common; from May 5, 1994 to February 21, 1997, 288 values per day (5-minute interval data) were common; and after February 21, 1997, 96 values per day (15-minute interval data) were common. Other small irregularities in unit values occurred during the period of analysis. In developing the approach, it would be helpful if a simple method were available to adapt to changing frequencies of unit value data. 

```{r unit_daily_stage, warning = FALSE }

if( file.exists('stage_unit.RData')){
  # print('Loading published unit stage data from RData file.')
  load('stage_unit.RData')
} else {
  print('Retrieving daily flow from Aquarius Retrieval')
  stage_unit <- read.csv('Data/04122500/stage/Gage_height.ft@04122500.EntireRecorda.csv', 
                         header = TRUE, sep = ',', stringsAsFactors = FALSE, comment = '#')
  # Standardize field names
  stage_unit     <- renameNWISColumns(stage_unit)
  stage_unit$dateTime <- as_datetime(stage_unit$Timestamp..UTC.05.00, tz = 'America/New_York')
  colnames(stage_unit)[3] <- 'stage'
  save(stage_unit, file = 'stage_unit.RData')
}

# Summarize unit stage to daily mean stage
stage_day <- stage_unit %>% 
  group_by(day = floor_date( dateTime, 'day')) %>% 
  summarize( stage_daily_mean  = mean( stage, na.rm = TRUE ),
             n         = n()) 

fig_no <- fig_no + 1
suppressMessages(stage_day %>% 
  # Remove row with missing values
  ggplot( aes( x = as.Date(day), y = n)) +
  geom_point( color = 'black', alpha = 0.5, shape = 16, size = 1 ) +
  geom_vline( xintercept = as.Date( c('1994-05-04', '1997-02-21'), format('%Y-%m-%d')), 
              linetype = 'dotted', color = 'black' ) +
  theme_bw() +
  xlab('Date') +
  ylab('Number of unit values per day') +
  ggtitle( paste0('Figure ', fig_no, '. Number of unit stage values per day at 04122500 Pere Marquette River \nat Scottville, Mich. during the period of analysis' )))

stage_day <- stage_day %>% 
  ungroup() %>% 
  arrange( day ) %>% 
  mutate( delta_day = as.numeric(day - lag( day ), units = 'days' ),
          day_date  = as.Date( day )) 

```
` 

```{r }

# Measured flows and daily stages for the interval when rating 19 was active
#  Note: Discrete measurement timing is approximated to days

df_stage_flow_allDays <- data.frame(day = seq.Date(from = date_beg_model, to = date_end_model, by = '1 day'))

# calculate_mode function from https://exploratory.io/note/kanaugust/1701090969905358
#   Used in subsequent summarize
calculate_mode <- function(x) {
  uniqx <- unique(na.omit(x))
  uniqx[which.max(tabulate(match(x, uniqx)))]
}

# Take the mean flow and stage for days of multiple flow measurements
df_discrete_meas <- discrete_meas %>% 
  group_by( Date ) %>% 
  summarize( flow_discrete_mea       = mean( flow_discrete_mea ) ,
             stage_discrete_mea      = mean( stage_discrete_mea     ) ,
             n                       = n(),
             qualifier               = calculate_mode(qualifier),
             control_type            = calculate_mode(control_type),
             se_pct                  = mean(se_pct),
             se_flow                 = mean(se_flow))

# Integrate daily flow and discrete measurements 
df_stage_flow_allDays <-df_stage_flow_allDays %>% 
  left_join( df_discrete_meas[, c('Date', 'flow_discrete_mea', 'stage_discrete_mea',
                                  'qualifier', 'control_type', 'se_pct', 'se_flow')], 
             by = c( 'day'  = 'Date')) %>% 
  left_join( stage_day[, c('day_date', 'stage_daily_mean')], by = c( 'day' = 'day_date' )) %>% 
  filter( day >= date_beg_model & day <= date_end_model )


```



### Historical Rating Curves


```{r read_ratings, results = 'hide' }

# Rating curve at the selected site.  The 'exsa' parameter is an extended table
#   provide detailed info on stage discharge

rating18 <- read.csv('Data/04122500/04122500_rating_18.0.txt', header = FALSE, sep = '\t') 
colnames( rating18 ) <- c('rated_flow', 'stage')

# Use rating18 as the initial rating
rating_init   <- rating18
#   Rename column rated18_flow to rated_flow_init
colnames(rating_init)[1] <- 'rated_flow_init'

rating19 <- read.csv('Data/04122500/04122500_rating_19.0.txt', header = FALSE, sep = '\t') 
colnames( rating19 ) <- c('rated_flow', 'stage')

rating20 <- read.csv('Data/04122500/04122500_rating_20.0.txt', header = FALSE, sep = '\t') 
colnames( rating20 ) <- c('rated_flow', 'stage')

rating21 <- read.csv('Data/04122500/04122500_rating_21.0.txt', header = FALSE, sep = '\t') 
colnames( rating21 ) <- c('rated_flow', 'stage')


suppressMessages(rating_wide <- rating18   %>%  
  full_join(   rating19 ) %>% 
  full_join(   rating20 ) %>% 
  full_join(   rating21 ) %>% 
  dplyr::select(stage, everything() ))
  
rating_long <- rbind(rating18, rbind(rating19, rbind(rating20, rating21)))
rating_long$rating_no <- c(rep(18, nrow(rating18)), rep(19, nrow(rating19)), 
                           rep(20, nrow(rating20)), rep(21, nrow(rating21)))

```



> Estimate missing stage based on daily mean flow and rating curve 18

The time series of computed daily mean stage values (fig. `r fig_no + 1`) shows a seasonal variation within which annual minimums appear to have a downward trend. In particular, during the early 1990s, minimum annual stage values were generally greater than 1.6 ft above gage datum, while during the 2010s, minimum annual stage values were generally 0.6 ft lower. This pattern is consistent with the possible trend in the discrete measurements of stage data discussed earlier. The datum of the gage is 597.66 ft above National Geodetic Vertical Datum of 1929 (NGVD29). Based on visual inspection of the time series, the variability of stage values may be increasing with time.

Most of the missing daily mean stage values occurred during the period Rating 18 was in effect (fig. `r fig_no + 1`). Thus, daily mean flows were used with Rating 18 flow-stage relation to estimate all missing daily mean stage values. 


```{r missing_stage, warning = FALSE, fig.width = 8 }

df_stage_flow_allDays <- df_stage_flow_allDays %>% 
  left_join( flow_daily_pub[, c('Date', 'Flow')], by = c('day' = 'Date')) %>% 
  rename( 'flow_daily_pub'    = 'Flow',
          'Date'              = 'day')


# find indices of missing stage_mean_day values
ndx_ina <- which( is.na(df_stage_flow_allDays$stage_daily_mean ))

# find indices of bad daily stage from bad unit stage (1994-09-13 to 1994-09-19)
ndx_bad <- which( df_stage_flow_allDays$Date >= as.Date('1994-09-13') & 
                  df_stage_flow_allDays$Date <= as.Date('1994-09-19'))

ndx_miss <- sort(c(ndx_ina, ndx_bad))

# Find dates for missing data
dates_missing <- df_stage_flow_allDays$Date[ndx_miss]

# Estimate daily mean stage from published daily mean flow using rating 18, which was active when most of the missing values occurred)
stage_missing <- approx( x = rating18$rated_flow, y = rating18$stage, 
                         xout =df_stage_flow_allDays$flow_daily_pub[ndx_miss])

# Estimated missing daily stage on the basis of published daily mean flow 
df_stage_flow_allDays$stage_daily_mean[ ndx_miss ] <- stage_missing$y
df_stage_flow_allDays$stage_miss               <- 'Measured'
df_stage_flow_allDays$stage_miss[ ndx_miss ]   <- 'Estimated'

df_stage_flow_allDays <- addWaterYear(df_stage_flow_allDays)

min_Stage_group_waterYear <- df_stage_flow_allDays %>% 
  dplyr::group_by( waterYear ) %>% 
  summarize(minStage = min(stage_daily_mean),
            minDate  = Date[ which.min(stage_daily_mean) ])


fig_no <- fig_no + 1
df_stage_flow_allDays[, ] %>% 
  ggplot( aes( x = Date, y = stage_daily_mean, color = stage_miss )) +
  geom_point( size = 1 ) +
  geom_vline( xintercept = as.Date( c('1994-10-25', '2011-03-01', '2013-10-01')) , color = 'blue') + 
  # geom_line( data = min_Stage_group_waterYear, aes( x = minDate, y = minStage ), color = 'red') +
  theme_bw() +
  xlab('Date') +
  theme( legend.position = 'bottom') +
  scale_y_sqrt( breaks = c(1,2,4,6), limits = c(1,7)) +
  annotate('text', x = as.Date('1990-01-01'), y = 7, label = 'Rating 18', hjust = 'left') +
  annotate('text', x = as.Date('2003-01-01'), y = 7, label = 'Rating 19', hjust = 'left') +
  annotate('text', x = as.Date('2012-01-01'), y = 7, label = '20',        hjust = 'left') +
  annotate('text', x = as.Date('2016-01-01'), y = 7, label = 'Rating 21', hjust = 'left') +
  ylab('Stage, in feet above gage datum of 597.66 ft NAVD29') +
  ggtitle( paste0('Figure ', fig_no, '. Time Series of daily mean stage computed from unit values at \n04122500 Pere Marquette River at Scottville, Mich., during the period of analysis'),
           subtitle = paste0('Vertical blue lines indicate ', length(ndx_miss), ' days of missing stage that were estimated.'))
  
```


***

> Stage values used in modeling

Figure `r fig_no + 1 ` shows the close correspondence  between discrete stage on days of field measurements and daily mean stage computed from unit values.  Only one estimated stage occurred on a day of a discrete measurement. For the purpose of modeling flows and tracking the changes in the stage-flow relation, daily mean stages were replace with measured discrete stages on days of field measurement so that even the small discrepancy between discrete stage measurements and daily mean stage did not degrade modeling accuracy.  



```{r mean_discrete stge}

# 
# Substitute discrete stage and time of discrete flow measurement
#   Find indices of discrete stage-flow measurements
ndx_ian <- which( !is.na( df_stage_flow_allDays$stage_discrete_mea ) )


fig_no <- fig_no + 1
df_stage_flow_allDays[ ndx_ian, ] %>% 
  mutate( stage_miss = factor( stage_miss, levels = c('Measured', 'Estimated'))) %>% 
  ggplot( aes( x = stage_discrete_mea, y = stage_daily_mean ) ) +
  geom_point( aes( color = stage_miss, shape = stage_miss ), alpha = 1 ) +
  theme_bw() +
  scale_x_log10( breaks = seq(1, 6, by = 1) ) +
  scale_y_log10( breaks = seq(1, 6, by = 1) ) +
  geom_abline( intercept = 0, slope = 1, color = 'salmon') +
  xlab('Stage at time of discrete flow measurements, in feet above streamgage datum') +
  ylab('Daily mean stage on days of discrete measurements, in feet') +
  ggtitle(paste0('Figure ', fig_no, '. Relation between discrete stage and daily mean stage \non days of discrete flow measurements.'))


#   Find discrepancies between daily mean and discrete stage on days of discrete measurements
del_stage <- df_stage_flow_allDays$stage_discrete[ ndx_ian ] - df_stage_flow_allDays[ ndx_ian, 'stage_daily_mean' ]
#   Find index of largest (positive) discrepancy
ndx_day <- which( del_stage == max( del_stage) )

# Find indices of unit stage on day of max stage discrepancy 
ndx_unit_stage <- which( as.Date(stage_unit$dateTime) == df_stage_flow_allDays$Date[ ndx_ian][ndx_day] )
# Result is an empty vector.  There are no unit stages on this day.
ndx_stage <- which( as.Date(stage_unit$dateTime) == as.Date('1992-07-22') ) 

# Substitude discrete stage values for mean daily stage on days of discrete measurement
df_stage_flow_allDays$stage_daily_mean[ ndx_ian ] <- df_stage_flow_allDays$stage_discrete_mea[ ndx_ian ]


```





***

### Historical Stage-Flow Ratings 

Historical Stage-Flow Ratings were retrieved from USGS NWIS data bases using Aquarius Time-Series Software [@Aquarius2017]. The expanded form of the ratings 18-21 were exported in column format. The expanded rating indicate the relation between stage and flow at 0.01 ft increments of gage height. Table 2 showing the starting and ending dates for each of the selected ratings and the number of discharge measurements in the corresponding interval when the ratings were applied. 


```{r rating_history, fig.width = 5 }

# Read rating history
rating_history <- read.csv( 'Data/04122500/rating_period_start_end.txt', sep = '\t', header = TRUE,
                            colClasses = c('numeric', 'Date', 'Date'), comment.char = '#')

# Increment table counter
tab_no <- tab_no + 1

# Print table showing the number of discrete measurements by rating period
suppressMessages(tmp <- discrete_meas %>% 
  # rename( 'rating_no' = 'effective_rating') %>% 
  group_by( rating_no ) %>% 
  summarize( no_meas = n()) %>% 
  left_join( rating_history ) %>% 
  dplyr::select(rating_no, start_date, end_date, no_meas))

tmp %>% 
  dplyr::filter( !is.na( rating_no )) %>% 
  kable( caption = paste0('Table ', tab_no, '. Number of discrete flow measurements during selected rating periods at streamgage 04122500 Pere Marquette River at Scottville, Mich. during the period of analysis '),
         col.names = c('Rating number', 'Starting date', 'Ending date', 'Number of discrete measurements')) %>% 
  kable_styling( full_width = F) %>% 
  column_spec(1, width = '5em' ) %>% 
  column_spec(2, width = '6em' ) %>% 
  column_spec(3, width = '6em' ) %>% 
  column_spec(4, width = '6em' ) %>% 
  add_header_above( c(' ', 'Rating Period' = 2, ' ') ) %>% 
  footnote('Discrete measurements that were obtained during the rating period were included in the number of count of discrete measurements for that rating. Discrete measurement 479 and 480, which occurred on Sept. 13, 1986, and Sept. 17, 1986, were included in rating period 18 because they were associated with the maximum stage measured at the streamgage.')




```

***

> Plot Historical Stage-Flow Ratings with corresponding discrete measurements


Figure `r fig_no + 1` shows historical rating curves 18-21 and corresponding discrete flow measurements during the selected rating periods.  A persistent shift to the right is evident in the plot indicating higher rates of flow are discharge at the same stage. The historical rating curves were developed manually.  


```{r plot_ratings_meas, fig.width = 9, fig.height = 8,}

# Increment fig_no
fig_no <- fig_no + 1

rating_long %>% 
  ggplot( aes( x = rated_flow, y = stage,  color = rating_no)) +
  geom_line( aes( color = factor(rating_no) ), size = 1.0, alpha = 0.50) +
  theme_bw() +
  geom_point( data = discrete_meas[ 9:nrow(discrete_meas), ], aes( x = flow_discrete_mea, y = stage_discrete_mea, 
                                         color = as.factor(rating_no))) +
  # geom_text(data = discrete_meas, aes( x = Date, y = stage_discrete_mea, label = measurement_nu ), hjust=0, vjust=0) +
  # facet_wrap( ~ rating_no) +
  theme( legend.position = 'bottom' ) +
  scale_x_continuous( breaks = seq(1000, 7000, by = 1000)) +
  scale_y_continuous( breaks = seq(   1,    8, by =    1)) +
  xlab( 'Flow, in cubic feet per second' ) +
  ylab( 'Stage, in feet above gage datum' ) +
  annotate('text', x = c(1300, 1950, 3200, 5200), y = c( 4.0, 4.7, 6.0, 7.0), 
           label = c('18', '19', '20', '21')) +
  labs( title =  paste0('Figure ', fig_no, '. Discrete flow measurements for selected rating curves at ', site_info$site_no, '\n', site_info$station_nm ) )

  
```

## Methods for Adapting Stage-Flow Rating to Sequential Discrete Flow Measurements


### Cubic Regression Splines

Cubic regression splines were developed within the framework of a generalized additive model (GAM) to flexibly represent the monotonically increasing relation between stage and flow. A cubic regression spline, according to [@Wood2017], is a smooth curve created from a set of third-degree polynomials, which are joined at knots where first and second order derivatives are continuous.  The GAM package 'mgcv' [@mgcv2019] provided parameter estimation, and statistics on model fit for the cubic regression splines within the R [@cite_R] programming environment.   

A linear form of the cubic regression spline was used by specifying a set of basis functions such that $b_j(x)$ is the $j^{th}$ function of the set.  A second-order polynomial basis can be represented with basis functions $b_0(x) = 1$, $b_1(x) = x$, $b_2(x) = x^2$ so that $y_i = \beta_0 + x_i \beta_1 + x_i^2 \beta_2 + e_i.$ Thus, a second order polynomial has three basis functions ${1, \ x, \ x^2}$ and three corresponding parameters ${\beta_0, \ \beta_1, \ \beta_2}$.  According to [@Wood2017], a polynomial basis is a useful approximation when interest in on a specific point, however, splines are preferred when interest in the response (flow) is over the entire range of x (stage). 

Splines used in this analysis were specified as having 5 knots, which corresponds to 5 basis functions and 5 estimated parameters. The first basis function, $b_0(x)$, is a vector of 1's describing an intercept term. Five knots were thought to be sufficiently flexible for the stage-flow relations at the selected streamgage. The five knots are automatically placed at the minimum and maximum of the stage range, and at the $25^{th}, 50^{th}, and \space 75^{th}$ percentiles of the stage values. For each stage value input, 5 basis values are returned.  The basis themselves are not affected by the response variable (flow).  

The linear basis function used in the analysis is written below:

$$ flow(stage_i) = {\displaystyle \sum_{n=0}^{4} b_j(stage_i) \space \beta_j + \varepsilon_i} $$

> Specification of a Generalized Additive Model for Rating Curve 18  

** Model specification in the R statistical computing environment: **    
gam18_rating   <- gam( rated_flow        ~ s(stage,             bs = 'cr', k = knts),  
                       data = rating18, family = Gamma(link = log))  



** data = rated_flow**   the response variable of flow, in cubic feet per second  

| Model Syntax            |  Explanation                                                                                              |  
|:------------------------|:----------------------------------------------------------------------------------------------------------|  
|: gam18_rating         |: The gam object produced as model output                                                                  |  
|: gam( ... )         |: Function from the R package 'mgcv' used to specify the gam model                                         |     
|:  ~                 |: Relational operator                                                                                      |  
|: s( ... )           |: Function used in definition of smooth terms within gam model formulae. The function does not evaluate    |  
|:                        |:   A (spline) smooth - it exists purely to help set up a model using spline based smooths [@mgcv2019],    |  
|: bs = 'cr'        |: bs is an input argument that accepts a code for the type of smoothing bases. The 'cr' indicates a Cubic  |  
|:                        |:    Regression Spline [@mgcv2019]                                                                         |  
|: k = knts         |: k is an input argument for the number of knots, which IS variable knts that is assigned a value of 5     |    
|: data = rating18  |: The dataframe used in the analysis, with partial listing below this table                                |  
|: family = Gamma(link = log) |: This is a family object specifying the distribution and link (transformation) to use in fitting the GAM. Based 
|:  |: on the distribution of daily mean flows, the Gamma distribution family following a natural logarithm transformation is a reasonable choice.| 
|:   |:

```{r echo = FALSE}
print('First 6 rows of data in data.frame rating18')
print( head( rating18 ))
```

 

### State-Space Model

A state-space model consists of two equations: a process or state equation, and a measurement equation. The left-hand-side state vector of length *n*, $\vec{x}_{t}$ contains unobservable states, which here correspond to the five parameters of the cubic spline regression at time *t.*  The states are updated from observations in the measurement vector (scalar in this case) $\vec{y}_{t}$. These updates are based on the Kalman gain, computed using variance of the state vector, *V* and the variance of the flow measurement *R* [@marss2018]. The two equations are solved alternately, starting with the state equation, and iteratively through time indexed by $t.$  The state-space model is written below:


$$
\begin{align}
\text{ State Equation:  } \vec{x}_{t}  &= \: \vec{x}_{t-1} + \vec{w}_{t} && \text{where } \vec{w}_{t} \text{~} MVN(0, Q) \text{,}\\
 \text{Measurement Equation:  } y_{t} &=     \vec{Z}_{t}{'} \text{ } \vec{x}_{t} + v_t && \text{where } v_t \text{~} N(0, R_{t = t'}) \text{,}\\
 \text{}
\end{align}
$$

where $MVN(\vec0,Q)$ indicates a *Multi-Variate Normal* vector with mean $\vec{0}_{n}$, and process variance $Q$.  
Similarly, $N(0, R_{t})$ indicates a *Normally* distributed error with mean $0$ and measurement variance $R$.  

> State Vector $\vec{x}_t

The vector $\vec{x}_t$ is the time series of state vectors, where each vector has a length of 5, corresponding to the number of parameters specified in the cubic spline regression. The state equation models the transition of the state vector from time $t-1$ to $t$ as a random walk. This implies that the state-vector is invariant between discrete flow measurements, when the measurement update is skipped because of missing values (NA) in the flow series $Y_t$. The variance of the state vector before the measurement update $V_t^-$, however, is incremented by $Q$ at each time step in the state equation, while  $V_t^+$ is state variance updated (reduced) after discrete measurements in the measurement equation. 

Several variations of $\vec{x}_t$ are generated by function *MARSSkf* as Kalman filter and smoother output [@marss2018].  These include *xtT*, $x_{t|T}$, which is the smoothed or offline estimate of the state vector at time *t* given all the time steps in the series *T*; *xtt*, $x_{t|t}$, is the filter estimate of the state vector at time *t* given information up to and including time *t*; and *xtt1*, $x_{t|t-1}$ is the predicted state at time *t* given information up to time *t-1*. The state vector was initialized at t = 0 $\vec{x}_{0}$ to the corresponding parameters estimated in a selected GAM model. Like the state vector, the covariance matrix of the state vector has three estimates in MARSS, $V_{t|T}$, $V_{t|t}$, and $V_{t|t-1}$ depending upon whether smoothed, filtered, or prediction variance are of interesst. 

> Measurement Time Series $y_t$

The scalar measurement time series $y_t$ has a length equal to the number of days in the period of analysis. Although the series represents flow at daily increments of time, $y_t$ contains the natural logarithms of discrete flows, in cubic feet per second, on days of field measurements. The natural logarithm transformation also makes the parameterization of the state-space model consistent with the associated GAM model.  Similarly, the stage on days of discrete measurements is the corresponding discrete stage measurement, rather than the daily mean of unit stage values for that day. The natural logarithm is used in $y_t$ to help ensure that $v_t \sim N(0, R)$ and that (small) variances of flow measurements, which are based on percentages of measured flows, are compatible flows modelled in natural logarithm units. On days when no flow measurement occurred, the elements in the vector contain an *NA* value, so that no measurement update occurs to the state vector or the state covariance matrix. In this analysis, $y_t$ vector was composed of mostly (97 percent) *not available* indicators. 

> Process Covariance Matrix $Q$

The time-varying process covariance matrix $Q_t$ increases the uncertainty of the state vector $\vec{x}_t$, which is characterized in the time-varying state covariance matrix $V_t$  [@simon2006]. The process variance reflects the forecast uncertainty in the transition from time $t-1$ to $t$. After the state equation update, the state covariance matrix increases from $V_{t-1}^+$ to $V_{t}_1$. If there is a discrete flow measurement at time $t$, the state covariance matrix is decreased by the information in the measurement to $V_{t}^+$. Otherwise, there is no reduction in the state covariance, which continues to increase as a function of $Q$ until a discrete flow measurement occurs. The initial state covariance matrix can be initialized at time zero, $V_0^+$.  Generally, the initial value is set sufficiently large so that the state vector is appropriately sensitive to information in discrete measurements after time zero. 

equation, and the reduction in variance in measurement equation $V_t^+$. If the process variance is estimated by use of daily time steps, the resulting variance reflects the day to day increase in the uncertainty of the spline parameters. If subsequently, the simulation time increment were to change to 15-minute intervals, for example, the daily process variance divided by 96 is suggested as a suitable estimate of the unit process variance. In this analysis, *Q* is unknown, but estimated as diagonal matrix in various forms. 


Symmetrical Process Covariance Matrix

$$ Q (symmetrical) =  \begin{bmatrix} q_{11} & q_{12} & q_{13} & q_{14} & q_{15} \\ q_{12} & q_{22} & q_{32} & q_{42} & q_{52} \\ q_{13} & q_{23} & q_{33} & q_{43} & q_{53} \\ q_{14} & q_{24} & q_{34} & q_{44} & q_{54} \\ q_{15} & q_{25} & q_{35} & q_{45} & q_{55}\\ \end{bmatrix}$$






$$ Q_1 (diagonal \ and \ equal) =  \begin{bmatrix} q & 0 & 0 & 0 & 0 \\ 0 & q & 0 & 0 & 0 \\ 0 & 0 & q & 0 & 0 \\ 0 & 0 & 0 & q & 0 \\ 0 & 0 & 0 & 0 & q\\ \end{bmatrix}$$
$$ Q_2 (diagonal \ and \ unequal) =  \begin{bmatrix} q_1 & 0 & 0 & 0 & 0 \\ 0 & q_2 & 0 & 0 & 0 \\ 0 & 0 & q_3 & 0 & 0 \\ 0 & 0 & 0 & q_4 & 0 \\ 0 & 0 & 0 & 0 & q_5\\ \end{bmatrix}$$
$$ Q_3 (covar \ multiplicative) =  \begin{bmatrix} '0.1166336*q' & 0 & 0 & 0 & 0 \\ 0 & '1.3179593*q' & 0 & 0 & 0 \\ 0 & 0 & '0.6810473*q' & 0 & 0 \\ 0 & 0 & 0 & '1.0737010*q' & 0 \\
0 & 0 & 0 & 0 & '1.8106588*q'\\ \end{bmatrix}$$

> Measurement Variance Time Series $R_t$

The measurement variance *R* reflects the uncertainties in discrete measurements of flow.  Hydrographers provide qualifiers to describe the precision of discrete flow measurements as: *Excellent,* for measurements thought to be within 2-percent of the actual flow; *Good* for measurements within 5-percent; 'Fair' for measurements within 8-percent, and *Poor* for measurements that may be >8 percent from the actual flow.  (https://help.waterdata.usgs.gov/codes-and-parameters/discharge-measurement-quality-code). In this analysis, these uncertainties were interpreted as standard errors, in percent, for *Excellent, Good, Fair,* and *Poor* measurement qualifiers as 2, 5, 8 and 12, respectively. In addition, to using the hydrographer estimates directly, the measurement uncertainties were estimated from the data. 


$$ R (constant) = R_t = [r] for \ all \ t, \ where \ r \ is \ estimated. $$
 $$ R_t(qualifier): Excellent \ R_t = [0.02^2]; \ Good \ R_t = [0.05^2]; \ Fair \ R_t = [0.08^2]; Poor \ R_t = [0.12^2]  $$
$$ R = mean(R_t(qualifier)) $$

${Z}_{t}$ is the set of $1\times 5$ time series of vectors that form the design matrix. This matrix contains elements of the basis vectors computed from stage data on all days during the period of analysis. The stage data was computed as daily means of unit value stage data, except on days of discrete measurements. On days of discrete measurements, the discrete stage measurement themselves were substituted for the daily mean stage to more precisely correspond to the discrete flow measurements.     

All parameters estimated in various forms of the $Q$ matrix and $R$ times series vector were computed using the MARSS software and supporting functions [@marss2018].

## Results of Cubic Spline Regression and State-Space Modeling

### Cubic Regression Splines
 
Genearlized Additive Modeling software [@mgcv2019] was used to develop cubic spline regression models describing the relations between stage and flow on Pere Marquette River at Scottville, Mich. for rating periods 18, 19, 20, and 21.  Cubic splines were developed directly from expanded (0.01 ft increments of stage) ratings and from paired discrete measurements of stage and flow obtained during periods when the ratings were effective. Both methods produced close approximations of the existing rating curves, but there are advantages and disadvantages to both methods. 

from expanded (0.01 ft increments of stage) rating
and flow at 0.01 ft increments of stage for rating 18, 19, 20, and 21. The results of the modeling efforts are shown in table 3. For models estimated directly from the expanded stage-flow ratings, the cubic spline models were essentially identical to the expanded ratings, reflected by coefficients of determination (R^2^) values of approximately 1. Preliminary plots showed no discerable differences between the rating table data and the cubic spline model, so the plot were not included.  Confidence intervals about the cubic spline, however, were too narrow to interpret the uncertainty in the cubic splines because the number of effective points were simply a function of the table increments.  

rating tables of stage and flow data, and from discrete measurements of stage and flow obtained during field visits.
Parameter estimates for the splines [$\beta_0$, $\beta_1$, $\beta_3$, and $\beta_4$] were fairly consistent among rating periods. 






```{r gammeasPair_estimate}


# Estimate GAM model for each rating periods
knts = 5; # knts is the number of knots 

# GAM based on rating table
gam18_rating   <- gam( rated_flow        ~ s(stage,             bs = 'cr', k = knts),
                       data = rating18, family = Gamma(link = log)) 

gam19_rating   <- gam( rated_flow        ~ s(stage,             bs = 'cr', k = knts),
                       data = rating19, family = Gamma(link = log))

gam20_rating   <- gam( rated_flow        ~ s(stage,             bs = 'cr', k = knts),
                       data = rating20, family = Gamma(link = log))

gam21_rating   <- gam( rated_flow        ~ s(stage,             bs = 'cr', k = knts),
                       data = rating21, family = Gamma(link = log))

# Summarize GAM of rating-based model
gam18_rating_sum  <- summary(gam18_rating)
gam19_rating_sum  <- summary(gam19_rating)
gam20_rating_sum  <- summary(gam20_rating)
gam21_rating_sum  <- summary(gam21_rating)


# GAM based on discrete stage-flow measurement data 
gam18_measPair <-  gam( flow_discrete_mea ~ s(stage_discrete_mea, bs = 'cr', k = knts), 
                      data = discrete_meas[ discrete_meas$rating == 18, ],
                      family = Gamma(link = log))

gam19_measPair <-  gam( flow_discrete_mea ~ s(stage_discrete_mea, bs = 'cr', k = knts), 
                      data = discrete_meas[ discrete_meas$rating == 19, ],
                      family = Gamma(link = log))

gam20_measPair  <- gam( flow_discrete_mea ~ s(stage_discrete_mea, bs = 'cr', k = knts),
                      data = discrete_meas[ discrete_meas$rating == 20, ],
                      family = Gamma(link = log))


gam21_measPair  <- gam( flow_discrete_mea ~ s(stage_discrete_mea, bs = 'cr', k = knts),
                      data = discrete_meas[ discrete_meas$rating == 21, ],
                      family = Gamma(link = log))


gamPOA_measPair <- gam( flow_discrete_mea ~ s(stage_discrete_mea, bs = 'cr', k = knts),
                      data = discrete_meas,
                      family = Gamma(link = log))

# Compute GAM summary statistics
gam18_measPair_sum  <- summary(gam18_measPair)
gam19_measPair_sum  <- summary(gam19_measPair)
gam20_measPair_sum  <- summary(gam20_measPair)
gam21_measPair_sum  <- summary(gam21_measPair)
gamPOA_measPair_sum <- summary(gamPOA_measPair)


gam_stats <- data.frame('Rating'          = c('18', '19', '20', '21', '18', '19', '20', '21', 'POA'),
                           'N'            = c(gam18_rating_sum$n, gam19_rating_sum$n , gam20_rating_sum$n,
                                              gam21_rating_sum$n, gam18_measPair_sum$n, gam19_measPair_sum$n , 
                                              gam20_measPair_sum$n, gam21_measPair_sum$n, gamPOA_measPair_sum$n),
                           'Residual.df'  =  format(c(gam18_rating_sum$residual.df, gam19_rating_sum$residual.df,
                                                        gam20_rating_sum$residual.df, gam21_rating_sum$residual.df,
                                                        gam18_measPair_sum$residual.df, gam19_measPair_sum$residual.df,
                                                        gam20_measPair_sum$residual.df, gam21_measPair_sum$residual.df,
                                                        gamPOA_measPair_sum$residual.df), digits = 4),
                           'Intercept'      = format(c(gam18_rating$coefficients[1], gam19_rating$coefficients[1],
                                                        gam20_rating$coefficients[1], gam21_rating$coefficients[1],
                                                        gam18_measPair$coefficients[1], gam19_measPair$coefficients[1],
                                                        gam20_measPair$coefficients[1], gam21_measPair$coefficients[1],
                                                        gamPOA_measPair$coefficients[1]), digits = 4),
                           'beta1'      = format(c(gam18_rating$coefficients[2], gam19_rating$coefficients[2],
                                                        gam20_rating$coefficients[2], gam21_rating$coefficients[2],
                                                        gam18_measPair$coefficients[2], gam19_measPair$coefficients[2],
                                                        gam20_measPair$coefficients[2], gam21_measPair$coefficients[2],
                                                        gamPOA_measPair$coefficients[2]), digits = 4),
                           'beta2'      = format(c(gam18_rating$coefficients[3], gam19_rating$coefficients[3],
                                                        gam20_rating$coefficients[3], gam21_rating$coefficients[3],
                                                        gam18_measPair$coefficients[3], gam19_measPair$coefficients[3],
                                                        gam20_measPair$coefficients[3], gam21_measPair$coefficients[3],
                                                        gamPOA_measPair$coefficients[3]), digits = 4),
                           'beta3'      = format(c(gam18_rating$coefficients[4], gam19_rating$coefficients[4],
                                                        gam20_rating$coefficients[4], gam21_rating$coefficients[4],
                                                        gam18_measPair$coefficients[4], gam19_measPair$coefficients[4],
                                                        gam20_measPair$coefficients[4], gam21_measPair$coefficients[4],
                                                        gamPOA_measPair$coefficients[4]), digits = 4),
                           'beta4'      = format(c(gam18_rating$coefficients[5], gam19_rating$coefficients[5],
                                                        gam20_rating$coefficients[5], gam21_rating$coefficients[5],
                                                        gam18_measPair$coefficients[5], gam19_measPair$coefficients[5],
                                                        gam20_measPair$coefficients[5], gam21_measPair$coefficients[5],
                                                        gamPOA_measPair$coefficients[5]), digits = 4),
                           'R2.adj'          = format(c(gam18_rating_sum$r.sq, gam19_rating_sum$r.sq,
                                                        gam20_rating_sum$r.sq, gam21_rating_sum$r.sq,
                                                        gam18_measPair_sum$r.sq, gam19_measPair_sum$r.sq,
                                                        gam20_measPair_sum$r.sq, gam21_measPair_sum$r.sq,
                                                        gamPOA_measPair_sum$r.sq), digits = 4))
# Increment tab_no
tab_no <- tab_no + 1

gam_stats %>% 
  kable( caption = paste('Table', tab_no, '. Summary statistics of five-knot cubic regression splines from Generalized Additive Models for selected rating periods 18 - 20 and the Period of Analysis (POA).' ), 
  col.names = c('Rating period', 'Number of data points', 'Residual degrees of freedom', "beta_0",
                "beta_1", 'beta_2', 'beta_3', "beta_4", 'R^2')) %>% 
  kable_styling() %>% 
  add_header_above( c(' ' = 3, 'Estimated Parameters' = 5, ' ' = 1 ) ) %>% 
  pack_rows('Based on stage-flow rating table', 1, 4) %>% 
  pack_rows('Based on discrete stage-flow measurement', 5, 9) %>% 
  column_spec(1, width = "5em") %>% 
  column_spec(2, width = "5em") %>% 
  column_spec(3, width = "8em")


```



```{r gam_measPair_predict}

# Predicted values are for all stages in rating 21, which has the widest range of the ratings. Note that the predicted values are back-transformed into cubic feet per second from estimation in natural logs. 

gam18_pred <- predict(gam18_measPair, newdata = data.frame( stage_discrete_mea = rating21$stage), se.fit = TRUE,
                       type = 'response')

gam19_pred <- predict(gam19_measPair, newdata = data.frame( stage_discrete_mea = rating21$stage), se.fit = TRUE,
                       type = 'response')

gam20_pred <- predict(gam20_measPair, newdata = data.frame( stage_discrete_mea = rating21$stage), se.fit = TRUE,
                       type = 'response')

gam21_pred <- predict(gam21_measPair, newdata = data.frame( stage_discrete_mea = rating21$stage), se.fit = TRUE,
                       type = 'response')

gamPOA_pred <- predict(gamPOA_measPair, newdata = data.frame( stage_discrete_mea = rating21$stage), se.fit = TRUE,
                       type = 'response')


# Compile and compute gam model results
gam18_df <- data.frame('stage'        = rating21$stage,
                        'gam18_flow'  = gam18_pred$fit, 
                        'gam18_se'    = gam18_pred$se.fit)
gam18_df$ci18_025 <- gam18_df$gam18_flow + qnorm(0.025) * gam18_df$gam18_se
gam18_df$ci18_975 <- gam18_df$gam18_flow + qnorm(0.975) * gam18_df$gam18_se

gam19_df <- data.frame('stage'        = rating21$stage,
                        'gam19_flow'  = gam19_pred$fit, 
                        'gam19_se'    = gam19_pred$se.fit)
gam19_df$ci19_025 <- gam19_df$gam19_flow + qnorm(0.025) * gam19_df$gam19_se
gam19_df$ci19_975 <- gam19_df$gam19_flow + qnorm(0.975) * gam19_df$gam19_se

gam20_df <- data.frame('stage'        = rating21$stage,
                        'gam20_flow'  = gam20_pred$fit, 
                        'gam20_se'    = gam20_pred$se.fit)
gam20_df$ci20_025 <- gam20_df$gam20_flow + qnorm(0.025) * gam20_df$gam20_se
gam20_df$ci20_975 <- gam20_df$gam20_flow + qnorm(0.975) * gam20_df$gam20_se

gam21_df <- data.frame('stage'        = rating21$stage,
                        'gam21_flow'  = gam21_pred$fit, 
                        'gam21_se'    = gam21_pred$se.fit)
gam21_df$ci21_025 <- gam21_df$gam21_flow + qnorm(0.025) * gam21_df$gam21_se
gam21_df$ci21_975 <- gam21_df$gam21_flow + qnorm(0.975) * gam21_df$gam21_se

rating18 <- rating18 %>% 
  rename( rated18_flow = rated_flow) 
rating19 <- rating19 %>% 
  rename( rated19_flow = rated_flow) 
rating20 <- rating20 %>% 
  rename( rated20_flow = rated_flow) 
rating21 <- rating21 %>% 
  rename( rated21_flow = rated_flow) 

gam_df <- gam18_df %>% 
  full_join(gam19_df, by = 'stage') %>% 
  full_join(gam20_df, by = 'stage') %>% 
  full_join(gam21_df, by = 'stage') %>% 
  full_join(rating18, by = 'stage') %>% 
  full_join(rating19, by = 'stage') %>% 
  full_join(rating20, by = 'stage') %>% 
  full_join(rating21, by = 'stage')

gam_long <- gam_df %>% 
  gather( key = 'flow_type', 'flow_cfs', -stage) %>% 
  dplyr::filter( !grepl('_se', flow_type) ) %>% 
  mutate( rating = case_when( grepl('18', flow_type) ~ 18,
                              grepl('19', flow_type) ~ 19,
                              grepl('20', flow_type) ~ 20,
                              grepl('21', flow_type) ~ 21))

ndx_025 <- which( grepl('025', gam_long$flow_type))
gam_long$flow_type[ndx_025] <- 'ci_025'
ndx_975 <- which( grepl('975', gam_long$flow_type))
gam_long$flow_type[ndx_975] <- 'ci_975'
ndx_rat <- which( grepl('rated', gam_long$flow_type ))
gam_long$flow_type[ndx_rat] <- 'rated_flow'
ndx_gam <- which( grepl('gam', gam_long$flow_type))
gam_long$flow_type[ndx_gam] <- 'gam_flow'
  

```



```{r plot_by_rating, fig.height = 10, fig.width = 10}

fig_no <- fig_no + 1
# Plot manual and gam estimates of each rating curve with corresponding discrete measurements
gam_long %>% 
  ggplot( aes( x = flow_cfs, y = stage, color = flow_type )) +
  geom_line( aes( size = flow_type ), alpha = 0.25 ) +
  facet_wrap( ~ rating, ncol = 2 ) +
  theme_bw() +
  scale_color_manual( values = c('forestgreen', 'forestgreen', 'navy', 'red') ) +
  scale_size_manual(  values= c(           1 ,            1,     1.25,  1.25) ) +
  scale_x_continuous( limits = c(0, 10000)) +
  theme( legend.position = 'bottom') +
  geom_point( data = discrete_meas, 
              aes( x = flow_discrete_mea, y = stage_discrete_mea, group = rating_no), color = 'blue') +
  xlab( 'Flow, in cubic feet per second') +
  ylab( 'Stage, in feet above gage datum') +
  ggtitle( label = paste0('Figure ', fig_no, '. Comparison of rating curves created manually with generalized additive models for selected rating periods'))
  
  

# Store GAM parameters in a matrix for comparison
gam_parameters <- matrix(NA, 5, 9)

gam_parameters[ ,1] <- gam18_measPair$coefficients
gam_parameters[ ,2] <- gam19_measPair$coefficients
gam_parameters[ ,3] <- gam20_measPair$coefficients
gam_parameters[ ,4] <- gam21_measPair$coefficients
gam_parameters[ ,5] <- gamPOA_measPair$coefficients

gam_parameters[ ,6] <- gam18_rating$coefficients
gam_parameters[ ,7] <- gam19_rating$coefficients
gam_parameters[ ,8] <- gam20_rating$coefficients
gam_parameters[ ,9] <- gam21_rating$coefficients



gam_parameters_df  <- as.data.frame( gam_parameters )

rownames( gam_parameters_df ) <- c('Knot 1', 'Knot 2', 'Knot 3', 'Knot 4', 'Knot 5')
colnames( gam_parameters_df ) <- c('18', '19', '20', '21' , 'POA',
                                   '18', '19', '20', '21')


tab_no <- tab_no + 1

# Show gam model parameters for 
gam_parameters_df %>% 
  kable( caption = paste0('Table ', tab_no, '. Generalized Additive Model parameters for each spline knot for selected rating period.' )) %>% 
  kable_styling() %>% 
  add_header_above( c(' ' = 1, 'Paired Measurement Data for Rating Period' = 5,
                      'Rating Curve for Rating Period' = 4)) %>% 
  footnote(general = 'POA: Period of Analysis')
  # pack_rows('Generalized Additive Models based on discrete measurement data pairs', 1, 5) %>% 
  # pack_rows('Generalized Additive Models estimated from stage-flow rating tables', 6, 9 )


```




***

> Basis Functions 

The basis functions developed using the GAM of the rating tables are plotted in figure `r fig_no + 1` along with the knot locations for each model and basis, excluding the (constant) intercept term.  The co-location of knots with peak magnitudes of the basis functions may result in parameter values associated with the knots being perturbed more when discrete measured stage-flow pairs are near a corresponding knot. 



```{r plot_basis_functions, fig.width = 11, fig.height = 10}

mm18_rating <- model.matrix(gam18_rating)
basis18_rat <- data.frame(rep('18', nrow(rating18)), rating18$stage, mm18_rating)
colnames(basis18_rat) <- c('Rating', 'Stage', 'Intercept', 'basis1', 'basis2', 'basis3', 'basis4')

mm19_rating <- model.matrix(gam19_rating)
basis19_rat <- data.frame(rep('19', nrow(rating19)), rating19$stage, mm19_rating)
colnames(basis19_rat) <- c('Rating', 'Stage', 'Intercept', 'basis1', 'basis2', 'basis3', 'basis4')

mm20_rating <- model.matrix(gam20_rating)
basis20_rat <- data.frame(rep('20', nrow(rating20)), rating20$stage, mm20_rating)
colnames(basis20_rat) <- c('Rating', 'Stage', 'Intercept', 'basis1', 'basis2', 'basis3', 'basis4')

mm21_rating <- model.matrix(gam21_rating)
basis21_rat <- data.frame(rep('21', nrow(rating21)), rating21$stage, mm21_rating)
colnames(basis21_rat) <- c('Rating', 'Stage', 'Intercept', 'basis1', 'basis2', 'basis3', 'basis4')

basis_rat <- rbind(basis18_rat, rbind(basis19_rat, rbind(basis20_rat, rbind(basis21_rat))))
colnames(basis_rat) <- c('Rating', 'Stage', 'Intercept', 'basis1', 'basis2', 'basis3', 'basis4')
                            
fig_no <- fig_no + 1

knot18 = data.frame( Stage = gam18_rating$smooth[[1]]$xp[2:5],
                                 Basis =  c( 'basis1', 'basis2', 'basis3', 'basis4'))
knot19 = data.frame( Stage = gam19_rating$smooth[[1]]$xp[2:5],
                                 Basis =  c( 'basis1', 'basis2', 'basis3', 'basis4'))
knot20 = data.frame( Stage = gam20_rating$smooth[[1]]$xp[2:5],
                                 Basis =  c( 'basis1', 'basis2', 'basis3', 'basis4'))
knot21 = data.frame( Stage = gam21_rating$smooth[[1]]$xp[2:5],
                                 Basis =  c( 'basis1', 'basis2', 'basis3', 'basis4'))

basis_rat %>% 
  gather( key = 'Basis', value = 'Computed', basis1, basis2, basis3, basis4) %>% 
  mutate( Basis = factor(Basis, levels = c( 'basis1', 'basis2', 'basis3', 'basis4'))) %>% 
  ggplot( aes( x = Stage, y = Computed, color = Rating)) +
  geom_line( aes( size = Rating, alpha = Rating ) ) +
  geom_vline(data = knot18,  aes( xintercept = Stage), color = 'orangered' ) +
  geom_vline(data = knot19,  aes( xintercept = Stage), color = 'yellow4' ) +
  geom_vline(data = knot20,  aes( xintercept = Stage), color = 'aquamarine3' ) +
  geom_vline(data = knot21,  aes( xintercept = Stage), color = 'purple' ) +
  scale_size_manual( values = c( 1, 1, 1, 1, 3 )  ) +
  scale_alpha_manual(values = c( 1, 1, 1, 1, 0.4) ) +
  facet_wrap(. ~ Basis, ncol = 2   ) +
  theme_bw() +
  theme( legend.position = 'bottom') +
  ggtitle( paste0('Figure ', fig_no, '. Basis functions of cubic splines approximating selected ratings as a function of stage computed on the basis of discrete flow measurements'),
           subtitle = 'Note: The basis0 equals 1 for all stages and was not included in the plots below' )                     


```









```{r R_variance_distribution, fig.height = 5}
# Set up data frame to characterize measurement variances

ndx_ian_Yt <- which(!is.na(df_stage_flow_allDays$flow_discrete_mea) )

  # Initialize measurement error vector with large variance for all days
  R  <- array(0, dim = c(1, 1, nrow(df_stage_flow_allDays)))
  # Measurement variance specification
  #   The standard error of flow as a percentage that varies with perceived accuracy of the measured flow
  R[1,1, ndx_ian_Yt] <- (df_stage_flow_allDays$se_pct[ndx_ian_Yt]/100)^2

data.frame( measurement_variance = R[1, 1, ndx_ian_Yt])  %>% 
  ggplot( aes( x = measurement_variance)) +
  geom_bar( aes(y = (..count..)/sum(..count..)), fill = 'darkgreen', alpha = 0.50 ) + 
  geom_text(aes(y = ((..count..)/sum(..count..)), label = scales::percent((..count..)/sum(..count..))), 
            stat = "count", vjust = -0.25) +
  scale_x_log10( limits = c(1e-3, 5e-2), breaks = c( 0.05^2, 0.08^2, 0.12^2, 0.03) ) +
  scale_y_continuous( labels = scales::percent, limits = c(0, 1) ) +
  geom_vline( xintercept = 300) +
  theme_bw() +
  xlab( 'Measurement variance based on qualifier of discrete flow measurement') +
  ylab( 'Probability density') +
  labs( title = paste0('Figure ', fig_no, '. Frequency density of streamflow measurement variance (R) series based on qualifiers'))

```


### Formatting data for a state space model


```{r set_marss_global_parameters}

# Yt is the measurement vector that contains directly measured flows (not daily means)
Yt <- matrix( log(df_stage_flow_allDays$flow_discrete_mea), nrow = 1)

# Indexes for measured flows in vector
# ndx_ian_Yt <- which( !is.na(Yt) )

per_NA <- format( length(which( is.na(Yt))) / length(Yt) * 100, digit = 3 )
print(paste('The measurement vector is', 
             per_NA,'percent NA values.'))


Zt <- t(predict( gam18_rating, newdata = data.frame( stage = df_stage_flow_allDays$stage_daily_mean),
                 type = 'lpmatrix') )

# Remove row names
rownames(Zt) <- NULL

# Populate the Zt array in the format that MARSS expects
Zt <- array(Zt, dim = c(1, knts, nrow(df_stage_flow_allDays)))

# A is an unused time-varying vector in the measurement equation
A  <- 'zero'

# Drift vector
U  <- 'zero'

# Initial state vector from GA model
X0 <- as.matrix( as.numeric(coefficients(gam18_rating)), 5, 1)

# Initial covariance of the state model
V0 <- gam18_rating$Vp * 1000


```
The R package MARSS [@MARSS2018] was used in developing the state-space model. The notation below follows the authors' notation. 

### State-Space Model

The state-space model [@MARSS2018]

$$
\begin{align}
    \text{State Equation:  }  x_{t} &= {I_n} \: x_{t-1} + w_t && \text{where } w_t \text{~} MVN(0, Q) \text{,}\\
    \text{Measurement Equation:  } y_{t} &= Z_t \text{ } x_{t} + v_t && \text{where } v_t \text{~} N(0, R_{t = t'}) \text{,}\\
    \text{}
\end{align}
$$

Following notation by [@MARSS2018], $y_t$ is the scalar measurement time series, which has a length equal to the number of days in the period of analysis. $y_t$ contains the natural logarithms of discrete flows on days of field measurements. Otherwise, the elements in the vector contain an *NA* value, indicating that a discrete flow measurement was *not available* on that day. In this analysis, $y_t$ contained `r per_NA` percent *not available* indicators. 

$\vec{x}_t$ is the time series of state vectors, where each vector has a length of 5, corresponding to the number of knots and subsequent parameters specified in the generalized additive model (GAM) cubic spline regression equation. Several variations of $\vec{x}_t$ are generated by function *MARSSkf* as Kalman filter and smoother output.  These include *xtT*, which is the smoothed or offline estimate of the state vector at time *t* given all the time steps in the series *T*; *xtt* is the filter estimate of the state vector at time *t* given information up to time *t*; and *xtt1* is the predicted state at time *t* given information up to time *t-1*. 

Q is a $5\times5$ diagonal process variance matrix, which increments the variance *V* of the state vector with each time step in the model.  In the state equation, the process variance indicates the incremental increase in variance of spline regression parameters associated with each day since a discrete measurement of flow occurred. The elements of Q are estimated in the Multivariate Autoregressive State Space (MARSS) model .  To simulate unit flows, the diagonal elements of the Q matrix would be divided by the number of unit values in a day.     

R  is an estimated constant measurement variance

Zt is the design matrix of   

Yt is sequential daily values of log_stage_round 


Note: A multiplicative form for the Q matrix can be specified as:
This is provided just as a reminder of the format for defining Q with a multiplicative constant.
Q <- matrix(list('qb', 0, 0, 0, 0, 0, 0, '10*qb'), 3, 3)



#### marss_fit_Q_ones_multiplier_R_one_multiplier

```{r marss_fit_Q_ones_multiplier_R_one_multiplier}

# 1
# Specify model
if( file.exists('marss_fit_Q_ones_multiplier_R_one_multiplier.rds') ){
  marss_fit_Q_ones_multiplier_R_one_multiplier <- read_rds('marss_fit_Q_ones_multiplier_R_one_multiplier.rds')
} else{
  Q = 'diagonal and equal'  # 'ones_multiplier'
  R = 'diagonal and equal'  # 'one_multiplier'
  model.gen=list(Z=Zt,A=A,R=R,U=U,Q=Q,x0=X0,V0=V0,tinitx=0)
  # Refit model: trace = 1 provides parameter etimates with each iteration with the corresponding log-likelihood
  marss_fit_Q_ones_multiplier_R_one_multiplier  <- MARSS(Yt, model = model.gen, control = list(maxit =  5000, trace = 1), 
                                                         method = 'BFGS' ) 
  write_rds(marss_fit_Q_ones_multiplier_R_one_multiplier, path = 'marss_fit_Q_ones_multiplier_R_one_multiplier.rds')
}


```

#### marss_fit_Q_unequal_scalars_R_one_multiplier

```{r marss_fit_Q_unequal_scalars_R_one_multiplier}
# 2
# Specify model
if( file.exists('marss_fit_Q_unequal_scalars_R_one_multiplier.rds') ){
  marss_fit_Q_unequal_scalars_R_one_multiplier <- read_rds('marss_fit_Q_unequal_scalars_R_one_multiplier.rds')
} else{
  Q = 'diagonal and unequal'  # 'unequal_scalars'
  R = 'diagonal and equal'    # 'one_multiplier'
  model.gen=list(Z=Zt,A=A,R=R,U=U,Q=Q,x0=X0,V0=V0,tinitx=0)
  # Refit model: trace = 1 provides parameter etimates with each iteration with the corresponding log-likelihood
  marss_fit_Q_unequal_scalars_R_one_multiplier  <- MARSS(Yt, model = model.gen, control = list(maxit =  5000, trace = 1), 
                 method = 'BFGS')
  write_rds(marss_fit_Q_unequal_scalars_R_one_multiplier, path = 'marss_fit_Q_unequal_scalars_R_one_multiplier.rds')
}

```


#### marss_fit_Q_ones_multiplier_R_mean_qualifier


```{r marss_fit_Q_ones_multiplier_R_mean_qualifier}
# 3

# Specify model
if( file.exists('marss_fit_Q_ones_multiplier_R_mean_qualifier.rds') ){
  marss_fit_Q_ones_multiplier_R_mean_qualifier <- read_rds('marss_fit_Q_ones_multiplier_R_mean_qualifier.rds')
} else{
  Q <- 'diagonal and equal'                                                    # 'ones_multiplier'
  # R is specified by qualifiers of discrete flow measurement              
  R  <- matrix(mean( (df_stage_flow_allDays$se_pct[ndx_ian_Yt]/100)^2),1,1)    # 'mean_qualifier'
  model.gen=list(Z=Zt,A=A,R=R,U=U,Q=Q,x0=X0,V0=V0,tinitx=0)
  # Refit model: trace = 1 provides parameter etimates with each iteration with the corresponding log-likelihood
  marss_fit_Q_ones_multiplier_R_mean_qualifier  <- MARSS(Yt, model = model.gen, control = list(maxit =  1000, trace = 1), 
                 method = 'BFGS')
  write_rds(marss_fit_Q_ones_multiplier_R_mean_qualifier, path = 'marss_fit_Q_ones_multiplier_R_mean_qualifier.rds')
}


```

#### marss_fit_Q_unconstrained_R_mean_qualifier


```{r marss_fit_Q_unconstrained_R_mean_qualifier}
# 4

# Specify model
if( file.exists('marss_fit_Q_ones_multiplier_R_mean_qualifier.rds') ){
  marss_fit_Q_unconstrained_R_mean_qualifier <- read_rds('marss_fit_Q_unconstrained_R_mean_qualifier.rds')
} else{
  Q <- 'unconstrained'                                                    # 'ones_multiplier'
  # R is specified by qualifiers of discrete flow measurement              
  R  <- matrix(mean( (df_stage_flow_allDays$se_pct[ndx_ian_Yt]/100)^2),1,1)    # 'mean_qualifier'
  model.gen=list(Z=Zt,A=A,R=R,U=U,Q=Q,x0=X0,V0=V0,tinitx=0)
  # Refit model: trace = 1 provides parameter etimates with each iteration with the corresponding log-likelihood
  marss_fit_Q_unconstrained_R_mean_qualifier  <- MARSS(Yt, model = model.gen, control = list(maxit =  1000, trace = 1), 
                 method = 'BFGS')
  write_rds(marss_fit_Q_unconstrained_R_mean_qualifier, path = 'marss_fit_Q_unconstrained_R_mean_qualifier.rds')
}


```

#### marss_fit_Q_equalvarcov_R_mean_qualifier


```{r marss_fit_Q_equalvarcov_R_mean_qualifier}
# 5

# Specify model
if( file.exists('marss_fit_Q_equalvarcov_R_mean_qualifier.rds') ){
  marss_fit_Q_equalvarcov_R_mean_qualifier <- read_rds('marss_fit_Q_equalvarcov_R_mean_qualifier.rds')
} else{
  Q <- 'equalvarcov'                                                    # 'ones_multiplier'
  # R is specified by qualifiers of discrete flow measurement              
  R  <- matrix(mean( (df_stage_flow_allDays$se_pct[ndx_ian_Yt]/100)^2),1,1)    # 'mean_qualifier'
  model.gen=list(Z=Zt,A=A,R=R,U=U,Q=Q,x0=X0,V0=V0,tinitx=0)
  # Refit model: trace = 1 provides parameter etimates with each iteration with the corresponding log-likelihood
  marss_fit_Q_equalvarcov_R_mean_qualifier  <- MARSS(Yt, model = model.gen, control = list(maxit =  1000, trace = 1), 
                 method = 'kem', inits = coef(marss_fit_Q_equalvarcov_R_mean_qualifier))
  write_rds(marss_fit_Q_equalvarcov_R_mean_qualifier, path = 'marss_fit_Q_equalvarcov_R_mean_qualifier.rds')
}


```




#### marss_fit_Q_covar_multiplier_R_one_multiplier

```{r marss_fit_Q_covar_multiplier_R_one_multiplier}

# 6

# Standardize the diagonal of gam18_rating for estimation
#  These standardized variance terms average 1 and form the diagonal of Q
# gam18_diag_std <- diag(gam18_rating$Vp)/mean(diag(gam18_rating$Vp))

if( file.exists( 'marss_fit_Q_covar_multiplier_R_one_multiplier.rds' )){
  marss_fit_Q_covar_multiplier_R_one_multiplier <- read_rds( 'marss_fit_Q_covar_multiplier_R_one_multiplier.rds' )
} else {
  # State variance matrix is based on the gam18_measPair$Vp
  Q <- matrix(list('0.1166336*q',0,0,0,0,                 # 'covar_multiplier'
                   0,'1.3179593*q',0,0,0  ,
                   0,0,'0.6810473*q',0,0,
                   0,0,0,'1.0737010*q',0,
                   0,0,0,0,'1.8106588*q'), 5, 5)
  R  <- 'diagonal and equal'                              # 'one_multiplier'
  
  # Specify model
  model.gen=list(Z=Zt,A=A,R=R,U=U,Q=Q,x0=X0,V0=V0,tinitx=0)
  
  # Refit model: trace = 1 provides parameter etimates with each iteration with the corresponding log-likelihood
  marss_fit_Q_covar_multiplier_R_one_multiplier  <- MARSS(Yt, model = model.gen, control = list(maxit =  1000, trace = 1), 
                                                            method = 'BFGS')
  
  write_rds(marss_fit_Q_covar_multiplier_R_one_multiplier, path = 'marss_fit_Q_covar_multiplier_R_one_multiplier.rds')
}
```


#### marss_fit_Q_covar_multiplier_R_mean_qualifier

```{r marss_fit_Q_covar_multiplier_R_mean_qualifier}

# 7

# Standardize the diagonal of gam18_rating for estimation
#  These standardized variance terms average 1 and form the diagonal of Q
# gam18_diag_std <- diag(gam18_rating$Vp)/mean(diag(gam18_rating$Vp))

if( file.exists( 'marss_fit_Q_covar_multiplier_R_mean_qualifier.rds' )){
  marss_fit_Q_covar_multiplier_R_mean_qualifier <- read_rds( 'marss_fit_Q_covar_multiplier_R_mean_qualifier.rds' )
} else {
  # State variance matrix is based on the gam18_measPair$Vp
  Q <- matrix(list('0.1166336*q',0,0,0,0,                                     # 'covar_multiplier'
                   0,'1.3179593*q',0,0,0  ,
                   0,0,'0.6810473*q',0,0,
                   0,0,0,'1.0737010*q',0,
                   0,0,0,0,'1.8106588*q'), 5, 5)
  
  # R is specified by qualifiers of discrete flow measurement
  R  <- matrix(mean( (df_stage_flow_allDays$se_pct[ndx_ian_Yt]/100)^2),1,1)   # 'mean_qualifier'
  
  # Specify model
  model.gen=list(Z=Zt,A=A,R=R,U=U,Q=Q,x0=X0,V0=V0,tinitx=0)
  
  # Refit model: trace = 1 provides parameter etimates with each iteration with the corresponding log-likelihood
  marss_fit_Q_covar_multiplier_R_mean_qualifier  <- MARSS(Yt, model = model.gen, control = list(maxit =  1000, trace = 1), 
                                                            method = 'BFGS')
  
  write_rds(marss_fit_Q_covar_multiplier_R_mean_qualifier, path = 'marss_fit_Q_covar_multiplier_R_mean_qualifier.rds')
}
```

#### marss_fit_Q_ones_multiplier_R_series_qualifier

```{r marss_fit_Q_ones_multiplier_R_series_qualifier}

# 8

if ( file.exists( 'marss_fit_Q_ones_multiplier_R_series_qualifier.rds' )){
  marss_fit_Q_ones_multiplier_R_series_qualifier <- read_rds( 'marss_fit_Q_ones_multiplier_R_series_qualifier.rds' )
} else {
  Q = 'diagonal and equal'                                                      # 'ones_multiplier'
  
  # Initialize measurement error vector with large variance for all days        # 'series_qualifier'
  R  <- array(0, dim = c(1, 1, nrow(df_stage_flow_allDays)))
  # Measurement variance specification
  #   The standard error of flow as a percentage that varies with perceived accuracy of the measured flow
  R[1,1, ndx_ian_Yt] <- (df_stage_flow_allDays$se_pct[ndx_ian_Yt]/100)^2
  
  # Specify model
  model.gen=list(Z=Zt,A=A,R=R,U=U,Q=Q,x0=X0,V0=V0,tinitx=0)
  
  # Refit model: trace = 1 provides parameter etimates with each iteration with the corresponding log-likelihood
  marss_fit_Q_ones_multiplier_R_series_qualifier  <- MARSS(Yt, model = model.gen, method = 'kem',
                                         control = list(maxit =  1000, trace = 1, conv.test.slope.tol = 0.05),
                                         inits = coef(marss_fit_QdiagEqual_Rseries))
  
  write_rds(marss_fit_Q_ones_multiplier_R_series_qualifier, path = 'marss_fit_Q_ones_multiplier_R_series_qualifier.rds')
}


```


#### marss_fit_Q_covar_multiplier_R_series_qualifier

```{r marss_fit_Q_covar_multiplier_R_series_qualifier}

# 9

# Standardize the diagonal of gam18_rating for estimation
# gam18_diag_std <- diag(gam18_rating$Vp)/mean(diag(gam18_rating$Vp))

if( file.exists( 'marss_fit_Q_covar_multiplier_R_series_qualifier.rds' )){
  marss_fit_Q_covar_multiplier_R_series_qualifier <- read_rds( 'marss_fit_Q_covar_multiplier_R_series_qualifier.rds' )
} else {
  # State variance matrix is based on the gam18_measPair$Vp
  Q <- matrix(list('0.1166336*q',0,0,0,0,                                  # 'covar_multiplier'
                   0,'1.3179593*q',0,0,0  ,
                   0,0,'0.6810473*q',0,0,
                   0,0,0,'1.0737010*q',0,
                   0,0,0,0,'1.8106588*q'), 5, 5)
  
  # Initialize measurement error vector with large variance for all days
  R  <- array(0, dim = c(1, 1, nrow(df_stage_flow_allDays)))               # 'series_qualifier'
  # Measurement variance specification
  #   The standard error of flow as a percentage that varies with perceived accuracy of the measured flow
  R[1,1, ndx_ian_Yt] <- (df_stage_flow_allDays$se_pct[ndx_ian_Yt]/100)^2
  
  # Specify model
  model.gen=list(Z=Zt,A=A,R=R,U=U,Q=Q,x0=X0,V0=V0,tinitx=0)
  
  # Refit model: trace = 1 provides parameter etimates with each iteration with the corresponding log-likelihood
  marss_fit_Q_covar_multiplier_R_series_qualifier  <- MARSS(Yt, model = model.gen, control = list(maxit =  3000, trace = 1, 
                                                            conv.test.slope.tol = 0.05), method = 'kem',
                      inits = coef(marss_fit_Q_covar_multiplier_R_series_qualifier))
  
  write_rds(marss_fit_Q_covar_multiplier_R_series_qualifier, path = 'marss_fit_Q_covar_multiplier_R_series_qualifier.rds')
}

```



### State-Space Modeling Results


```{r state_space_compare}

# MARSS results for process variance Q 'diagonal and equal' for alternative specifications for the measurement variance R
#   Q = 'diagonal and unequal'
#   R is 'diagonal and equal'
marss_fit_Q_ones_multiplier_R_one_multiplier   <- read_rds('marss_fit_Q_ones_multiplier_R_one_multiplier.rds' )

#   Q = 'diagonal and unequal'
#   R = 'diagonal and equal'
marss_fit_Q_unequal_scalars_R_one_multiplier <- read_rds('marss_fit_Q_unequal_scalars_R_one_multiplier.rds')

#   R is based on the mean perceived measurement accuracy based on the qualifier: 
#     Standard error was interpreted as equivalent to  SE of flows in natural log units.
marss_fit_Q_ones_multiplier_R_mean_qualifier <- read_rds('marss_fit_Q_ones_multiplier_R_mean_qualifier.rds')

 
Q <- 'unconstrained'                                                    # 'ones_multiplier'
# R is specified by qualifiers of discrete flow measurement              
# R  <- matrix(mean( (df_stage_flow_allDays$se_pct[ndx_ian_Yt]/100)^2),1,1)    # 'mean_qualifier'
marss_fit_Q_unconstrained_R_mean_qualifier <- read_rds('marss_fit_Q_unconstrained_R_mean_qualifier.rds')

#State variance matrix is based on the gam18_measPair$Vp
# Q <- matrix(list('0.1166336*q',0,0,0,0,
# 
# R is estimated
# R  <- 'diagonal and equal'
marss_fit_Q_covar_multiplier_R_one_multiplier <- read_rds('marss_fit_Q_covar_multiplier_R_one_multiplier.rds')


#   R is based on the mean perceived measurement accuracy based on the qualifier: 
#     Standard error was interpreted as equivalent to  SE of flows in natural log units.
#   Q is based on the standardized diagonal of gam18_rating$Vp
marss_fit_Q_covar_multiplier_R_mean_qualifier <- read_rds( 'marss_fit_Q_covar_multiplier_R_mean_qualifier.rds' )
 
#   R is based on the perceived measurement accuracy based on the qualifier: 
#     Standard error was interpreted as equivalent to  SE of flows in natural log units.
#     Excellant measurements were assigned a standard error of 0.02
#     Good measurements were assigned a standard error of 0.05,
#     Fair measurements were assigned a standard error of 0.08,
#     Poor measurements were assigned a standard error of 0.15.
marss_fit_Q_ones_multiplier_R_series_qualifier     <- read_rds('marss_fit_Q_ones_multiplier_R_series_qualifier.rds')


#   Q is diagonal and unequal variances based on gam18_rating, with equal multiplier
#   R is based on the perceived measurement quality, in percent
marss_fit_Q_covar_multiplier_R_series_qualifier     <- read_rds('marss_fit_Q_covar_multiplier_R_series_qualifier.rds')


# Set up data.frame to compare models,
marss_compare <- data.frame( Model = seq(1,8),
  Q.Process     = c('ones_multiplier', 'unequal_scalars', 'ones_multiplier',  'covar_multiplier', 'covar_multiplier',
                    'symmetrical', 'ones_multiplier', 'covar_multiplier' ),
                             R.Measurement = c('estimated scalar', 'estimated scalar', 'specified mean', 'estimated scalar', 
                                                'specified mean', 'specified mean', rep('specified series', 2)),
  'Method'         = c( 'BFGS', 'BFGS', 'BFGS', 'BFGS', 'BFGS', 'BFGS', 'kem', 'kem' ),
  'Q_Process'      =
    format( c(marss_fit_Q_ones_multiplier_R_one_multiplier$par$Q, 
              mean(marss_fit_Q_unequal_scalars_R_one_multiplier$par$Q),
              marss_fit_Q_ones_multiplier_R_mean_qualifier$par$Q, 
              marss_fit_Q_covar_multiplier_R_one_multiplier$par$Q,
              marss_fit_Q_covar_multiplier_R_mean_qualifier$par$Q,
              mean(marss_fit_Q_unconstrained_R_mean_qualifier$par$Q),
              marss_fit_Q_ones_multiplier_R_series_qualifier$par$Q,
              marss_fit_Q_covar_multiplier_R_series_qualifier$par$Q), digits = 4), 
  'R_Measurement'  = c(format(marss_fit_Q_ones_multiplier_R_one_multiplier$par$R, digits = 4),
                       format(marss_fit_Q_unequal_scalars_R_one_multiplier$par$R, digits = 4),
                       format(marss_fit_Q_ones_multiplier_R_mean_qualifier$model$fixed$R, digits = 4),
                       format(marss_fit_Q_covar_multiplier_R_one_multiplier$par$R, digits = 4),
                       format(marss_fit_Q_covar_multiplier_R_mean_qualifier$model$fixed$R, digits = 4),
                       format(marss_fit_Q_unconstrained_R_mean_qualifier$model$fixed$R, digits = 4),
                       rep('R[qualifier(t)]',2)),
  'Likelihood'     = format(c(marss_fit_Q_ones_multiplier_R_one_multiplier$logLik, 
                              marss_fit_Q_unequal_scalars_R_one_multiplier$logLik,
                              marss_fit_Q_ones_multiplier_R_mean_qualifier$logLik,
                              marss_fit_Q_covar_multiplier_R_one_multiplier$logLik,
                              marss_fit_Q_covar_multiplier_R_mean_qualifier$logLik,
                              marss_fit_Q_unconstrained_R_mean_qualifier$logLik,
                              marss_fit_Q_ones_multiplier_R_series_qualifier$logLik,
                              marss_fit_Q_covar_multiplier_R_series_qualifier$logLik), digits = 4),
  'AIC'            = format(c(marss_fit_Q_ones_multiplier_R_one_multiplier$AIC,
                              marss_fit_Q_unequal_scalars_R_one_multiplier$AIC,
                              marss_fit_Q_ones_multiplier_R_mean_qualifier$AIC,
                              marss_fit_Q_covar_multiplier_R_one_multiplier$AIC,
                              marss_fit_Q_covar_multiplier_R_mean_qualifier$AIC,
                              marss_fit_Q_unconstrained_R_mean_qualifier$AIC,
                              marss_fit_Q_ones_multiplier_R_series_qualifier$AIC,
                              marss_fit_Q_covar_multiplier_R_series_qualifier$AIC), digits = 4),
  no_parameters = c(2, 6, 1, 2, 1, 15, 1, 1))
                                               
                   
marss_compare %>% 
  kable( caption = 'Table 5. Summary statistics from Multivariate Autoregressive State-Space estimation.',
         col.names = c('Model', 'Process Variance, Q', 'Measurement Variance, R', 'Parameter Estimation Method', 
                       'Process Variance, Q', 'Measurement Variance, R','Log-likelihood', 'Akaike Information Criteria',
                       'Number of param-eters')) %>% 
  kable_styling( position = 'center', latex_options = 'scale down') %>% 
  column_spec(1, width = '8em' ) %>% 
  column_spec(3, width = '8em' ) %>%
  column_spec(6, width = '8em' ) %>% 
  column_spec(9, width = '8em' )%>% 
  add_header_above( c(' ' = 1, 'Structure' = 2, ' ' ,'Characterization' = 2, 
                      ' ' = 1, ' ' = 1, ' ' = 1), align = 'center') %>% 
  footnote(general = c('log-likelihood, a goodness of fit statistic that is maximized in model selection,',
           'AICc: Akaike Information Criteria is a measure of out-of-sample prediction error that is minimized in model selection.',
           'SE Qualifier: Standard Error of Measurement Qualifier, in percent. Excellent = 2%, Good = 5%, Fair = 8%, and Poor = 12%',
           'Process Variance Q estimated as 5 x 5 diagonal matrix with equal variances.',
           'BFGS: BroydenFletcherGoldfarbShanno quasi Newton parameter estimation algorithm.',
           'kem:  kernelized expectation-maximization algorithm for parameter estimation.'))
             
  




```

## Results of State Space Modeling

Results of two model parameterizations are compared.  The first model estimated the process variance matrix Q as a time invariant 5 x 5 diagonal matrix with equal diagonal components and a time-varying scalar measurement variance R. The elements of R were based on the subject quality of the measurement as generally 'Good' with less then 5 percent error, and 'Fair' with an error of 


```{r eval_state_space}


# Select model 3 for analysis
# marss_fit   <- marss_fit_QdiagUnequal_Rseries
# marss_fit <- marss_fit_QdiagEqual_RdiagEqual
marss_fit   <- marss_fit_Q_ones_multiplier_R_series_qualifier
marss_kf    <- MARSSkf( marss_fit )
# kfss function provids Innov, J, J0, Kt, Sigma, Vtt
# marss_kfss  <- MARSSkfss( marss_fit )
  


df_stage_flow_allDays$ytT    <- marss_fit$ytT[1, ]

df_stage_flow_allDays$ytT.se <- marss_fit$ytT.se[, ]

# df_stage_flow_allDays$Kt     <- marss_kfss$Kt

df_stage_flow_allDays <- addWaterYear(df_stage_flow_allDays)

df_stage_flow_allDays <- df_stage_flow_allDays %>% 
  mutate( cl025 = ytT + qnorm( 0.025 ) * ytT.se ,
          cl975 = ytT + qnorm( 0.975 ) * ytT.se ) 

```


## Plot Changes in the state parameters over time


```{r, state_parameter_matrix, fig.height = 9, fig.width = 9 }

# marss_state_tT is the estimate of the state at time t given all time T
marss_state_tT           <- data.frame( t(as.matrix( marss_kf$xtT, ncol = 5, nrow = 10966)))
marss_state_tT$Date      <- flow_daily_pub$Date
colnames(marss_state_tT) <- c('x1 (b0)', 'x2 (b1)', 'x3 (b2)', 'x4 (b3)', 'x5 (b4)', 'Date')

# Identify days of Discrete Flow Measurement (# # # DFM # # #)
ndx_DFM <- which( !is.na( df_stage_flow_allDays$flow_discrete_mea ) )


marss_state_tT %>% 
  gather( key = 'Parameter', value = 'Estimate', -Date) %>% 
  ggplot( aes( x = Date, y = Estimate, group = Parameter )) +
  geom_line() +
  geom_vline( xintercept = df_stage_flow_allDays$Date[ ndx_DFM ], color = 'lightblue' ) +
  theme_bw() +
  xlab( 'Date') + 
  ylab( 'State Estimates') +
  theme( legend.position = 'bottom') +
  facet_wrap( ncol = 1, Parameter ~ ., scales = 'free' ) +
  ggtitle( 'Figure xx. Daily estimates of the state vector as parameters of the corresponding basis functions.',
           subtitle = 'Vertical lines indicate dates of discrete flow measurements.')



```



```{r adaptive_rating, fig.height = 16, fig.width = 9}
# Design matrix for allDays
# design_matrix_allDays <- predict( gam18_rating, newdata = 
#                                   data.frame( stage =  df_stage_flow_allDays$stage_daily_mean), 
#                                   type = 'lpmatrix')
# 
# # Allocate vector for marss flows for all days
# marss_ytT_allDays <- rep(NA, length( df_stage_flow_allDays ))
# 
# # Compute marss flow for all days
# for (i in 1:nrow( df_stage_flow_allDays )){
#   marss_ytT_allDays[i] <- exp( design_matrix_allDays[ i, ] %*% matrix(marss_kf$xtT[ , i], 5, 1) )
# }
# 
sel_yr <- seq(from = 1990, to = 2019, by = 3)

# Design matrix formed with from Rating 18 basis functions using the stage range in Rating 21
#   The range in stage is the widest range of stage (0.53 - 8.10 ft)
design_matrix_rating <- predict( gam18_rating,  newdata = data.frame( stage = rating21$stage), 
                                 type = 'lpmatrix')



# Design matrix for rating augmented with corresponding stage
stage_design_matrix_rating <- cbind(rating21$stage, design_matrix_rating) 
colnames(stage_design_matrix_rating)[1] <- 'stage'




# Preallocate matrix to contain time-varying flows at times of dDscrete Flow Measurement (DFM)
flow_rating_DFM <- matrix(NA, nrow = nrow(rating21), ncol = length( ndx_DFM ) )

# design_matrix <- as.matrix(design_matrix_rating[,1:5])

for (i in 1:length(ndx_DFM )){
  flow_rating_DFM[ , i ] <- exp( design_matrix_rating %*% matrix( marss_kf$xtT[ , ndx_DFM[ i ]], 5, 1 ) )
}


flow_rating_DFM_vec <- matrix( flow_rating_DFM, ncol = 1, byrow = FALSE)

adaptive_rating_measDays <- data.frame( Date       = rep(df_stage_flow_allDays$Date[ ndx_DFM ], each  = nrow(   rating21)),
                                        stage      = rep(rating21$stage,                        times = length( ndx_DFM )),
                      marss_flow = flow_rating_DFM_vec )
adaptive_rating_measDays <- adaptive_rating_measDays %>% addWaterYear()


discrete_meas_allDays <- discrete_meas %>% 
  left_join( df_stage_flow_allDays, by = 'Date' )

# Find indices of allDays in which discrete flow measurements occurred
ndx_allDays_measDays <- which( df_stage_flow_allDays$Date %in% discrete_meas$Date )


tmp <- df_discrete_meas %>%
  # mutate( Date = measurement_dt ) %>% 
  addWaterYear() %>% 
  filter( waterYear %in% sel_yr)


adaptive_rating_measDays %>% 
  dplyr::filter( waterYear %in% sel_yr) %>%
  arrange( marss_flow ) %>% 
  ggplot( aes(x = marss_flow, y = stage, group = waterYear ), alpha = 0.25) +
  geom_line(  aes( x = marss_flow, y = stage), size = 0.25, color = 'blue', alpha = 0.5) +
  # geom_point( aes( x = discrete_meas$Date, y = discrete_meas$flow_discrete_mea, 
  #        ), color = 'blue') +
  labs(title = 'Date of Measurement:' ) + 
  theme_bw() +
  facet_wrap( ~ waterYear, ncol = 2) +
  theme( legend.position = 'right') +
  xlab( 'Streamflow, in cubic feet per second' ) +
  ylab( 'Stage, in feet above gage datum' ) + 
  labs(title = 'Stage-Flow Ratings at 04122500 Pere Marquette River at Scottville, MI') +
  geom_line( data = rating18, aes( x = rated18_flow, y = stage, group = NULL), color = 'red', size = 0.5 ) +
  geom_line( data = rating21, aes( x = rated21_flow, y = stage, group = NULL), color = 'forestgreen', size = 0.5 ) +
  geom_point( data = tmp, aes( x = flow_discrete_mea, y = stage_discrete_mea, color = factor(waterYear))) +
  theme( legend.position = 'bottom')

```


```{r plot_daily_flow_se, fig.width = 12, fig.height = 20 }
sel_yr <- seq(from = 1990, to = 2019, by = 3)

df_stage_flow_allDays %>% 
  dplyr::filter( waterYear %in% sel_yr ) %>% 
  mutate( y_ice = pmin(flow_daily_pub, exp(ytT)) ) %>%  
  ggplot( aes( x = Date  )) +
  geom_line( aes( y = exp(ytT) ), color = 'red' ) +
  geom_point(aes( y = flow_discrete_mea,   color = qualifier), size = 1.5 ) +
  geom_point(aes( y = flow_discrete_mea ), size = 5, shape = 1, color = 'black'  ) +
  geom_line( aes( y = exp(cl025) ), color = 'blue', size = 0.5) +
  geom_line( aes( y = exp(cl975) ), color = 'blue', size = 0.5) +
  geom_line(  aes( y = flow_daily_pub ), color = 'darkgreen', size = 0.7) +
  geom_ribbon( aes(ymin = y_ice, ymax = exp(ytT) ), fill = 'grey', alpha = 0.5  ) +
  theme_bw() +
  scale_y_log10() +
  xlab( 'Date') + 
  ylab( 'Daily mean streamflow, in cubic feet per second') +
  theme( legend.position = 'bottom') +
  facet_wrap( ncol = 2, waterYear ~ ., scales = 'free' ) +
  ggtitle( 'Figure x. Daily mean expected streamflow (red) with 95-percent confidence intervals (blue) based on adaptive spline \nrating model with historically published streamflow (green).',
           subtitle = 'Grey filled areas between published and expected daily flow primarily show ice affects.')


```


```{r rating_updates, eval = FALSE, fig.height = 18, eval = FALSE, fig.width = 10}

Z <- predict( gam18_rating, newdata = data.frame( stage = rating18$stage ),
              type = 'lpmatrix')

dim(Z)

xtT_DFM <- marss_kf$xtT[, ndx_DFM ]

dim( xtT_DFM )


ytT_DFM <- Z %*% xtT_DFM

ytT_DFM_vec <- matrix(exp(ytT_DFM), ncol = 1, byrow = TRUE)

print(ytT_DFM[1:15, 1])

print(ytT_DFM_vec[1:15])

ytT_DFM_df <- data.frame( Date    = rep( df_stage_flow_allDays$Date[ ndx_DFM ], each = nrow(rating21)),
                          stage   = rep( rating21$stage, length(ndx_DFM) ), 
                          ytT     = ytT_DFM_vec)

ytT_DFM_df <- addWaterYear( ytT_DFM_df )

colnames(discrete_meas)[2] <- 'Date'


tmp <- tmp_df %>% 
  dplyr::filter( waterYear %in% sel_yr)


ytT_DFM_df %>% 
  dplyr::filter( waterYear %in% sel_yr ) %>%
  ggplot( aes(x = ytT, y = stage, group = Date ), alpha = 0.25) +
  geom_line(  aes( x = ytT, y = stage), size = 0.5, color = 'blue', alpha = 0.5) +
  # geom_point( aes( x = discrete_meas$Date, y = discrete_meas$flow_discrete_mea, 
  #        ), color = 'blue') +
  labs(title = 'Date of Measurement:' ) + 
  scale_x_log10() +
  geom_point(data   = df_discrete_meas, 
             aes( x = flow_discrete_mea, y = stage_discrete_mea, group = Date), 
             color  = 'orange3', size = 3) +
  theme_bw() +
  facet_wrap( ~ waterYear, ncol = 2) +
  theme( legend.position = 'right') +
  xlab( 'Streamflow, in cubic feet per second' ) +
  ylab( 'Stage, in feet above gage datum' ) + 
  labs(title = 'Stage-Flow Ratings at 04122500 Pere Marquette River at Scottville, MI') +
  geom_line( data = rating18, aes( x = rated_flow, y = stage, group = NULL), color = 'red', size = 1 ) +
  geom_line( data = rating21, aes( x = rated_flow, y = stage, group = NULL), color = 'forestgreen', size = 1 )


```
