---
title: "Bayesian Rating Curve"
author: "Dave Holtschlag"
date: "`r format(Sys.Date(), '%A %b %d, %Y') `"
output: html_document
---

```{r setup, include=FALSE}
library( "rethinking" )
options(mc.cores = parallel::detectCores())
library(dataRetrieval)
library(tidyverse)
library(ggthemes)
library(splines)
library(mgcv)
# Package expss contains the function vlookup
library(expss)
knitr::opts_chunk$set(echo = TRUE)
```

## Specify Streamgage for analysis

```{r specify_Gage}
gageInfo  <- readNWISsite('04112500')

# print(t(gageInfo))

siteNo    <- gageInfo$site_no

```

## Read in rating and flow measurements

```{r read_rating_data}
# Rating curve at the selected site.  The 'exsa' parameter is an extended table
#   provide detailed info on stage discharge
ratingData <- readNWISrating(siteNo, 'exsa') %>% 
  rename('flow'         = 'DEP',
         'stage'        = 'INDEP') 

# Note: the 'base' rating information provides initial estimates of knot placement
# ratingKnots <- readNWISrating(siteNo, 'base')
# colnames(ratingKnots) <- c('stage', 'flow', 'stor') 

# Rename fields to convention reference
# colnames(ratingData) <- c('stage','shift','flow')

measData   <- readNWISmeas(siteNo, expanded = TRUE, convertType = TRUE)

measData <- measData %>% 
  rename('stage'        = 'gage_height_va',
         'flow'         = 'discharge_va',
         'quality'      = 'measured_rating_diff') %>% 
  filter(quality       != 'Unspecified',
         !is.na(quality)) %>% 
  mutate(quality        = ordered(quality, levels = c('Excellent', 'Good', 'Fair', 'Poor'))) %>% 
  arrange( stage )
  
  
# Only cases where flow and stage are measured
measData <- measData[complete.cases(measData$stage, measData$flow),]

# Add small uniform random number to stage for uniqueness
measData <- measData %>% 
  mutate(log_stage_ = log_stage + runif(nrow(measData), -0.000049, 0.000049)) %>% 
  arrange(log_stage_)

measSumm <- measData %>% 
  group_by( log_stage ) %>% 
  dplyr::summarize( mean_log_flow = mean(log_flow),
                    sd_log_flow   = sd(log_flow),
                    cnt_log_flow  = n() )

  
# Generate alternative knots based on measurement data and equally spaced
# measurements in log10 space.

# ratingKnots <- data.frame(
#   stage_knots <- seq(from       = sqrt(min(measData$stage)),
#                      to         = sqrt(max(measData$stage)),
#                      length.out = 7)^2)
# 
# 
# 
# ratingKnots$flow_alt  <- approx(measData$stage, measData$flow,
#                                 ratingKnots$stage_alt, ties = mean, 
#                                 rule = 2)$y

num_knots   <- 11

# Did not like quantile estimator of knots because curvature was disregarded
# knot_list   <- quantile( measData$stage, probs = (seq(0, 1, length.out = num_knots)^0.5) )
v <- 1.5
knot_list   <- 10^(10^(seq(log10(log10(min(measData$stage)-v)),
                           log10(log10(max(measData$stage)-v)), length.out = num_knots))) + v
ratingKnots <- data.frame( stage_knots = knot_list )


```

## Plot empirical distributions of measured flow and stage


```{r dist_flow, fig.cap = 'Figure 1. Distributin of Measured Flows'}

measData %>% 
  ggplot( aes(log10(flow)) ) +
  geom_density(fill = 'blue', color = 'blue', alpha = 0.2, show.legend = TRUE) +
  stat_function(fun  = dnorm, 
                args = list( mean = mean(log10(measData$flow)),
                             sd   =   sd(log10(measData$flow))),
                  lwd = 1, color = 'purple', show.legend = TRUE) +
  theme_bw() + 
  xlab('Log10 of Measured Flow, in cubic feet per second') +
  ylab('Probability Density')

```

## Distribution of Stage


```{r dist_flow, fig.cap = 'Figure 2. Distributin of Measured Stages'}

# Gamma distribution is questionable applicable because of arbitrary offset
# Fit gamma distribution
fit.gamma.logstage <- fitdist(log10(measData$stage)-min(log10(measData$stage))+.0001,
                              distr = 'gamma', method = 'mle')

summary(fit.gamma.logstage)

measData %>% 
  ggplot( aes(log10(stage)-min(log10(stage))),  color = 'blue' ) +
  geom_density(fill = 'blue', alpha = 0.2, show.legend = TRUE) +
  stat_function(fun   = dgamma, 
                args  = list(shape = fit.gamma.logstage$estimate[[1]],
                             rate  = fit.gamma.logstage$estimate[[2]])) +
                #              sd   =   sd(log10(measData$stage))),
                #  lwd = 1, color = 'purple', show.legend = TRUE) +
  theme_bw() + 
  xlab('Log10 of Measured Stage minus point of zero flow offset, in feet') +
  ylab('Probability Density')

```



## Plot Rating and Measurements

```{r plot_rating, fig.height = 7}

ratingData %>% 
  ggplot( aes(y = flow, x = stage)) +
  geom_line( color = 'blue') + 
  geom_point(data = measData, aes(x = stage, y = flow, 
                                  color = quality)) +
  # geom_point(data = ratingKnots, aes(x = stage, y = flow),
  #             olor = 'black', shape = 4, size = 5) +
  
  geom_vline(data = ratingKnots, aes(xintercept = stage_knots), 
             color = 'salmon', linetype = 'dashed' ) +

  # geom_point(data = ratingKnots, aes(x = stage_alt, y = flow_alt),
  #            color = 'red',   shape = 3, size = 5 ) +
  
  theme_bw() +
  theme(legend.position = "bottom") +
 
  scale_y_continuous(name   = 'Flow, in cubic feet per second',
                     trans  = "log10",
                     breaks = c(1, 10, 50, 100, 200, 500, 1000, 2000, 4000, 6000)) +
  scale_x_continuous(name   = "Stage, in feet",
                     trans  = "log10",
                     breaks = c(seq(3, 12, by = 1))) +
  ggtitle(label = paste('Figure 1. Stage-flow relation at streamgage',gageInfo$site_no, gageInfo$station_nm))
```

## Frequency of measurement qualities

The quality of measurements may need to be recomputed based on the rating in place at the time of measurement.

```{r table_quality}

measData$flow_exp <- approx( ratingData$stage, ratingData$flow, xout = measData$stage, rule = 2 )$y

measData$flow_err <- (measData$flow - measData$flow_exp) / (measData$flow_exp) * 100.

# Frequency of measurement qualities
print(table(measData$quality))


measData %>% 
  ggplot( aes( quality, flow_err) ) +
  geom_boxplot() +
  xlab('Rated Quality of Measurement') +
  ylab('Percent error in flow measurement') +
  geom_hline( aes(yintercept = 0) , color = 'red', linetype = 'dashed' )


```

## Distribution of Square root of measured flow

```{r flow_dist}

dens( log10(measData$flow), norm.comp = TRUE )


```

## Bayesian spline regression model with uncertainty in stage


```{r bayes_spline}

measData <- measData %>% 
  mutate( log_flow = log10(flow), log_stage = log10(stage))

log_stage <- measData$log_stage_
log_flow  <- measData$log_flow


Gspline <- s(log_stage, bs = 'cr')

# gam(eval(parse(text='log_flow ~ s(log_stage)')))

Gmodel  <- gam(log_flow ~ eval(Gspline$label) )

tmp2    <- gam(eval(parse(text='log_flow ~ s(log_stage)')))

Gmodel  <- gam(log_flow ~ s(log_stage))

# Gmodel  <- gam(log_flow ~ s(log_stage, bs = 'cr'))

Bspline <- bs(log_stage , 
              # knots = knot_list,
              # knots = knot_list[-c(1, num_knots)],
              df     = 9,
              degree = 3, intercept = TRUE)


m_bs_reg_02 <- quap(
  alist(
    log_flow ~  dnorm( mu, sigma ),
    mu       <- a + G %*% w,
    a        ~  dnorm( 2, 1.5 ),
    w        ~  dnorm( 0, 1   ),
    sigma    ~  dexp(1)
  ),
  data = list( log_flow = log_flow, log_stage = log_stage, G = Gmodel$smooth[[1]]$UZ),
  start = list(w = rep( 0, Gmodel$rank))
)

precis( m_bs_reg_02, depth = 2, prob = 0.95, digits = 5)

```

## Evaluate spline model

```{r eval_bs}

stage.seq <- seq(from = min(measData$log_stage), to = max(measData$log_stage), length = 200)

stage.mu  <- link( m_bs_reg_02)

# Find minimum difference between entry in 
meas_data_ndx <- rep(NA, length(stage.seq))
for (i in 1:length(stage.seq)){
  adiff <- abs(stage.seq[i] - log_stage)
  meas_data_ndx[i] <- which.min(adiff)
  print( paste(i, ndx, adiff[ndx]))
}

# summarize the distribution of stage.mu
mu.mean <- apply( stage.mu[,meas_data_ndx] , 2 , mean )
mu.PI   <- apply( stage.mu[,meas_data_ndx] , 2 , PI , prob=0.95 )

smooth.spline(x = log_stage, y = log_flow)
# , w = NULL, df = measData, lambda = NULL, cv = FALSE,
#               all.knots = FALSE, nknots = .nknots.smspl,
#               keep.data = TRUE, df.offset = 0, penalty = 1,
#               control.spar = list(), tol = 1e-6 * IQR(log_stage), keep.stuff = FALSE)

require(graphics)
plot(dist ~ speed, data = cars, main = "data(cars)  &  smoothing splines")
cars.spl <- with(cars, smooth.spline(speed, dist))
cars.spl
## This example has duplicate points, so avoid cv = TRUE

lines(cars.spl, col = "blue", lwd = 2)
ss10 <- smooth.spline(cars[,"speed"], cars[,"dist"], df = 10)
lines(ss10, lty = 2, col = "red")
legend(5,120,c(paste("default [C.V.] => df =",round(cars.spl$df,1)),
               "s( * , df = 10)"), col = c("blue","red"), lty = 1:2,
       bg = 'bisque')



plot(log_flow ~ log_stage, data = measData, main = "data(cars)  &  smoothing splines",
     col = 'grey')
rate1.spl <- with(measData, smooth.spline(log_stage, log_flow))
rate1.spl
## This example has duplicate points, so avoid cv = TRUE

lines(rate1.spl, col = "blue", lwd = 2)
rate2.spl <- smooth.spline(measData[,"log_stage"], measData[,"log_flow"], df = 8)
lines(rate2.spl, lty = 2, col = "red")

legend(5,120,c(paste("default [C.V.] => df =",round(cars.spl$df,1)),
               "s( * , df = 10)"), col = c("blue","red"), lty = 1:2,
       bg = 'bisque')


# plot raw data
# fading out points to make line and interval more visible
plot( log_flow ~ log_stage , data=measData , col=col.alpha(rangi2,0.5) )
# plot the MAP line, aka the mean mu for each weight
lines( stage.seq , mu.mean )
# plot a shaded region for 89% PI
shade( mu.PI , stage.seq )


# pred      <- link( m_bs_reg_01, data = measData  )
# 
# 
# post <- extract.samples(m_bs_reg_01)
# w <- apply( post$w , 2 , mean )
# plot( NULL , xlim=range(tmp$log_stage) , ylim=c(-5,5) ,
# xlab="year" , ylab="basis * weight" )
# for ( i in 1:ncol(Bspline) ) lines( tmp$log_stage , w[i]*Bspline[,i] )
# 
# mu <- link( m_bs_reg_01 )
# mu_PI <- apply(mu,2,PI,0.95)
# plot( tmp$log_stage , tmp$log_flow , col=col.alpha(rangi2,0.3) , pch=16 )
# shade( mu_PI , tmp$log_stage , col=col.alpha("black",0.5) )


```

## GAM models using mgcv

```{r simple_gam}

ct1 <- gam(Volume ~ s(Height), family=Gamma(link=log), data = trees)
print(ct1)

 pd <- data.frame(Height=c(75,80),Girth=c(12,13))
 
 predict(ct1,newdata=pd,se=TRUE)
 
 Xp <- predict(ct1,newdata=pd,type="lpmatrix")
 
 Xp %*% coef(ct1)
 
```


```{r gam_rate}

measData <- measData %>% 
  mutate( log_flow = log10(flow), log_stage = log10(stage))

log_stage <- measData$log_stage_
log_flow  <- measData$log_flow

gam_spline <- gam(log_flow ~ s(log_stage), data = measData, method = "ML")
measData$gam_hat <- gam_spline$fitted.values


measData %>% 
  ggplot( aes( x = log_stage, y = log_flow)) +
  geom_point() +
  geom_line( aes( x = log_stage, y = gam_hat), color = 'red')


plot(gam_spline)

m_bs_reg_01 <- quap(
  alist(
    log_flow ~  dnorm( mu, sigma ),
    mu       <- a + Bspline %*% w,
    a        ~  dnorm( 2, 1.5 ),
    w        ~  dnorm( 0, 1   ),
    sigma    ~  dexp(1)
  ),
  data = list( log_flow = log_flow, log_stage = log_stage, B = Bspline),
  start = list(w = rep( 0, ncol(Bspline)))
)

precis( m_bs_reg_01, depth = 2, prob = 0.95, digits = 5)




```


}